### 2012-02-18 Paper.tex
First tex version of the paper

\documentclass[bachelor,english]{hgbthesis}

\begin{document}

\title{Hybrid Threads for the Parrot Virtual Machine}

\author{Stefan Seifert}
\studiengang{Software Engineering}
\studienort{Hagenberg}
\abgabemonat{Februar}
\abgabejahr{2012}

\nummer{1010307037-A}
\gegenstand{SWE}
\semester{Wintersemester 2012}
\betreuer{Heinz Dobler}

\frontmatter
\maketitle
\tableofcontents

\chapter{Thanks (1)}

Andrew "whiteknight" Whitworth for laying the foundations for my work, being of tremendous help at all stages of the project and for fixing some of my bugs.

Nat "Chandon" Tuck for implementing green thread basics on which my work is based.

Christoph "cotto" Otto for bringing me into the Parrot project.

Brian "benabik" Gernhardt for lurking in the IRC channel and at least trying to help with my many questions.

\chapter{Abstract (1)}

\begin{german}
\chapter{Kurzfassung (1)}
\end{german}

\chapter{Table of Contents (1)}

	Perl 6 Concurrency
	About Parrot
	Parrot threads
	Nat "Chandon" Tuck green\_threads

\mainmatter

\chapter{Introduction and Motivation (3)}

On July 19th 2000, the Perl 6 design process was announced. Perl 5 had been a very flexible and widely used programming language but had started to show it's age and suffered from early design decisions. The Perl interpreter is written in C and has accumulated a lot of cruft over the years. The general consensus among the core developers was that the code had reached a state where maintenance was approaching impossibility. [http://use.perl.org/~masak/journal/40451]. An attempt to reimplement these internals had failed but led to the decision that the interpreter for a Perl 6 language should be developed independently of the needs of Perl 5. Since the Perl 6 syntax was very much in flux (and parts of it still are) the designers of these new internals tried to work very independently of any syntax related questions. [http://www.developer.com/lang/perl/article.php/10940\_3076571\_3/Perl-6-and-the-Parrot-Project.htm]. Taking up the name born in an April Fool's Day joke announcing the mergin of the Perl and Python programming languages, these new internals were called Parrot.

Parrot evolved from being just the interpreter for the new version of Perl to being a language independent virtual machine providing features like garbage collection, exception handling and dynamic typing. At the time when the Parrot project started, the Java and .NET virtual machines were widely used, but both targeted statically typed languages. Parrot thus filled a quickly growing niche.

The Perl 6 design process began with asking the Perl users for what they were expecting from the new version of the language. The very first feature that got asked for was well integrated multi-threading support [http://dev.perl.org/perl6/rfc/1.html]. Perl 5 had two different implementations of thread support. In the first model, called 5005threads, data was shared by default and shared access to data had to be explicitely synchronized. This was similiar to the models used by languages such as C or Java. The implementation however suffered from data corruption and crashes and thus was not recommended for production use [http://search.cpan.org/~nwclark/perl-5.8.8/lib/Thread.pm]. Perl 5.6 introduced the newer model called ithreads, mostly as a way to emulate fork() on Win32 platforms. Perl 5.8 exposed this API to the user of the programming language. In this new model, all data would is copied to each thread and afterwards thread local. Data must be explicitely shared between threads. In other words, in Perl threads are not lightweight at all. They have severe impact on memory usage, writes to shared data are expensive and still not all features of the language are usable in threaded programs.

Being born at a time when Perl 6 still looked much more similiar to Perl 5 than it does nowadays, Parrot's threading support initially was very close to Perl's ithreads model. Previous attempts to change this into the more conventional model of data shared by default or implementing new technologies like Software Transactional Memory failed. For example Parrot has never supported running multiple threads and having garbage collection at the same time.

\section{Why is multi-threading support so important?}

In the year 2005 development of faster CPUs shifted from increased speed of a single core to adding more cores. Modern processors contain up to 12 cores with even mobile phones having up to four. To utilize a modern CPU's power, code needs to be run in parallel. In UNIX (and thus Perl) tradition, this is accomplished using multiple processes which indeed is a good solution for many use cases. For many others like Perl 6's auto threading of hyper operators, the cost of process setup and communication would be prohibitively high except for very large data sets.

\section{Why is it so difficult to implement multi-threading support in Perl or Parrot?}

Low level programming languages like C provide only the bare necessities, leaving the responsibility for preventing data corruption entirely to the user. A high level language like Perl 6 on the other hand offers complex and compund data types, handles garbage collection and a very dynamic object system. Even seeminlgy simple things like a method call can become very complex sequences. In a statically typed programming language the definition of a class is immutable. Thus calling a method on an object contains just the steps of determining the object's class, fetching the required method from this class and calling it. Calling the same method again may then even omit the first two steps since their results cannot change.

In a dynamic language, the object may change it's class at runtime. The inheritance hierarchy of the class may be changed by adding or removing parent classes. Methods may be added to or removed from classes (or objects) at runtime and even the way to find a method of a class may change. So a simple method call may result in the following steps:
%
\begin{itemize}
\item determining the class of the object
\item determining the method resolution method of the class
\item finding the actual method to call
\item calling the method
\end{itemize}
%
These steps have to be repeated for every following method call, because their results may change any time. In a threaded environment, a thread running in parallel may change the underlying data and meta data in between those sequences and even between those steps. As a consequence, this meta data has to be protected from corruption introducing the need for many locks in a very performance critical area.

While there are multi threaded garbage collection schemes, their implementation has not yet been attempted in Parrot.

Many interpreters for dynamic languages like Python or Ruby handle this problem by using a global interpreter lock to effectively serialize all operations. This is a proven and reliable way but again leaves much of the hardware's potential unused.

\section{Current status}

During these years of back and forth and failed attempts of adding threading support to Parrot, the Perl 6 specification evolved to a point where the largest parts of the language were covered and its features implemented in the compilers. The lack of concurrency primitives in Parrot prevents however prevents any progress in the area of concurrency support.

Currently Parrot does not have any threading support at all. The previous, defunct implementation has been removed.

This paper suggests a new approach based on a hybrid threading system. So called green threads are used to simplify the implementation of a nearly lock free multi threading implementation. This approach is based on a design by Andrew Whitworth and Nat Tuck.

\chapter{Parrot (4)}

	July 19, 2000, Perl 6 design process announced
	http://www.perl.com/pub/2001/04/01/parrot.htm
	April Fool's Day joke

	Historical development (1)

\chapter{Concurrency in other Programming Platforms (5)}

\section{Python}

\chapter{Green threads (3)}

"Green threads" or "lightweight threads" are threads which are managed by the virtual machine instead of by the operating system. The virtual machine contains a scheduler and support for preempting running tasks. This is very similiar to an operating system running on a system with a single CPU core.

Their advantages are:
%
\begin{itemize}
\item They allow pseudo concurrent processing without endangering the interpreter's internal consistency.
\item They support using a naive single threaded garbage collector.
\item Green threads are very light weight with low memory overhead and close to zero creation time.
\item The interpreter controls the point at which a green thread is preempted.
\end{itemize}

Their disadvantages are:
%
\begin{itemize}
\item They do not allow more than one CPU core to be used for computations
\item Blocking calls like I/O block the whole interpreter including other green threads.
\item The interpreter has to contain logic and timers to control green threads.
\end{itemize}
%
	Chandon's green\_threads branch

\chapter{Design of Hybrid Threads (3)}

\chapter{Implementation of Hybrid Threads (15)}

\section{Runloops (1)}

\section{The Scheduler (4)}

\section{Threads (5)}

\subsection{Creation (1)}

\subsection{Proxies (2)}

\subsection{Writing shared data (2)}

\subsection{Garbage Collection (3)}

\chapter{Tests and Benchmarks (4)}

\chapter{Conclusion, Further Work and Experiences (3)}

\begin{itemize}
\item replace thread\_data->interp\_lock usage by preallocated proxy objects
\item allow threads to create threads
\item allow multi level proxying (don't create a proxy for a proxy, unpack first)
\end{itemize}

\MakeBibliography{References, Literature (1)}

\include{messbox}
\end{document}

### 2012-02-18 Paper.tex
First part of the Parrot chapter

@@
 Taking up the name born in an April Fool's Day joke announcing the mergin of the Perl and Python programming languages, these new internals were called Parrot.
+[http://www.perl.com/pub/2001/04/01/parrot.htm]
 
@@
 
-	July 19, 2000, Perl 6 design process announced
-	http://www.perl.com/pub/2001/04/01/parrot.htm
-	April Fool's Day joke
+Parrot consists of the virtual machine (also called interpreter), and various tools to facilitate the implementation of programming languages on top of the Parrot VM (the Parrto Compiler Toolkit).
+In this paper, we concentrate on the virtual machine itself.
+The interpreter itself is written in C.
+Example code and testcases are written in PIR (Parrot Intermediate Representation) a high level assembly language which abstracts away register allocations and function calling conventions.
 
-	Historical development (1)
+Contrary to other widely used virtual machines like the JVM or the CLR which are stack based, Parrot mirrors contemporary hardware CPUs more closely by being register based.
+The rationale behind giving up the simplicity of a stack based implementation is the hope of simplifying just in time compilation and improved performance of nested function and method calls.
+
+The current design uses four sets of registers with each and unlimited number of usable registers. The four sets are:
+%
+\begin{itemize}
+\item integer
+\item floating point
+\item string
+\item polymorphic container (PMC)
+\end{itemize}
+
+The first three should be self explanatory.
+Strings in Parrot are a low level type with the interpreter handling all memory allocation issues and Unicode encoding.
+String values themselves are immutable.
+
+\section{PMCs}
+
+PMCs are containers for all higher level types such as objects, arrays, hash tables or code.
+Thus they are very similar to Python's PythonObject types.
+PMCs are defined by C structs and are fully garbage collected.
+Their definition looks like:
+\begin{CCode}
+struct PMC {
+    Parrot_UInt    flags;
+    VTABLE         *vtable;             /* Pointer to vtable. */
+    DPOINTER       *data;               /* Pointer to attribute structure. */
+    PMC            *_metadata;          /* Pointer to metadata PMC. */
+};
+\end{CCode}
+In short it contains a pointer to a type specific data structure and some meta data.
+The vtable is where the type's behaviour is defined.
+It contains a long list of function pointers forming Parrot's unified type interface.
+This interface is a union of numeric, string, array, hash and object like behaviours.
+For example the  \textit{get\_integer} function returns an integer value for the data type.
+For a simple \textit{int} it's the contained value.
+For an array it may be the number of elements.
+The \textit{find\_method} function makes user defined methods of objects possible.
+
+In this way, a language implementor can define the basic types of the language and available operations on them.
+For each vtable entry, there is a corresponding op code in Parrot's bytecode.
+Thus an \textit{inc \$P0} instruction is calling the \textit{increment} vtable function of the PMC contained in the \textit{\$P0} register.
+
+\section{Continuation passing}
+\section{Runloops}
+\section{Exception handling}
+\section{Garbage collector}
+
+\section{Historical development (1)}
 
@@
 
+\section{Java}
 \section{Python}
+\section{Erlang}
 
@@
 \end{itemize}
-%
+
+Green threads differ from coroutines in that the interpreter decides when a running task is to be preempted while a coroutine depends on manual yield calls.
+
 Chandon's green\_threads branch

### 2012-02-19 Paper.tex
Parrot and Green threads chapters

@@
 \semester{Wintersemester 2012}
-\betreuer{Heinz Dobler}
+\betreuer{Dr. Heinz Dobler}
 
@@
 
+\chapter{Concurrency in other Programming Platforms (5)}
+
+\section{Java}
+\section{Python}
+\section{Erlang}
+
 \chapter{Parrot (4)}
 
-Parrot consists of the virtual machine (also called interpreter), and various tools to facilitate the implementation of programming languages on top of the Parrot VM (the Parrto Compiler Toolkit).
+Parrot consists of the virtual machine (also called interpreter), and various tools to facilitate the implementation of programming languages on top of the Parrot VM (the Parrot Compiler Toolkit).
 In this paper, we concentrate on the virtual machine itself.
@@
 
+\section{ParrotInterpreter}
+
+The interpreter itself is represented by a C struct called \textit{parrot\_interp\_t}.
+This structure contains pointers to the garbage collector's runtime data, the loaded types, vtables, the runloop, the current continuation and other global data.
+A pointer to this structure is passed to almost every function as the first parameter.
+
 \section{Continuation passing}
+
+Control flow is modeled in Parrot using continuation passing.
+A continuation is a data structure containing the a state of a program at a given point in it's execution.
+It contains all information which is needed to continue a program in a certain state, e.g.
+a call stack, the instruction pointer and contents of local variables.
+``Calling'' a continuation means restoring the state encapsulated in the continuation.
+
+When a function is called, instead of pushing all this information on the stack, the function is given a return continuation as part of it's parameters.
+Returning from a function means calling the return continuation with possible return values stored in the registers where the calling code is expecting them.
+Continuation passing makes things like tail call optimization \footnote{When as the last statement of a function another function is called and it's value returned unchanged, the outer function may just pass on it's return continuation thus saving the need to create a new continuation and one step on returning from the nested function} very simple and as we will see later will be a very important piece in implementing green threads.
+
 \section{Runloops}
+
+A runloop is in principle the most inner loop executing bytecode in Parrot.
+It is also a data structure containing some information needed to support exception handling.
+When some operation in Parrot bytecode is calling a C function and this C function in turn is again executing Parrot bytecode, a nested runloop is started.
+Examples for such situations include calling a library function with a callback as parameter and exceptions thrown inside Parrot's C code which call a previously defined exception handler in user code.
+
 \section{Exception handling}
+
+Before entering a runloop, the \textit{setjmp} C function is used to save the current stack position and register contents to a data structure stored in the runloop meta data.
+This is effectively creating something resembling a continuation at C level.
+When an exception is created within the interpreter, the runloop stack is searched for the runloop containing a suitable exception handler.
+\textit{longjmp} is then used to unwind the call stack up to the point where setjmp was called and the call environment restored.
+While having great similarities with continuations, this mechanism is more limited.
+It only allows to jump back to a point higher in the call stack.
+
 \section{Garbage collector}
+
+Parrot supports different garbage collector implementations which can be selected at interpreter startup. Currently there are four existing implementations of different algorithms:
+\begin{itemize}
+\item Inf: a GC for debugging purposes which never collects any garbage
+\item MS: a basic mark and sweep implementation
+\item MS2: a non-recursive mark and sweep implementation
+\item GMS: a generational, non-compacting, mark and sweep GC
+\end{itemize}
+with GMS being the default.
+
+A mark and sweep garbage collector operates in two phases.
+In the mark phase, starting from a known root set of objects, the GC follows pointers down the object tree, marking each encountered object as alive.
+In the sweep phase, all objects which are not alive are destroyed and the live flags reset.
+
+In a process with many objects, having to traverse the whole tree may take a considerable amount of time.
+To mitigate this, a generational garbage collector extends this algorithm by assuming that the longer an object has lived, the lower the chances are that it will become unused.
+So the objects are partitioned into different generations.
+The youngest generation would always be traversed while the older generations would be handled much less frequently or even never more at all.
 
@@
 
-\chapter{Concurrency in other Programming Platforms (5)}
+\chapter{Green threads (3)}
 
-\section{Java}
-\section{Python}
-\section{Erlang}
+There are three possible ways how two subroutines could be (at least seemingly) run in parallel:
+\begin{itemize}
+\item Coroutines
+\item Green threads 
+\item Operating system threads
+\end{itemize}
 
-\chapter{Green threads (3)}
+\section{Coroutines}
+
+Coroutines are functions which retain their state between calls.
+Instead of returning, they yield control back to the calling function, possibly with an intermediate result returned.
+A very simple example in Python could look like:
+\begin{GenericCode}
+def counter():
+    for i in range(0, 100):
+	yield i
+\end{GenericCode}
+This function would return one integer for every call, counting from 0 to 100.
+For two functions to run in parallel, they have to cooperate.
+One by calling the other repeatedly and the callee by yielding control back to the caller.
+
+\section{Operating system threads}
+
+Operating system threads are like multiple processes running at the same time but share their memory. The operating system is responsible for giving each of these threads CPU time to run.
+
+\section {Green threads}
 
@@
 The virtual machine contains a scheduler and support for preempting running tasks.
-This is very similiar to an operating system running on a system with a single CPU core.
-
-Their advantages are:
+Preemption means to suspend a task regardless of what it's currently doing and probably resuming it later on.
+This is an implementation of the many-to-one threading model very similiar to what an operating system running on a system with a single CPU core does.
+%
+Advantages of green threads are:
 %
@@
 \item They allow pseudo concurrent processing without endangering the interpreter's internal consistency.
-\item They support using a naive single threaded garbage collector.
 \item Green threads are very light weight with low memory overhead and close to zero creation time.
+\item They do not depend on OS threading support
 \item The interpreter controls the point at which a green thread is preempted.
+\item The interpreter controls the exact scheduling policy and may allow the user to influence it or even take over completely.
+\item Garbage collection can be implemented like in a single threaded process.
+\item Critical sections may simply be protected by disabling the scheduler.
 \end{itemize}
-
-Their disadvantages are:
+%
+Disadvantages include:
 %
 \begin{itemize}
-\item They do not allow more than one CPU core to be used for computations
+\item They do not allow more than one CPU core to be used for computations.
 \item Blocking calls like I/O block the whole interpreter including other green threads.
@@
 
-Chandon's green\_threads branch
+Let's look at a simple example to explain why concurrent processing could endanger the interpreter's internal consistency.
+The basic problem is concurrent writes to shared data.
+Assume we have an implementation of an array class consisting of the field holding the data and the number of contained elements in a seperate member variable:
+\begin{CCode}
+pmclass ResizableIntegerArray auto_attrs provides array {
+    ATTR INTVAL   size;      /* number of INTVALs stored in this array */
+    ATTR INTVAL * int_array; /* INTVALs are stored here */
+\end{CCode}
+To append a new value, the array would have to read the current size, write the new value at the position size + 1 and then write the incremented size back into the member variable.
+Now if two threads simultainously try to do this, it may happen that both read the same size, write to the same position (with one overwriting the value of the other) and write back the same incremented size.
+In this case one of the appended values would be lost.
+
+In a more complicated example, the array could have to resize it's data buffer to accomodate the new value.
+It would read the current size, allocate a new buffer, copy the values to the new buffer and then destroying the old one.
+Again if two threads try to do this simultainously, one of them could still be copying data, while the other is already at the point where it destroys the old buffer.
+This would lead to the copying thread accessing freed memory.
+
+\section{Green threads in Parrot}
+
+Parrot's green threads implementation is based on Nat Tuck's green\_threads branch developed during his Google Summer of Code internship.
+
+In Parrot green threads are called \textit{Task}.
+Each task is assigned a fixed amount of execution time.
+On expiry a timer callback sets a flag which is checked for at execution of every branch operation.
+Since the interpreter's state is well defined at this point, it's internal consistency is guaranteed.
+The same goes for the garbage collector.
+Since task preemption is only done while executing user level code, the GC can do it's work undisturbed and without the need for measures like locking.
+Since user level code is allowed to disable the scheduler, it can be guaranteed to run undisturbed through critical sections.
+
+The scheduler is implemented as a PMC (Parrot's polymorphic container type). The plan is to allow the user to subclass this PMC thus allowing very fine grained control over the scheduling policy.
 
 \chapter{Design of Hybrid Threads (3)}
+
+Using green threads we're now solving the following problems which occur when trying to implement threading support:
+%
+\begin{itemize}
+\item How to ensure internal interpreter consistency when doing writes to shared data.
+\item How to handle garbage collection.
+\end{itemize}
+
+\section{Shared data}
+
+\section{Garbage Collection}
 

### 2012-02-19 Paper.tex
More on concurrency issues

@@
 \item Garbage collection can be implemented like in a single threaded process.
-\item Critical sections may simply be protected by disabling the scheduler.
+\item Critical sections may simply be protected by disabling the scheduler until the code is clear of the section.
 \end{itemize}
@@
 
+If we bring garbage collection into the mix, the possibilities for corruption grow even more.
+At the point where the first thread needs to allocate the new buffer, the garbage collector could decide that it needs to run and clear some unused memory.
+It would have to traverse the object tree to mark alive objects.
+Between the mark and the sweep runs, the second thread could change the pointer to the buffer from the old one to the new one.
+So the old buffer would have been marked alive, while during the sweep run the new buffer would be in it's place, but not yet marked alive.
+
 \section{Green threads in Parrot}
@@
 
-The scheduler is implemented as a PMC (Parrot's polymorphic container type). The plan is to allow the user to subclass this PMC thus allowing very fine grained control over the scheduling policy.
+The scheduler is implemented as a PMC (Parrot's polymorphic container type).
+The plan is to allow the user to subclass this PMC thus allowing very fine grained control over the scheduling policy.
+Features a user could add this way would be for example giving different priorities to tasks or implementing the possibility to suspend and resume a task.
 

### 2012-02-21 Paper.tex
Remove manual table of contents
Feedback by NK++ and Design of Hybrid Threads

@@
 \end{german}
-
-\chapter{Table of Contents (1)}
-
-	Perl 6 Concurrency
-	About Parrot
-	Parrot threads
-	Nat "Chandon" Tuck green\_threads
 
@@
 During these years of back and forth and failed attempts of adding threading support to Parrot, the Perl 6 specification evolved to a point where the largest parts of the language were covered and its features implemented in the compilers.
-The lack of concurrency primitives in Parrot prevents however prevents any progress in the area of concurrency support.
+The lack of concurrency primitives in Parrot however prevents any progress in the area of concurrency support.
 
-Currently Parrot does not have any threading support at all. The previous, defunct implementation has been removed.
+Before my work on this paper, Parrot did not have any threading support at all. The previous, defunct implementation had been removed.
 
@@
 Contrary to other widely used virtual machines like the JVM or the CLR which are stack based, Parrot mirrors contemporary hardware CPUs more closely by being register based.
-The rationale behind giving up the simplicity of a stack based implementation is the hope of simplifying just in time compilation and improved performance of nested function and method calls.
+A stack based virtual machine usually fetches the operands of an operation from the top of a stack and puts the result back on top.
+Thus the operands chosen implicitely by ordering of the operations allowing the opcodes to be quite small.
+In a register based VM on the other hand each operation has to specify the operands explicitely.
+Compilers for stack machines are simpler because they do not have to care about register allocations and code is independent of prior or subsequent code.
+[http://static.usenix.org/events/vee05/full\_papers/p153-yunhe.pdf] The rationale behind giving up the simplicity of a stack based implementation is the hope of simplifying just in time compilation and improved performance of nested function and method calls.
 
@@
 
-There are three possible ways how two subroutines could be (at least seemingly) run in parallel:
-\begin{itemize}
-\item Coroutines
-\item Green threads 
-\item Operating system threads
-\end{itemize}
+Green threads are one way to model concurrent control flows in a program. To get a better understanding of what they are and how they work, we will first discuss the possible options.
 
@@
 \end{GenericCode}
-This function would return one integer for every call, counting from 0 to 100.
-For two functions to run in parallel, they have to cooperate.
+This function would return one integer for every call, counting from 0 to 100. An example usage would look like:
+\begin{GenericCode}
+x = counter()
+print x
+x = counter()
+print x
+\end{GenericCode}
+This example would print the numbers 0 and 1.
+As we can see, for the two functions to run in parallel, they have to cooperate.
 One by calling the other repeatedly and the callee by yielding control back to the caller.
@@
 
-Operating system threads are like multiple processes running at the same time but share their memory. The operating system is responsible for giving each of these threads CPU time to run.
+Operating system threads are like multiple processes running at the same time but share their memory.
+The operating system is responsible for giving each of these threads CPU time to run.
+Operating system threads are the only of these options where more than one thread may be executed by the CPU(s) at the same time if the hardware permits.
 
@@
 "Green threads" or "lightweight threads" are threads which are managed by the virtual machine instead of by the operating system.
+These threads only run pseudo parallel.
 The virtual machine contains a scheduler and support for preempting running tasks.
@@
 
-Using green threads we're now solving the following problems which occur when trying to implement threading support:
+This chapter describes how green threads were used to solve the following problems which occur when trying to implement threading support:
 %
@@
 \item How to ensure internal interpreter consistency when doing writes to shared data.
+\item How to implement critical sections.
 \item How to handle garbage collection.
@@
 
+In keeping the analogy of the interpreter being a software CPU, multithreading is implemented by having a separate interpreter with it's own register set for each thread.
+When a user starts a new task, the scheduler first looks for an idle thread.
+If one can be found, the task is scheduled on the thread's interpreter.
+If none can be found, a new thread with a new interpreter is started.
+Parrot tries to optimize the number of utilized threads by creating at most one for each CPU core in the system.
+If more tasks are started than the maximum number of threads, the tasks are distributed evenly among the running interpreters.
+This is effectively an implementation of the N:M threading model.
+
 \section{Shared data}
+
+As described in the introduction, cross thread writes to shared data may endanger the internal consistency of the interpreter.
+Traditionally the response to this problem is the use of locks of varying granularity.
+Fine grained locking allows code to run well parallelized but taking and releasing locks also costs performance.
+It not only increases the instruction count and memory accesses but it also forces the CPU cores to coordinate and thus communicate.
+Even a seemingly simple operation like an atomic increment may take two orders of magnitude longer than a normal increment [http://irl.cs.ucla.edu/~yingdi/paperreading/smp\_locking.pdf].
+While the gain through being able to utilize multiple CPU cores may offset this cost, it is still impacting the very common case of having only a single thread running.
+
+Too coarse locking on the other hand would reduce scalability and the performance gains through parallel execution.
+In the extreme case of having a global interpreter lock it may even effectively serialize all computations costing much of the benefits of using threads in the first place.
+
+The other problem with locking is the possibility of introducing dead locks.
+For example, two functions F1 and F2 both use two resources A and B protected by locks.
+If F1 first locks A and then tries to lock B while F2 has already locked B and is now trying to lock A, the program would come to a halt.
+Both functions would be left waiting for the other to unlock the resource which will never happen.
+With fine grained locking the possibilities for such bugs grow quickly.
+At the same time it's easy to miss a case where a lock would be appropriate leading to difficult to diagnose corruption bugs.
+
+The solution for these problems implemented for this paper is to side step them altogether by disallowing write access to shared data.
+The programmer (or in most cases the compiler) is obliged to declare a list of all shared variables before a newly created task is started.
+The interpreter would then create proxy objects for these variables which the task can use to access the data.
+These proxies contain references to the original objects.
+They use these references to forward all reading vtable functions to the originals.
+Write access on the other hand would lead to a runtime error.
+
+In other words all data is owned by the thread creating it and only the owner may write to it. Other threads have only read access.
+
+For threads to be able to communicate with their creators and other threads, they still need to write to shared data.
+This is where green threads come into play.
+Since green threads are very light weight, it is feasible for a thread to create a task just for updating a variable.
+This task would be scheduled on the interpreter owning this variable.
+To reduce latency, the task is flagged to run immediately.
+The data owning interpreter would preempt the currently running task and process the new write task.
+Put another way, the data owning interpreter is simply told what to write to it's variables, so other threads don't have to.
+
+\section{Critical Sections}
+
+In this scheme critical sections can simply be implemented by disabling preemption till the code is clear of the section.
+With other threads not being allowed to write to the data and the current task running uninterrupted it it guaranteed to complete without the data being changed from outside.
 

### 2012-02-21 Paper.tex
Some words about garbage collection

@@
 
+A garbage collector has to traverse and process the entire tree of objects in memory.
+If during this traversal another thread is changing this tree by creating new objects, moving them around in the tree or removing them, the garbage collector's data would be compromised.
+For example it could happen that the GC does not mark referenced objects alive because they were added to parents which already have been processed.
+
+In previous attempts to implement threading support in Parrot, the answer to this problem was to disable the garbage collector while multiple threads are active.
+Other VMs solve this problem by suspending all threads while the GC is running.
+
 \chapter{Implementation of Hybrid Threads (15)}

### 2012-02-23 Paper.tex
First round of fixes

@@
 On July 19th 2000, the Perl 6 design process was announced.
-Perl 5 had been a very flexible and widely used programming language but had started to show it's age and suffered from early design decisions.
+Perl 5 had been a very flexible and widely used programming language but had started to show its age and suffered from early design decisions.
 The Perl interpreter is written in C and has accumulated a lot of cruft over the years.
@@
 [http://www.developer.com/lang/perl/article.php/10940\_3076571\_3/Perl-6-and-the-Parrot-Project.htm].
-Taking up the name born in an April Fool's Day joke announcing the mergin of the Perl and Python programming languages, these new internals were called Parrot.
+Taking up the name born in an April Fool's Day joke announcing the merging of the Perl and Python programming languages, these new internals were called Parrot.
 [http://www.perl.com/pub/2001/04/01/parrot.htm]
@@
 Perl 5.8 exposed this API to the user of the programming language.
-In this new model, all data would is copied to each thread and afterwards thread local.
+In this new model, all data is copied to each thread and afterwards thread local.
 Data must be explicitely shared between threads.
@@
 
-In a dynamic language, the object may change it's class at runtime.
+In a dynamic language, the object may change its class at runtime.
 The inheritance hierarchy of the class may be changed by adding or removing parent classes.
@@
 
-Before my work on this paper, Parrot did not have any threading support at all. The previous, defunct implementation had been removed.
+Before the work on this paper, Parrot did not have any threading support at all. The previous, defunct implementation had been removed.
 
@@
 Control flow is modeled in Parrot using continuation passing.
-A continuation is a data structure containing the a state of a program at a given point in it's execution.
+A continuation is a data structure containing the a state of a program at a given point in its execution.
 It contains all information which is needed to continue a program in a certain state, e.g.
@@
 
-When a function is called, instead of pushing all this information on the stack, the function is given a return continuation as part of it's parameters.
+When a function is called, instead of pushing all this information on the stack, the function is given a return continuation as part of its parameters.
 Returning from a function means calling the return continuation with possible return values stored in the registers where the calling code is expecting them.
-Continuation passing makes things like tail call optimization \footnote{When as the last statement of a function another function is called and it's value returned unchanged, the outer function may just pass on it's return continuation thus saving the need to create a new continuation and one step on returning from the nested function} very simple and as we will see later will be a very important piece in implementing green threads.
+Continuation passing makes things like tail call optimization \footnote{When as the last statement of a function another function is called and its value returned unchanged, the outer function may just pass on its return continuation thus saving the need to create a new continuation and one step on returning from the nested function} very simple and as we will see later will be a very important piece in implementing green threads.
 
@@
 
-In a more complicated example, the array could have to resize it's data buffer to accomodate the new value.
+In a more complicated example, the array could have to resize its data buffer to accomodate the new value.
 It would read the current size, allocate a new buffer, copy the values to the new buffer and then destroying the old one.
@@
 Between the mark and the sweep runs, the second thread could change the pointer to the buffer from the old one to the new one.
-So the old buffer would have been marked alive, while during the sweep run the new buffer would be in it's place, but not yet marked alive.
+So the old buffer would have been marked alive, while during the sweep run the new buffer would be in its place, but not yet marked alive.
 
@@
 On expiry a timer callback sets a flag which is checked for at execution of every branch operation.
-Since the interpreter's state is well defined at this point, it's internal consistency is guaranteed.
+Since the interpreter's state is well defined at this point, its internal consistency is guaranteed.
 The same goes for the garbage collector.
-Since task preemption is only done while executing user level code, the GC can do it's work undisturbed and without the need for measures like locking.
+Since task preemption is only done while executing user level code, the GC can do its work undisturbed and without the need for measures like locking.
 Since user level code is allowed to disable the scheduler, it can be guaranteed to run undisturbed through critical sections.
@@
 
-In keeping the analogy of the interpreter being a software CPU, multithreading is implemented by having a separate interpreter with it's own register set for each thread.
+In keeping the analogy of the interpreter being a software CPU, multithreading is implemented by having a separate interpreter with its own register set for each thread.
 When a user starts a new task, the scheduler first looks for an idle thread.
@@
 The data owning interpreter would preempt the currently running task and process the new write task.
-Put another way, the data owning interpreter is simply told what to write to it's variables, so other threads don't have to.
+Put another way, the data owning interpreter is simply told what to write to its variables, so other threads don't have to.
 
@@
 Other VMs solve this problem by suspending all threads while the GC is running.
+While multi threaded garbage collection algorithms exist, their implementation would have been to complex and brittle for this paper [http://doc.cat-v.org/inferno/concurrent\_gc/].
 

### 2012-02-23 Paper.tex
More on GC design

@@
 
+Since cross threaded writes are already forbidden and all read access to other thread's data goes through the very narrow channel of proxy objects, forcing a complete separation of the thread's memory domains was only a small step.
+By having separate memory areas for each interpreter, it became possible to have each interpreter to run its own garbage collector.
+This way, the known to work single threaded GC implementation could be continued to be used.
+
+Separation of memory domains is a problem for objects which are only referenced from other threads.
+But since all objects which could be accessed from other threads have to be pushed onto the Task object representing such threads, they can still be referenced from there.
+
 \chapter{Implementation of Hybrid Threads (15)}

### 2012-02-23 Paper.tex
Implementation/Runloops section

@@
 
-Too coarse locking on the other hand would reduce scalability and the performance gains through parallel execution.
+Too coarse locking on the other hand would reduce scalability and the performance gains through parallel execution by having threads wait for extended periods for locks to become open.
 In the extreme case of having a global interpreter lock it may even effectively serialize all computations costing much of the benefits of using threads in the first place.
@@
 
+Runloops are created whenever a part of Parrot C internals start to execute bytecode.
+This could be simply at interpreter startup to start the actual program to execute or when some library function needs to execute a user callback.
+An exception handler could also be such a callback.
+Each runloop has a unique id.
+These ids are numbered from 0 and simply increased by 1 whenever a new runloop is created.
+
+When an exception is thrown, Parrot searches for a previously set up exception handler and executes it.
+As part of exception handling finalization, it cleans up, freeing any information no longer needed.
+The exception handler does not necessarily have to have been set up in the same runloop which the exception has been thrown from.
+Parrot uses the runloop id to identifiy such situations and free the nested runloops up to the level of the exception handler.
+
+A continuation also contains the id of the runloop in which it has been created.
+When resuming a preempted task, this id would not be the same as the one of the runloop created for executing its code.
+This would lead to Parrot not finding the correct runloop when finalizing an exception.
+
+To mitigate this, the \textit{reset\_runloop\_id\_counter(interp)} function is used to reset the global runloop id counter back to 0 when resuming a task.
+This way the task retains the same runloop id over its life time.
+
+If Parrot were to preempt a task while it is executing a nested runloop, it would have to somehow capture not only the interpreter's current state but also the C stack between the outermost runloop and the currently executing one.
+It would also have to recreate these call and runloop stacks when resuming the task.
+Otherwise the task would be ended as soon as the executing callback would be completed since instead of some C function called by an op in the bytecode, the scheduler is owner of the runloop.
+
+Thus the scheduler checks for the current runloop id and does not preempt a task if it is currently running a nested runloop.
+Furthermore even a manual yield to another task is not possible in such a situation.
+The only way to solve this problem would be to get rid of nested runloops in the interpreter.
+
 \section{The Scheduler (4)}

### 2012-02-24 Paper.tex
Document the functions in src/scheduler.c

@@
 \section{Runloops}
+\label{sec:Runloops}
 
@@
 
-\chapter{Implementation of Hybrid Threads (15)}
+\chapter{Implementation of Hybrid Threads (13)}
+
+\section{The Scheduler (4)}
+
+The scheduler is the place where most of the green thread logic is implemented. It's functions use the prefix \textit{Parrot\_cx} and are located in \textit{src/scheduler.c}.
+
+\textit{Parrot\_cx\_init\_scheduler} is called from \textit{Parrot\_interp\_initialize\_interpreter} whenever a new interpreter is created, e.g.
+on interpreter startup and whenever a new thread is created.
+It creates the scheduler PMC and sets up alarm singal handling.
+
+The scheduler is hooked into the system by replacing the call to execute the \textit{main\_sub} by a call to \textit{Parrot\_cx\_begin\_execution}.
+The latter creates the main task using \textit{main\_sub} as the task's code.
+This task is then put onto the run queue.
+The scheduler timer gets enabled and control given to \textit{Parrot\_cx\_outer\_runloop}.
+
+\textit{Parrot\_cx\_outer\_runloop} is the core loop taking tasks from the run queue and executing them.
+Despite its name, it should not be confused with runloops discussed in section~\ref{sec:Runloops}.
+\textit{Parrot\_cx\_outer\_runloop} actually consists of two nested loops.
+The inner loop simply fetches tasks and executes as long as tasks are available.
+Even when all tasks are finished there may still be alarms pending which upon expiry would trigger new tasks to be scheduled.
+For this reason there's an outer loop checking for pending alarms and if there are putting the thread to sleep until the next alarm expires.
+
+\textit{Parrot\_cx\_next\_task} contains the code to take the next task from the run queue and executing it.
+Before calling the task, it checks if there are other tasks still in the queue.
+Only if other tasks are waiting, task preemption gets enabled.
+Otherwise the current task would run until it finishes by itself, schedules other tasks or an alarm expires.
+this is an important optimization reducing the runtime overhead of green threads in the important single tasking case essentially to 0.
+
+To enable task preemption, \textit{Parrot\_cx\_enable\_preemption} sets a flag on the scheduler PMC and uses \textit{Parrot\_cx\_set\_scheduler\_alarm} to set an alarm at a point \textit{PARROT\_TASK\_SWITCH\_QUANTUM} milliseconds in the future.
+
+When an alarm is triggered, Parrot only increments the global \textit{alarm\_serial}.
+This serial is checked for changes by \textit{Parrot\_alarm\_check} which is called by \textit{Parrot\_cx\_check\_scheduler}.
+This function in turn is called once for every \textit{branch} op.
+When the serial is different from the last check or the \textit{SCHEDULER\_wake\_requested} flag is set, \textit{Parrot\_cx\_check\_scheduler} wakes the scheduler by calling \textit{Parrot\_cx\_run\_scheduler}.
+
+Since POSIX alarms can only be delivered to the main thread of a process, \textit{Parrot\_cx\_run\_scheduler} uses \textit{Parrot\_thread\_notify\_threads} to relay alarms to other thread's schedulers.
+For this purpose every thread has a pipe which \textit{Parrot\_thread\_notify\_threads} writes a dummy byte to.
+The next step for \textit{Parrot\_cx\_run\_scheduler} is to check for any expired alarms using \textit{Parrot\_cx\_check\_alarms}.
+An alarm contains a task as callback.
+On expiry this callback gets scheduled for immediate execution.
+As alarm signals do not contain any information about their origin, the scheduler does not know if the alarm was scheduled by user code or if it's the scheduler's preemption alarm.
+\textit{Parrot\_cx\_check\_quantum} is therefore used to check if the current task has used its assigned quantum and if it should be preempted.
+
+Preemption of a task is implemented in \textit{Parrot\_cx\_preempt\_task}.
+It uses \textit{Parrot\_cx\_stop\_task} to create a continuation at the current point of execution and store it as the task's code.
+The task itself is then appended at the end of the run queue.
+The NULL opcode return value is propagated all the way up through the stack back to the \textit{branch} op checking for alarms.
+The op then recognizes this as the signal to stop processing and to exit the runloop.
+
+The control flow ends up back at \textit{Parrot\_cx\_next\_task} at the point after executing the task at which it returns to \textit{Parrot\_cx\_outer\_runloop}.
+
+The \textit{schedule} op code is used to schedule a new task from user code.
+It uses \textit{Parrot\_cx\_schedule\_task} which starts new worker threads if needed and possible and pushes the task on the target thread's scheduler's task queue.
+If the target thread previously was executing only a single task, its preemption has been disabled for optimization as described above.
+So in this case \textit{Parrot\_cx\_schedule\_task} has to enable preemption to give the new task a chance to run.
+
+\textit{Parrot\_cx\_schedule\_immediate} is used in various places like at alarm expiry for putting a task to the head of the run queue and immediately causing preemption of the currently running task.
+It does this by setting \textit{SCHEDULER\_wake\_requested} and \textit{SCHEDULER\_resched\_requested} flags the same as when the preemption alarm expires.
+To be precise this mechanism leads to the current task being preempted at the next \textit{branch} op since this is the place when the mentioned flags will be checked.
+
+Alarms can be registered using \textit{Parrot\_cx\_schedule\_alarm}.
+It puts the alarm in the appropriate place in the ordered alarm list and uses \textit{Parrot\_alarm\_set} which sets the actual alarm but only if there's not another alarm already set for an earlier time since there can only be one POSIX alarm pending at any time.
+
+\textit{Parrot\_cx\_schedule\_sleep} is the actual implementation of the \textit{sleep} op.
+Like in\textit{Parrot\_cx\_preempt\_task}, \textit{Parrot\_cx\_stop\_task} is used to get an updated task at the current execution position, but instead of pushing this on the run queue, it is used as a callback for a newly set alarm.
+Again the NULL opcode is used to stop processing of the current task.
 
@@
 
-Runloops are created whenever a part of Parrot C internals start to execute bytecode.
+Runloops are created whenever a part of Parrot's C internals starts to execute bytecode.
 This could be simply at interpreter startup to start the actual program to execute or when some library function needs to execute a user callback.
@@
 
-To mitigate this, the \textit{reset\_runloop\_id\_counter(interp)} function is used to reset the global runloop id counter back to 0 when resuming a task.
+To mitigate this, the \textit{reset\_runloop\_id\_counter(interp)} function is used by \textit{Parrot\_cx\_outer\_runloop} to reset the global runloop id counter back to 0 when resuming a task.
 This way the task retains the same runloop id over its life time.
@@
 It would also have to recreate these call and runloop stacks when resuming the task.
-Otherwise the task would be ended as soon as the executing callback would be completed since instead of some C function called by an op in the bytecode, the scheduler is owner of the runloop.
+Otherwise the task would be ended as soon as the executing callback would be completed since instead of some C function called by an op in the bytecode, the scheduler is the owner of the runloop.
 
@@
 
-\section{The Scheduler (4)}
-
-\section{Threads (5)}
+\section{Threads (8)}
 
@@
 \begin{itemize}
-\item replace thread\_data->interp\_lock usage by preallocated proxy objects
-\item allow threads to create threads
-\item allow multi level proxying (don't create a proxy for a proxy, unpack first)
+\item Windows implementation for timer/alarm handling
+\item Replace thread\_data->interp\_lock usage by preallocated proxy objects
+\item Allow threads to create threads
+\item Allow multi level proxying (don't create a proxy for a proxy, unpack first)
+\item Allow redistribution of tasks to other threads
 \end{itemize}

### 2012-02-25 Paper.tex
Scheduler PMC

@@
 
+This chapter is talking about programming platforms.
+A platform is seen as a combination of a programming language and a runtime.
+For example for the Python programming language there are multiple runtimes with different implementations of threading support.
+
 \section{Java}
@@
 Thus an \textit{inc \$P0} instruction is calling the \textit{increment} vtable function of the PMC contained in the \textit{\$P0} register.
+
+PMCs are implemented in .pmc files with method bodies written in C. These files get preprocessed to plain C before compliation.
 
@@
 
-\section{The Scheduler (4)}
+\section{The Scheduler (3)}
 
-The scheduler is the place where most of the green thread logic is implemented. It's functions use the prefix \textit{Parrot\_cx} and are located in \textit{src/scheduler.c}.
+The scheduler is the place where most of the green thread logic is implemented.
+It consists of two parts: functions using the prefix \textit{Parrot\_cx} which are located in \textit{src/scheduler.c} and the scheduler PMC type containing the scheduler's data and public interface.
 
@@
 
+\section{Scheduler PMC (1)}
+
+The scheduler PMC contains the following attributes:
+%
+\begin{itemize}
+\item PMC *task\_queue: List of tasks/green threads waiting to run
+\item Parrot\_mutex task\_queue\_lock: a lock protecting the task\_queue so other threads might access it safely
+\item PMC *alarms: List of future alarms ordered by time
+\item PMC *all\_tasks: Hash of all active tasks by ID
+\item UINTVAL next\_task\_id: ID to assign to the next created task
+\item Parrot\_Interp interp: A reference to the scheduler's interpreter.
+\end{itemize}
+
+\textit{push\_pmc} and \textit{unshift\_pmc} are used to add a task at the end respectively the beginning of the task\_queue.
+\textit{shift\_pmc} is used to fetch and remove a task from the tip of the task\_queue.
+\textit{get\_integer} returns the number of tasks in the task\_queue.
+These four methods use the task\_queue\_lock to make accessing the task\_queue thread safe.
+
+\textit{alarms} is list containing all pending alarms, sorted by their expiry time.
+Sorted insertion is used to keep alarms in order.
+Sorting makes it simple and efficient to retrieve all alarms that have expired at a certain point in time.
+
+The textit{active\_tasks} returns an array containing all tasks which have been run at least once and are not yet finished, e.g. they are currently being executed or are preempted.
+
+\section{Task PMC}
+
 \section{Runloops (1)}

### 2012-02-25 Paper.tex
Task PMC

@@
 The scheduler is the place where most of the green thread logic is implemented.
-It consists of two parts: functions using the prefix \textit{Parrot\_cx} which are located in \textit{src/scheduler.c} and the scheduler PMC type containing the scheduler's data and public interface.
+It consists of two parts: functions using the prefix \textit{Parrot\_cx} which are located in \textit{src/scheduler.c} and the Scheduler PMC type containing the scheduler's data and public interface.
 
@@
 
-The scheduler PMC contains the following attributes:
+The Scheduler PMC contains the following attributes:
 %
@@
 \textit{push\_pmc} and \textit{unshift\_pmc} are used to add a task at the end respectively the beginning of the task\_queue.
-\textit{shift\_pmc} is used to fetch and remove a task from the tip of the task\_queue.
-\textit{get\_integer} returns the number of tasks in the task\_queue.
-These four methods use the task\_queue\_lock to make accessing the task\_queue thread safe.
+\textit{shift\_pmc} is used to fetch and remove a task from the tip of the \textit{task\_queue}.
+\textit{get\_integer} returns the number of tasks in the \textit{task\_queue}.
+These four methods use the \textit{task\_queue\_lock} to make accessing the \textit{task\_queue} thread safe.
+Thus the Scheduler PMC acts as a container while using \textit{task\_queue} to actually store the data.
 
@@
 
-The textit{active\_tasks} returns an array containing all tasks which have been run at least once and are not yet finished, e.g. they are currently being executed or are preempted.
+The \textit{active\_tasks} returns an array containing all tasks which have been run at least once and are not yet finished, e.g. they are currently being executed or are preempted.
 
 \section{Task PMC}
+
+The Task PMC contains the following attributes:
+%
+\begin{itemize}
+\item UINTVAL id: Unique identifier for the task
+\item FLOATVAL birthtime: The creation time stamp of the task
+\item Parrot\_Interp interp: The interpreter which created the task
+\item PMC *code: The code to run
+\item PMC *data: Additional data for the task given as parameter to \textit{code}
+\item INTVAL killed: Flag marking killed tasks
+\item PMC *waiters: Tasks waiting on this one
+\item PMC *shared: List of variables shared with this task
+\item PMC *partner: Copy of this task on the other side of a GC barrier, meaning in another thread
+\end{itemize}
+
+The \textit{invoke} method is used to run the task.
+It first checks the \textit{killed} flag to see if the task has been killed while waiting in the task queue.
+If it's still alive, the current recursion depth is saved to a local variable.
+This is a workaround needed because \textit{Parrot\_Sub\_invoke} which ends up being called when invoking the task's \textit{code} increments the recursion depth, but it does not get decremented anymore in case the task got preempted.
+This would lead to an ``recursion depth exceeded'' exception when many tasks are invoked during a program's runtime.
+The next steps are to add the task to the list of active tasks and to invoke the task's \textit{code}.
+\textit{data} is an optional parameter to the subroutine given as \textit{code}.
+Since this may be a compound object, a single parameter can cover all use cases.
+
+If the task has been killed while being in the task queue or while running or it just ended, it gets removed from the list of active tasks and all tasks registered with this task's \textit{waiters} array are added to the scheduler's task queue to be run.
+
+\textit{push\_pmc} and \textit{pop\_pmc} are used to add or remove variables to the task's list of shared data. For these variables, proxies will be created when the task is run on a different thread.
+
+The \textit{kill} method is used to set the task's \textit{killed} flag.
 

### 2012-02-25 Paper.tex
Thread creation

@@
 
+Each thread is represented by an instance of the ParrotInterpreter PMC.
+These interpreters are kept in the \textit{threads\_array} defined in \textit{src/thread.c}.
+It's a simple C array instead of one of Parrot's dynamic arrays which are implemented as PMCs because though ParrotInterpreter is implemented as a PMC itself, it's not fully garbage collected because of the bootstrapping issues this would create.
+
+\textit{src/thread.c} also contains thread management functions all prefixed by \textit{Parrot\_thread\_}.
+Threads are created when \textit{Parrot\_cx\_schedule\_task} determines that all existing threads are busy and new threads may yet be started.
+It uses \textit{Parrot\_thread\_create} which creates a new interpreter by cloning the current one but with a cleared out \textit{thread\_data} structure.
+It also creates the \textit{thread\_data->notifierfd} pipe used to notify threads of expired alerts.
+
+The task is then scheduled with the new interpreter's scheduler using \textit{Parrot\_thread\_schedule\_task}.
+This function uses \textit{Parrot\_thread\_create\_local\_task} to create a corresponding task on the new interpreter and then simply pushes it onto the scheduler.
+Since each interpreter has its own garbage collector, all objects used on the interpreter must origin from this GC's memory pools.
+That's why \textit{Parrot\_thread\_create\_local\_task} creates a new \textit{local\_task}.
+Proxies are created for the task's \textit{code} and \textit{data} attributes.
+The task object of the originating interpreter and the new \textit{local\_task} are linked using their \textit{partner} attributes which just contains a pointer to the other object.
+This is the point where for all variables in the original task's \textit{shared} array proxies are created an put into the \textit{shared} array of the \textit{local\_task}.
+\textit{Parrot\_thread\_create\_proxy} is used for this purpose.
+
+\textit{Parrot\_thread\_insert\_thread} is used to put the new interpreter into the \textit{threads\_array}.
+
+\textit{Parrot\_thread\_run} is the function where the actual operating system thread is created.
+It uses macros defined in the \textit{include/parrot/thr\_*.h} files which act as an OS abstraction layer.
+
+\textit{Parrot\_thread\_outer\_runloop} is used as the thread's main function.
+It's very similiar to \textit{Parrot\_cx\_outer\_runloop}.
+The most important difference is the way it waits for pending alarms in case no tasks are to be run.
+While \textit{Parrot\_cx\_outer\_runloop} waits for an alarm signal to arrive, \textit{Parrot\_thread\_outer\_runloop} waits for a dummy write to be written to its \textit{notifierfd} pipe by \textit{Parrot\_cx\_run\_scheduler}.
+In addition it's holding the \textit{interp->thread\_data->interp\_lock} lock while executing a task.
+This is the lock which \textit{Parrot\_cx\_schedule\_task} is taking during its call to \textit{Parrot\_thread\_schedule\_task} because the latter needs to create new objects local to the thread.
+As this could cause garbage collection to be triggered, the thread itself should not be executing code at the same time.
+
 \subsection{Proxies (2)}

### 2012-02-25 Paper.tex
A few words on how to write to shared data

@@
 \chapter{Design of Hybrid Threads (3)}
+\label{cha:design}
 
@@
 \subsection{Creation (1)}
+\label{subsec:thread_creation}
 
@@
 \textit{Parrot\_thread\_run} is the function where the actual operating system thread is created.
-It uses macros defined in the \textit{include/parrot/thr\_*.h} files which act as an OS abstraction layer.
+It uses macros defined in the \textit{include/parrot/thr\_*.h} include files which act as an OS abstraction layer.
 
@@
 The most important difference is the way it waits for pending alarms in case no tasks are to be run.
-While \textit{Parrot\_cx\_outer\_runloop} waits for an alarm signal to arrive, \textit{Parrot\_thread\_outer\_runloop} waits for a dummy write to be written to its \textit{notifierfd} pipe by \textit{Parrot\_cx\_run\_scheduler}.
+While \textit{Parrot\_cx\_outer\_runloop} waits for an alarm signal to arrive, \textit{Parrot\_thread\_outer\_runloop} waits for a dummy to be written to its \textit{notifierfd} pipe by \textit{Parrot\_cx\_run\_scheduler}.
 In addition it's holding the \textit{interp->thread\_data->interp\_lock} lock while executing a task.
@@
 
+Proxies are the arbiters between threads. They are the only means for a thread to access another thread's data and are implemented by the \textit{Proxy} PMC type.
+
 \subsection{Writing shared data (2)}
+
+As described in chapter~\ref{cha:design} to write to shared data, a thread creates a small task and schedules it on the data owning interpreter. Such a task could look like:
+%
+\begin{GenericCode}
+.sub write_to_variable
+    .param pmc variable
+    variable = 1
+.end
+\end{GenericCode}
+%
+This is a subroutine with just one parameter.
+The variable passed in as this parameter is the one the task should write to.
+In this case the constant value 1 would be written to the variable.
+In PIR an assignment to a PMC gets translated as a method call.
+In this case the \textit{set\_integer\_native} would be called changing the variable's value.
+Since PMCs get passed by reference, it's the original variable which gets written to.
+
+The code to create the task would look like:
+%
+\begin{GenericCode}
+    write_task = new ['Task']
+    setattribute write_task, 'code', write_to_variable
+    setattribute write_task, 'data', shared_variable
+    interp.'schedule_proxied'(write_task, shared_variable)
+\end{GenericCode}
+%
+Line 1 simply creates a new Task object.
+The example subroutine is used for the task's \textit{code} attribute.
+\textit{shared\_variable} is used for \textit{data}.
+At this point \textit{shared\_variable} is actually the proxy object created for the shared integer PMC.
+The interpreter object contains a \textit{schedule\_proxied} method which is used to schedule the \textit{write\_task} on the thread owning the original variable.
+This owner thread's interpreter cannot be used directly for scheduling the task, since it would have to be stored in a register to be accessible to PIR code.
+But then the same problem as described in subsection~\ref{subsec:thread_creation} would occur with the ParrotInterpreter PMC tripping up the garbage collector.
 

### 2012-02-25 Paper.tex
A few words on proxies

@@
 \section{PMCs}
+\label{sec:PMCs}
 
@@
 
-Proxies are the arbiters between threads. They are the only means for a thread to access another thread's data and are implemented by the \textit{Proxy} PMC type.
+Proxies are the arbiters between threads.
+They are the only means for a thread to access another thread's data and are implemented by the \textit{Proxy} PMC type.
+This type has only to attributes, none of which are garbage collected:
+%
+\begin{itemize}
+\item PMC *target: The PMC this object proxies to
+\item Parrot\_Interp interp: The interpreter owning target
+\end{itemize}
+
+As described in section~\ref{sec:PMCs} PMC source files to get preprocessed before compilation by a C compiler.
+This preprocessing gets done by the \textit{tools/build/pmc2c.pl} Perl script.
+This script got extended by implementing the \textit{Parrot::Pmc2c::PMC::Proxy} module which creates default implementations for all vtable functions not otherwise defined in the \textit{proxy.pmc} file.
+The default implementation for all writing functions would just call \textit{cant\_do\_write\_method} which creates a runtime exception.
+All other methods call the vtable method of the same name on the \textit{target} passing \textit{interp} as the current interpreter and all other parameters unchanged.
+If the method returns a PMC, another proxy object is created and wrapped around it so it may be safely returned to the caller.
 

### 2012-02-25 Paper.tex
Describe schedule_proxied

@@
 
-As described in chapter~\ref{cha:design} to write to shared data, a thread creates a small task and schedules it on the data owning interpreter. Such a task could look like:
+As described in chapter~\ref{cha:design} to write to shared data, a thread creates a small task and schedules it on the data owning interpreter. An example task could look like this:
 %
@@
 
+\textit{schedule\_proxied} uses \textit{Parrot\_thread\_create\_local\_task} which in this case detects that the \textit{data} given as parameter for the task's \textit{code} is actually a proxy already and unwraps the proxied object.
+\textit{Parrot\_cx\_schedule\_immediate} is then used to make the data owning interpreter execute the task as soon as possible.
+
 \subsection{Garbage Collection (3)}

### 2012-02-25 Paper.tex
Some words about garbage collection implementation

@@
 By having separate memory areas for each interpreter, it became possible to have each interpreter to run its own garbage collector.
-This way, the known to work single threaded GC implementation could be continued to be used.
+This way, the known to work, single threaded GC implementation could be used nearly unchanged.
 
@@
 
+Since each interpreter has its own garbage collector and consequenty its own memory areas and may only access data directly which it owns, the garbage collector can simply act as if there were no other threads.
+Thus the main work has been to clean up the interpreter structure itself and to replace references to global data by proxies.
+Only a few special cases had to be built into the GC itself.
+The global variable PMCNULL is used in many places in the code, but should be handled only by the main thread's garbage collector.
+
+Before the implementation of threading support, having multiple interpreter objects in the system was used to implement things like a safe compartment in which a code could be run in a restricted environment.
+For such cases it was useful to share the garbage collector between interpreters.
+This special handling had to be removed.
+
+Only the main thread is allowed to load additional bytecode or create new classes since it owns the data structures used to manage these things.
+Child threads use these via proxies.
+In previous implementation attempts code segments, classes and vtables have been copied for each new interpreter making thread startup a very costly operation while still causing subtle problems.
+
+Code segments and vtables are only marked by the main thread's garbage collector.
+
 \chapter{Tests and Benchmarks (4)}

### 2012-02-25 Paper.tex
A bit on history

@@
 \section{Historical development (1)}
+
+The remainder of the previous threading implementation has been removed with the merging of the kill\_threads branch on September, 21st 2011.
+Previously much threading related code has been removed to clean up the code and improve performance.
+Since the existing threading support was known to be unreliable and seriously flawed, this was no trade off.
+
+A year before, Nat Tuck began working on a green\_threads branch during his Google Summer of Code internship.
+The feature got prototyped using pure PIR and then implemented in Parrot's core.
+He got it to work in simple cases and started to work on operating system thread support but the internship ended before the code was ready to be merged into the master branch.
+The code lay dormant until the work on this paper started about a year later.
 

### 2012-02-26 Paper.tex
Second round of language fixes.

@@
 
-Nat "Chandon" Tuck for implementing green thread basics on which my work is based.
+Nat "Chandon" Tuck for implementing green thread basics on which my work is based on.
 
@@
 To utilize a modern CPU's power, code needs to be run in parallel.
-In UNIX (and thus Perl) tradition, this is accomplished using multiple processes which indeed is a good solution for many use cases.
+In UNIX (and thus Perl) tradition, this is accomplished using multiple processes being a good solution for many use cases.
 For many others like Perl 6's auto threading of hyper operators, the cost of process setup and communication would be prohibitively high except for very large data sets.
@@
 Thus calling a method on an object contains just the steps of determining the object's class, fetching the required method from this class and calling it.
-Calling the same method again may then even omit the first two steps since their results cannot change.
+Calling the same method again can then even omit the first two steps since their results cannot change.
 
@@
 Methods may be added to or removed from classes (or objects) at runtime and even the way to find a method of a class may change.
-So a simple method call may result in the following steps:
+So a simple method call results in the following steps:
 %
@@
 Parrot consists of the virtual machine (also called interpreter), and various tools to facilitate the implementation of programming languages on top of the Parrot VM (the Parrot Compiler Toolkit).
-In this paper, we concentrate on the virtual machine itself.
+This paper concentrates on the virtual machine itself.
 The interpreter itself is written in C.
-Example code and testcases are written in PIR (Parrot Intermediate Representation) a high level assembly language which abstracts away register allocations and function calling conventions.
+Example code and testcases are written in PIR (Parrot Intermediate Representation) a high level assembly language abstracting away register allocations and function calling conventions.
 
@@
 For example the  \textit{get\_integer} function returns an integer value for the data type.
-For a simple \textit{int} it's the contained value.
+For a simple \textit{int} it is the contained value.
 For an array it may be the number of elements.
@@
 A continuation is a data structure containing the a state of a program at a given point in its execution.
-It contains all information which is needed to continue a program in a certain state, e.g.
+It contains all information needed to continue a program in a certain state, e.g.
 a call stack, the instruction pointer and contents of local variables.
@@
 Returning from a function means calling the return continuation with possible return values stored in the registers where the calling code is expecting them.
-Continuation passing makes things like tail call optimization \footnote{When as the last statement of a function another function is called and its value returned unchanged, the outer function may just pass on its return continuation thus saving the need to create a new continuation and one step on returning from the nested function} very simple and as we will see later will be a very important piece in implementing green threads.
+Continuation passing makes things like tail call optimization \footnote{When as the last statement of a function another function is called and its value returned unchanged, the outer function can just pass on its return continuation thus saving the need to create a new continuation and one step on returning from the nested function} very simple and as described in chapter~\ref{cha:green_threads} is a very important piece in implementing green threads.
 
@@
 When some operation in Parrot bytecode is calling a C function and this C function in turn is again executing Parrot bytecode, a nested runloop is started.
-Examples for such situations include calling a library function with a callback as parameter and exceptions thrown inside Parrot's C code which call a previously defined exception handler in user code.
+Examples for such situations include calling a library function with a callback as parameter and exceptions thrown inside Parrot's C code calling a previously defined exception handler in user code.
 
@@
 \begin{itemize}
-\item Inf: a GC for debugging purposes which never collects any garbage
+\item Inf: a GC for debugging purposes never collecting any garbage
 \item MS: a basic mark and sweep implementation
@@
 In the mark phase, starting from a known root set of objects, the GC follows pointers down the object tree, marking each encountered object as alive.
-In the sweep phase, all objects which are not alive are destroyed and the live flags reset.
+In the sweep phase, all objects not being alive are destroyed and the live flags reset.
 
@@
 \chapter{Green threads (3)}
+\label{cha:green_threads}
 
-Green threads are one way to model concurrent control flows in a program. To get a better understanding of what they are and how they work, we will first discuss the possible options.
+Green threads are one way to model concurrent control flows in a program. To get a better understanding of what they are and how they work, the possible options are discussed first.
 
@@
 
-Coroutines are functions which retain their state between calls.
+Coroutines are functions retaining their state between calls.
 Instead of returning, they yield control back to the calling function, possibly with an intermediate result returned.
-A very simple example in Python could look like:
+A very simple example in Python looks like:
 \begin{GenericCode}
@@
 This example would print the numbers 0 and 1.
-As we can see, for the two functions to run in parallel, they have to cooperate.
+For the two functions to run in parallel, they have to cooperate.
 One by calling the other repeatedly and the callee by yielding control back to the caller.
@@
 
-"Green threads" or "lightweight threads" are threads which are managed by the virtual machine instead of by the operating system.
+"Green threads" or "lightweight threads" are threads managed by the virtual machine instead of by the operating system.
 These threads only run pseudo parallel.
 The virtual machine contains a scheduler and support for preempting running tasks.
-Preemption means to suspend a task regardless of what it's currently doing and probably resuming it later on.
+Preemption means to suspend a task regardless of what it is currently doing and probably resuming it later on.
 This is an implementation of the many-to-one threading model very similiar to what an operating system running on a system with a single CPU core does.
@@
 \item They do not depend on OS threading support
-\item The interpreter controls the point at which a green thread is preempted.
+\item The interpreter controls the point of preemption of a green thread.
 \item The interpreter controls the exact scheduling policy and may allow the user to influence it or even take over completely.
 \item Garbage collection can be implemented like in a single threaded process.
-\item Critical sections may simply be protected by disabling the scheduler until the code is clear of the section.
+\item Critical sections can simply be protected by disabling the scheduler until the code is clear of the section.
 \end{itemize}
@@
 
-Let's look at a simple example to explain why concurrent processing could endanger the interpreter's internal consistency.
+Let's look at a simple example to explain why concurrent processing can endanger the interpreter's internal consistency.
 The basic problem is concurrent writes to shared data.
-Assume we have an implementation of an array class consisting of the field holding the data and the number of contained elements in a seperate member variable:
+Assumeing an implementation of an array class consisting of the field holding the data and the number of contained elements in a seperate member variable:
 \begin{CCode}
@@
 \end{CCode}
-To append a new value, the array would have to read the current size, write the new value at the position size + 1 and then write the incremented size back into the member variable.
+to append a new value, the array would have to read the current size, write the new value at the position size + 1 and then write the incremented size back into the member variable.
 Now if two threads simultainously try to do this, it may happen that both read the same size, write to the same position (with one overwriting the value of the other) and write back the same incremented size.
@@
 
-If we bring garbage collection into the mix, the possibilities for corruption grow even more.
+When garbage collection is brought into the mix, the possibilities for corruption grow even more.
 At the point where the first thread needs to allocate the new buffer, the garbage collector could decide that it needs to run and clear some unused memory.
@@
 
-This chapter describes how green threads were used to solve the following problems which occur when trying to implement threading support:
+This chapter describes how green threads were used to solve the following problems occuring when trying to implement threading support:
 %
@@
 It not only increases the instruction count and memory accesses but it also forces the CPU cores to coordinate and thus communicate.
-Even a seemingly simple operation like an atomic increment may take two orders of magnitude longer than a normal increment [http://irl.cs.ucla.edu/~yingdi/paperreading/smp\_locking.pdf].
+Even a seemingly simple operation like an atomic increment can take two orders of magnitude longer than a normal increment [http://irl.cs.ucla.edu/~yingdi/paperreading/smp\_locking.pdf].
 While the gain through being able to utilize multiple CPU cores may offset this cost, it is still impacting the very common case of having only a single thread running.
@@
 Too coarse locking on the other hand would reduce scalability and the performance gains through parallel execution by having threads wait for extended periods for locks to become open.
-In the extreme case of having a global interpreter lock it may even effectively serialize all computations costing much of the benefits of using threads in the first place.
+In the extreme case of having a global interpreter lock it would even effectively serialize all computations costing much of the benefits of using threads in the first place.
 
@@
 With fine grained locking the possibilities for such bugs grow quickly.
-At the same time it's easy to miss a case where a lock would be appropriate leading to difficult to diagnose corruption bugs.
+At the same time it is easy to miss a case where a lock would be appropriate leading to difficult to diagnose corruption bugs.
 
@@
 If during this traversal another thread is changing this tree by creating new objects, moving them around in the tree or removing them, the garbage collector's data would be compromised.
-For example it could happen that the GC does not mark referenced objects alive because they were added to parents which already have been processed.
+For example it could happen that the GC does not mark referenced objects alive because they were added to parents having already been processed.
 
@@
 By having separate memory areas for each interpreter, it became possible to have each interpreter to run its own garbage collector.
-This way, the known to work, single threaded GC implementation could be used nearly unchanged.
+This way, the known to work, single threaded GC implementation can be used nearly unchanged.
 
@@
 Otherwise the current task would run until it finishes by itself, schedules other tasks or an alarm expires.
-this is an important optimization reducing the runtime overhead of green threads in the important single tasking case essentially to 0.
+this is an optimization reducing the runtime overhead of green threads in the important single tasking case to 0.
 
@@
 When an alarm is triggered, Parrot only increments the global \textit{alarm\_serial}.
-This serial is checked for changes by \textit{Parrot\_alarm\_check} which is called by \textit{Parrot\_cx\_check\_scheduler}.
+This serial is checked for changes by \textit{Parrot\_alarm\_check} being called by \textit{Parrot\_cx\_check\_scheduler}.
 This function in turn is called once for every \textit{branch} op.
@@
 On expiry this callback gets scheduled for immediate execution.
-As alarm signals do not contain any information about their origin, the scheduler does not know if the alarm was scheduled by user code or if it's the scheduler's preemption alarm.
+As alarm signals do not contain any information about their origin, the scheduler does not know if the alarm was scheduled by user code or if it is the scheduler's preemption alarm.
 \textit{Parrot\_cx\_check\_quantum} is therefore used to check if the current task has used its assigned quantum and if it should be preempted.
@@
 
-The control flow ends up back at \textit{Parrot\_cx\_next\_task} at the point after executing the task at which it returns to \textit{Parrot\_cx\_outer\_runloop}.
+The control flow ends up back at \textit{Parrot\_cx\_next\_task} at the point after executing the task. \textit{Parrot\_cx\_next\_task} then returns to \textit{Parrot\_cx\_outer\_runloop}.
 
@@
 \item PMC *task\_queue: List of tasks/green threads waiting to run
-\item Parrot\_mutex task\_queue\_lock: a lock protecting the task\_queue so other threads might access it safely
+\item Parrot\_mutex task\_queue\_lock: a lock protecting the task\_queue so other threads can access it safely
 \item PMC *alarms: List of future alarms ordered by time
@@
 It first checks the \textit{killed} flag to see if the task has been killed while waiting in the task queue.
-If it's still alive, the current recursion depth is saved to a local variable.
-This is a workaround needed because \textit{Parrot\_Sub\_invoke} which ends up being called when invoking the task's \textit{code} increments the recursion depth, but it does not get decremented anymore in case the task got preempted.
+If it is still alive, the current recursion depth is saved to a local variable.
+This is a workaround needed because \textit{Parrot\_Sub\_invoke}, which ends up being called when invoking the task's \textit{code}, increments the recursion depth, but it does not get decremented anymore in case the task got preempted.
 This would lead to an ``recursion depth exceeded'' exception when many tasks are invoked during a program's runtime.
@@
 \textit{data} is an optional parameter to the subroutine given as \textit{code}.
-Since this may be a compound object, a single parameter can cover all use cases.
+Since this can be a compound object, a single parameter can cover all use cases.
 
@@
 Runloops are created whenever a part of Parrot's C internals starts to execute bytecode.
-This could be simply at interpreter startup to start the actual program to execute or when some library function needs to execute a user callback.
-An exception handler could also be such a callback.
+This can be simply at interpreter startup to start the actual program to execute or when some library function needs to execute a user callback.
+An exception handler can also be such a callback.
 Each runloop has a unique id.
@@
 As part of exception handling finalization, it cleans up, freeing any information no longer needed.
-The exception handler does not necessarily have to have been set up in the same runloop which the exception has been thrown from.
+The exception handler does not necessarily have to have been set up in the same runloop from which the exception has been thrown.
 Parrot uses the runloop id to identifiy such situations and free the nested runloops up to the level of the exception handler.
@@
 These interpreters are kept in the \textit{threads\_array} defined in \textit{src/thread.c}.
-It's a simple C array instead of one of Parrot's dynamic arrays which are implemented as PMCs because though ParrotInterpreter is implemented as a PMC itself, it's not fully garbage collected because of the bootstrapping issues this would create.
+It's a simple C array instead of one of Parrot's dynamic arrays implemented as PMCs because though ParrotInterpreter is implemented as a PMC itself, it is not fully garbage collected because of the bootstrapping issues this would create.
 
 \textit{src/thread.c} also contains thread management functions all prefixed by \textit{Parrot\_thread\_}.
-Threads are created when \textit{Parrot\_cx\_schedule\_task} determines that all existing threads are busy and new threads may yet be started.
-It uses \textit{Parrot\_thread\_create} which creates a new interpreter by cloning the current one but with a cleared out \textit{thread\_data} structure.
+Threads are created when \textit{Parrot\_cx\_schedule\_task} determines that all existing threads are busy and new threads can yet be started.
+It uses \textit{Parrot\_thread\_create} to create a new interpreter by cloning the current one and giving it a cleared out \textit{thread\_data} structure.
 It also creates the \textit{thread\_data->notifierfd} pipe used to notify threads of expired alerts.
@@
 Proxies are created for the task's \textit{code} and \textit{data} attributes.
-The task object of the originating interpreter and the new \textit{local\_task} are linked using their \textit{partner} attributes which just contains a pointer to the other object.
+The task object of the originating interpreter and the new \textit{local\_task} are linked using their \textit{partner} attributes containing a pointer to the other object.
 This is the point where for all variables in the original task's \textit{shared} array proxies are created an put into the \textit{shared} array of the \textit{local\_task}.
@@
 While \textit{Parrot\_cx\_outer\_runloop} waits for an alarm signal to arrive, \textit{Parrot\_thread\_outer\_runloop} waits for a dummy to be written to its \textit{notifierfd} pipe by \textit{Parrot\_cx\_run\_scheduler}.
-In addition it's holding the \textit{interp->thread\_data->interp\_lock} lock while executing a task.
+In addition it is holding the \textit{interp->thread\_data->interp\_lock} lock while executing a task.
 This is the lock which \textit{Parrot\_cx\_schedule\_task} is taking during its call to \textit{Parrot\_thread\_schedule\_task} because the latter needs to create new objects local to the thread.
@@
 They are the only means for a thread to access another thread's data and are implemented by the \textit{Proxy} PMC type.
-This type has only to attributes, none of which are garbage collected:
+This type has only two attributes, none of which are garbage collected:
 %
@@
 All other methods call the vtable method of the same name on the \textit{target} passing \textit{interp} as the current interpreter and all other parameters unchanged.
-If the method returns a PMC, another proxy object is created and wrapped around it so it may be safely returned to the caller.
+If the method returns a PMC, another proxy object is created and wrapped around it so it can be safely returned to the caller.
 
@@
 
-As described in chapter~\ref{cha:design} to write to shared data, a thread creates a small task and schedules it on the data owning interpreter. An example task could look like this:
+As described in chapter~\ref{cha:design} to write to shared data, a thread creates a small task and schedules it on the data owning interpreter. An example task looks like this:
 %
@@
 In this case the \textit{set\_integer\_native} would be called changing the variable's value.
-Since PMCs get passed by reference, it's the original variable which gets written to.
+Since PMCs get passed by reference, it is the original variable which gets written to.
 
@@
 
-Since each interpreter has its own garbage collector and consequenty its own memory areas and may only access data directly which it owns, the garbage collector can simply act as if there were no other threads.
+Since each interpreter has its own garbage collector and consequenty its own memory areas and can only access its own data directly, the garbage collector can simply act as if there were no other threads.
 Thus the main work has been to clean up the interpreter structure itself and to replace references to global data by proxies.
@@
 
-Before the implementation of threading support, having multiple interpreter objects in the system was used to implement things like a safe compartment in which a code could be run in a restricted environment.
+Before the implementation of threading support, having multiple interpreter objects in the system was used to implement things like a safe compartment where code can be run in a restricted environment.
 For such cases it was useful to share the garbage collector between interpreters.

### 2012-02-26 Paper.tex
Put references into literature.bib

@@
 \documentclass[bachelor,english]{hgbthesis}
+
+\AddBibFile{literature.bib}
 
@@
 \betreuer{Dr. Heinz Dobler}
+\license{This work is published under the conditions of the Createive Commons Attribution 3.0 Unported License (CC BY) -- see \url{http://creativecommons.org/licenses/by/3.0/}}
 
@@
 
+Markus "zimmski" Zimmerman for correcting many errors.
+
 \chapter{Abstract (1)}
@@
 The Perl interpreter is written in C and has accumulated a lot of cruft over the years.
-The general consensus among the core developers was that the code had reached a state where maintenance was approaching impossibility.
-[http://use.perl.org/~masak/journal/40451].
+The general consensus among the core developers was that the code had reached a state where maintenance was approaching impossibility\cite{Masak40451}.
 An attempt to reimplement these internals had failed but led to the decision that the interpreter for a Perl 6 language should be developed independently of the needs of Perl 5.
-Since the Perl 6 syntax was very much in flux (and parts of it still are) the designers of these new internals tried to work very independently of any syntax related questions.
-[http://www.developer.com/lang/perl/article.php/10940\_3076571\_3/Perl-6-and-the-Parrot-Project.htm].
-Taking up the name born in an April Fool's Day joke announcing the merging of the Perl and Python programming languages, these new internals were called Parrot.
-[http://www.perl.com/pub/2001/04/01/parrot.htm]
+Since the Perl 6 syntax was very much in flux (and parts of it still are) the designers of these new internals tried to work very independently of any syntax related questions\cite{Developer}.
+Taking up the name born in an April Fool's Day joke announcing the merging of the Perl and Python programming languages, these new internals were called Parrot\cite{Perl}.
 
@@
 The Perl 6 design process began with asking the Perl users for what they were expecting from the new version of the language.
-The very first feature that got asked for was well integrated multi-threading support [http://dev.perl.org/perl6/rfc/1.html].
+The very first feature that got asked for was well integrated multi-threading support\cite{RFC1}.
 Perl 5 had two different implementations of thread support.
@@
 This was similiar to the models used by languages such as C or Java.
-The implementation however suffered from data corruption and crashes and thus was not recommended for production use [http://search.cpan.org/~nwclark/perl-5.8.8/lib/Thread.pm].
+The implementation however suffered from data corruption and crashes and thus was not recommended for production use\cite{ThreadManual}.
 Perl 5.6 introduced the newer model called ithreads, mostly as a way to emulate fork() on Win32 platforms.
@@
 \section{Java}
+
+In Java responsibility for preventing concurrency issues is obliged on the user code.
+The language provides synchronization primitives like mutexes, but the interpreter (the Java Virtual Machine, JVM) does not protect itself.
+The class library provides the user with high level data structures explicitely designed for multi threaded scenarios.
+
+Java version 1.1 used ``green threads'' to support multithreaded execution of Java programs.
+Green threads are threads simulated by the virtual machine but unable to use more than one CPU core for processing.
+Details are described in chapter~\ref{cha:green_threads}.
+Version 1.2 introduced native OS thread support which since has become the standard way to do multithreading in Java\cite{JavaThreadManual}.
+
 \section{Python}
+
+Python provides threading support through the \textit{threading} module.
+
+The CPython implementation of the Python runtime uses a Global Interpreter Lock (GIL) to protect its internal consistency\cite{PythonThreadingManual}.
+This is a single lock taken whenever the interpreter executes Python bytecode.
+Because of this lock, only one thread can execute bytecode at any time so all built-in types and the object model are implicitely type safe.
+The drawback is that Python code cannot benefit from having multiple CPU cores available.
+However I/O operations and calls to external libraries are done without holding the GIL, so in applications with multiple I/O bound threads, there may still be a performance benefit from using multi threading.
+
+To actually run Python code in parallel, multiple processes have to be used.
+The \textit{multiprocessing} module provides support for spawning processes using an API similar to the threading module\cite{PythonMultiProcessingManual}.
+Since processes may not dirctly access other processes' memory, the \textit{multiprocessing} module provides several means of communication between processes: \textit{Queues}, \textit{Pipes} and shared memory support.
+
 \section{Erlang}
@@
 In a register based VM on the other hand each operation has to specify the operands explicitely.
-Compilers for stack machines are simpler because they do not have to care about register allocations and code is independent of prior or subsequent code.
-[http://static.usenix.org/events/vee05/full\_papers/p153-yunhe.pdf] The rationale behind giving up the simplicity of a stack based implementation is the hope of simplifying just in time compilation and improved performance of nested function and method calls.
+Compilers for stack machines are simpler because they do not have to care about register allocations and code is independent of prior or subsequent code\cite{VMShowdown}.
+The rationale behind giving up the simplicity of a stack based implementation is the hope of simplifying just in time compilation and improved performance of nested function and method calls.
 
@@
 
-In keeping the analogy of the interpreter being a software CPU, multithreading is implemented by having a separate interpreter with its own register set for each thread.
+In keeping the analogy of the interpreter being a software CPU, multi threading is implemented by having a separate interpreter with its own register set for each thread.
 When a user starts a new task, the scheduler first looks for an idle thread.
@@
 It not only increases the instruction count and memory accesses but it also forces the CPU cores to coordinate and thus communicate.
-Even a seemingly simple operation like an atomic increment can take two orders of magnitude longer than a normal increment [http://irl.cs.ucla.edu/~yingdi/paperreading/smp\_locking.pdf].
+Even a seemingly simple operation like an atomic increment can take two orders of magnitude longer than a normal increment\cite{LockingInOSKernels}.
 While the gain through being able to utilize multiple CPU cores may offset this cost, it is still impacting the very common case of having only a single thread running.
@@
 Other VMs solve this problem by suspending all threads while the GC is running.
-While multi threaded garbage collection algorithms exist, their implementation would have been to complex and brittle for this paper [http://doc.cat-v.org/inferno/concurrent\_gc/].
+While multi threaded garbage collection algorithms exist, their implementation would have been to complex and brittle for this paper\cite{VCGC}.
 
@@
 
+To protect a critical section, preemption can be disabled so the critical section runs uninterrupted:
+%
+\begin{GenericCode}
+.sub swap_variables
+    .param pmc a, b
+    .local temp
+    disable_preemption
+    temp = a
+    a = b
+    b = temp
+    enable_preemption
+.end
+\end{GenericCode}
+
 \subsection{Garbage Collection (3)}

### 2012-02-27 Paper.tex
First feedback

@@
 
+\usepackage{wrapfig}
 \AddBibFile{literature.bib}
@@
 
+\begin{wrapfigure}{r}{0.5\textwidth}
+A simple example of Perl:
+\include{perlexample}
+\end{wrapfigure}
+%
 On July 19th 2000, the Perl 6 design process was announced.
@@
 The Perl 6 design process began with asking the Perl users for what they were expecting from the new version of the language.
-The very first feature that got asked for was well integrated multi-threading support\cite{RFC1}.
+The very first feature that got asked for was well integrated multithreading support\cite{RFC1}.
 Perl 5 had two different implementations of thread support.
@@
 Perl 5.6 introduced the newer model called ithreads, mostly as a way to emulate fork() on Win32 platforms.
-Perl 5.8 exposed this API to the user of the programming language.
+Perl 5.8 exposed this API (Application Programming Interface) to the user of the programming language.
 In this new model, all data is copied to each thread and afterwards thread local.
@@
 
-\section{Why is multi-threading support so important?}
+\section{Why is multithreading support so important?}
 
-In the year 2005 development of faster CPUs shifted from increased speed of a single core to adding more cores.
+In the year 2005 development of faster CPUs (Central Processing Units) shifted from increased speed of a single core to adding more cores.
 Modern processors contain up to 12 cores with even mobile phones having up to four.
@@
 
-\section{Why is it so difficult to implement multi-threading support in Perl or Parrot?}
+\section{Why is multithreading support so difficult to implement?}
 
@@
 
-While there are multi threaded garbage collection schemes, their implementation has not yet been attempted in Parrot.
-
 Many interpreters for dynamic languages like Python or Ruby handle this problem by using a global interpreter lock to effectively serialize all operations.
@@
 This paper suggests a new approach based on a hybrid threading system.
-So called green threads are used to simplify the implementation of a nearly lock free multi threading implementation.
+So called green threads are used to simplify the implementation of a nearly lock free multithreading implementation.
 This approach is based on a design by Andrew Whitworth and Nat Tuck.
+The goal of this paper is to demonstrate the advantages of this model and produce a working implementation which can be used to investigate the performance characteristics of a hybrid threading system.
 
@@
 The language provides synchronization primitives like mutexes, but the interpreter (the Java Virtual Machine, JVM) does not protect itself.
-The class library provides the user with high level data structures explicitely designed for multi threaded scenarios.
+The class library provides the user with high level data structures explicitely designed for multithreaded scenarios.
 
@@
 Details are described in chapter~\ref{cha:green_threads}.
-Version 1.2 introduced native OS thread support which since has become the standard way to do multithreading in Java\cite{JavaThreadManual}.
+Version 1.2 introduced native OS (Operating System) thread support which since has become the standard way to do multithreading in Java\cite{JavaThreadManual}.
 
@@
 The drawback is that Python code cannot benefit from having multiple CPU cores available.
-However I/O operations and calls to external libraries are done without holding the GIL, so in applications with multiple I/O bound threads, there may still be a performance benefit from using multi threading.
+However I/O operations and calls to external libraries are done without holding the GIL, so in applications with multiple I/O bound threads, there may still be a performance benefit from using multithreading.
 
@@
 
-Parrot consists of the virtual machine (also called interpreter), and various tools to facilitate the implementation of programming languages on top of the Parrot VM (the Parrot Compiler Toolkit).
+Parrot consists of the virtual machine (VM, also called interpreter), and various tools to facilitate the implementation of programming languages on top of the Parrot VM (the Parrot Compiler Toolkit).
 This paper concentrates on the virtual machine itself.
@@
 
-PMCs are containers for all higher level types such as objects, arrays, hash tables or code.
+PMCs (PolyMorphic Container) are containers for all higher level types such as objects, arrays, hash tables or code.
 Thus they are very similar to Python's PythonObject types.
@@
 
-\section{Garbage collector}
+\section{Garbage Collector}
 
-Parrot supports different garbage collector implementations which can be selected at interpreter startup. Currently there are four existing implementations of different algorithms:
+Parrot supports different GC (Garbage Collector) implementations which can be selected at interpreter startup. Currently there are four existing implementations of different algorithms:
 \begin{itemize}
@@
 
-Let's look at a simple example to explain why concurrent processing can endanger the interpreter's internal consistency.
+Following is a simple example to explain why concurrent processing can endanger the interpreter's internal consistency.
 The basic problem is concurrent writes to shared data.
@@
 
-The scheduler is implemented as a PMC (Parrot's polymorphic container type).
+The scheduler is implemented as a PMC (Parrot's polymorphic container) type.
 The plan is to allow the user to subclass this PMC thus allowing very fine grained control over the scheduling policy.
@@
 
-In keeping the analogy of the interpreter being a software CPU, multi threading is implemented by having a separate interpreter with its own register set for each thread.
+In keeping the analogy of the interpreter being a software CPU, multithreading is implemented by having a separate interpreter with its own register set for each thread.
 When a user starts a new task, the scheduler first looks for an idle thread.
@@
 Other VMs solve this problem by suspending all threads while the GC is running.
-While multi threaded garbage collection algorithms exist, their implementation would have been to complex and brittle for this paper\cite{VCGC}.
+While multithreaded garbage collection algorithms exist, their implementation would have been to complex and brittle for this paper\cite{VCGC}.
 

### 2012-03-08 Paper.tex
Fix a typo
dukeleto++
Another typo
tadzik++

@@
 \betreuer{Dr. Heinz Dobler}
-\license{This work is published under the conditions of the Createive Commons Attribution 3.0 Unported License (CC BY) -- see \url{http://creativecommons.org/licenses/by/3.0/}}
+\license{This work is published under the conditions of the Creative Commons Attribution 3.0 Unported License (CC BY) -- see \url{http://creativecommons.org/licenses/by/3.0/}}
 
@@
 Other VMs solve this problem by suspending all threads while the GC is running.
-While multithreaded garbage collection algorithms exist, their implementation would have been to complex and brittle for this paper\cite{VCGC}.
+While multithreaded garbage collection algorithms exist, their implementation would have been too complex and brittle for this paper\cite{VCGC}.
 

### 2012-03-08 Paper.tex
s/explicitely/explicitly/ and s/implicitely/implicitly/

@@
 Perl 5 had two different implementations of thread support.
-In the first model, called 5005threads, data was shared by default and shared access to data had to be explicitely synchronized.
+In the first model, called 5005threads, data was shared by default and shared access to data had to be explicitly synchronized.
 This was similiar to the models used by languages such as C or Java.
@@
 In this new model, all data is copied to each thread and afterwards thread local.
-Data must be explicitely shared between threads.
+Data must be explicitly shared between threads.
 In other words, in Perl threads are not lightweight at all.
@@
 The language provides synchronization primitives like mutexes, but the interpreter (the Java Virtual Machine, JVM) does not protect itself.
-The class library provides the user with high level data structures explicitely designed for multithreaded scenarios.
+The class library provides the user with high level data structures explicitly designed for multithreaded scenarios.
 
@@
 This is a single lock taken whenever the interpreter executes Python bytecode.
-Because of this lock, only one thread can execute bytecode at any time so all built-in types and the object model are implicitely type safe.
+Because of this lock, only one thread can execute bytecode at any time so all built-in types and the object model are implicitly type safe.
 The drawback is that Python code cannot benefit from having multiple CPU cores available.
@@
 A stack based virtual machine usually fetches the operands of an operation from the top of a stack and puts the result back on top.
-Thus the operands chosen implicitely by ordering of the operations allowing the opcodes to be quite small.
-In a register based VM on the other hand each operation has to specify the operands explicitely.
+Thus the operands chosen implicitly by ordering of the operations allowing the opcodes to be quite small.
+In a register based VM on the other hand each operation has to specify the operands explicitly.
 Compilers for stack machines are simpler because they do not have to care about register allocations and code is independent of prior or subsequent code\cite{VMShowdown}.

### 2012-03-08 Paper.tex
Improved Design of GC section

--- 2012-02-27 Paper.tex
@@
 In previous attempts to implement threading support in Parrot, the answer to this problem was to disable the garbage collector while multiple threads are active.
+While multithreaded garbage collection algorithms exist, their implementation would have been too complex and brittle for this paper\cite{VCGC}.
 Other VMs solve this problem by suspending all threads while the GC is running.
-While multithreaded garbage collection algorithms exist, their implementation would have been to complex and brittle for this paper\cite{VCGC}.
+Suspending all threads is not as simple as it sounds because these threads can be in any state at the time.
+For example, they could be blocking on some long running I/O operation like waiting for a reply on a network connection.
+A thread can stay in this condition indefinitely thereby never being able to confirm synchronization.
 
@@
 
-Separation of memory domains is a problem for objects which are only referenced from other threads.
-But since all objects which could be accessed from other threads have to be pushed onto the Task object representing such threads, they can still be referenced from there.
+With separated memory domains it can happen that an object is created on one thread, used on another thread but not any more on the owner thread.
+Without any additional measures, the GC would collect such objects, since it does not know about references from other threads.
+But since all objects which could be accessed from other threads have to be pushed onto the Task object representing these threads, the objects can still be referenced from the Task object.
+Since the task is still owned by the original thread, the GC knows that these objects are still in use.
 

### 2012-03-09 Paper.tex
Another typo
tadzik++

@@
 The basic problem is concurrent writes to shared data.
-Assumeing an implementation of an array class consisting of the field holding the data and the number of contained elements in a seperate member variable:
+Assuming an implementation of an array class consisting of the field holding the data and the number of contained elements in a seperate member variable:
 \begin{CCode}

### 2012-03-29 Paper.tex
Update to match the current version of the code + spelling fixes

@@
 Perl 5 had two different implementations of thread support.
-In the first model, called 5005threads, data was shared by default and shared access to data had to be explicitely synchronized.
-This was similiar to the models used by languages such as C or Java.
+In the first model, called 5005threads, data was shared by default and shared access to data had to be explicitly synchronized.
+This was similar to the models used by languages such as C or Java.
 The implementation however suffered from data corruption and crashes and thus was not recommended for production use\cite{ThreadManual}.
@@
 
-Being born at a time when Perl 6 still looked much more similiar to Perl 5 than it does nowadays, Parrot's threading support initially was very close to Perl's ithreads model.
+Being born at a time when Perl 6 still looked much more similar to Perl 5 than it does nowadays, Parrot's threading support initially was very close to Perl's ithreads model.
 Previous attempts to change this into the more conventional model of data shared by default or implementing new technologies like Software Transactional Memory failed.
@@
 Low level programming languages like C provide only the bare necessities, leaving the responsibility for preventing data corruption entirely to the user.
-A high level language like Perl 6 on the other hand offers complex and compund data types, handles garbage collection and a very dynamic object system.
-Even seeminlgy simple things like a method call can become very complex sequences.
+A high level language like Perl 6 on the other hand offers complex and compound data types, handles garbage collection and a very dynamic object system.
+Even seemingly simple things like a method call can become very complex sequences.
 In a statically typed programming language the definition of a class is immutable.
@@
 The \textit{multiprocessing} module provides support for spawning processes using an API similar to the threading module\cite{PythonMultiProcessingManual}.
-Since processes may not dirctly access other processes' memory, the \textit{multiprocessing} module provides several means of communication between processes: \textit{Queues}, \textit{Pipes} and shared memory support.
+Since processes may not directly access other processes' memory, the \textit{multiprocessing} module provides several means of communication between processes: \textit{Queues}, \textit{Pipes} and shared memory support.
 
@@
 The interpreter itself is written in C.
-Example code and testcases are written in PIR (Parrot Intermediate Representation) a high level assembly language abstracting away register allocations and function calling conventions.
+Example code and test cases are written in PIR (Parrot Intermediate Representation) a high level assembly language abstracting away register allocations and function calling conventions.
 
@@
 In short it contains a pointer to a type specific data structure and some meta data.
-The vtable is where the type's behaviour is defined.
+The vtable is where the type's behavior is defined.
 It contains a long list of function pointers forming Parrot's unified type interface.
-This interface is a union of numeric, string, array, hash and object like behaviours.
+This interface is a union of numeric, string, array, hash and object like behaviors.
 For example the  \textit{get\_integer} function returns an integer value for the data type.
@@
 
-In this way, a language implementor can define the basic types of the language and available operations on them.
+In this way, a language implementer can define the basic types of the language and available operations on them.
 For each vtable entry, there is a corresponding op code in Parrot's bytecode.
@@
 
-PMCs are implemented in .pmc files with method bodies written in C. These files get preprocessed to plain C before compliation.
+PMCs are implemented in .pmc files with method bodies written in C. These files get preprocessed to plain C before compilation.
 
@@
 Control flow is modeled in Parrot using continuation passing.
-A continuation is a data structure containing the a state of a program at a given point in its execution.
+A continuation is a data structure containing the state of a program at a given point in its execution.
 It contains all information needed to continue a program in a certain state, e.g.
@@
 "Green threads" or "lightweight threads" are threads managed by the virtual machine instead of by the operating system.
-These threads only run pseudo parallel.
+For better distinction green threads are henceforth referred to as "tasks".
+These tasks only run pseudo parallel.
 The virtual machine contains a scheduler and support for preempting running tasks.
 Preemption means to suspend a task regardless of what it is currently doing and probably resuming it later on.
-This is an implementation of the many-to-one threading model very similiar to what an operating system running on a system with a single CPU core does.
+This is an implementation of the many-to-one threading model very similar to what an operating system running on a system with a single CPU core does.
 %
@@
 The basic problem is concurrent writes to shared data.
-Assuming an implementation of an array class consisting of the field holding the data and the number of contained elements in a seperate member variable:
+Assuming an implementation of an array class consisting of the field holding the data and the number of contained elements in a separate member variable:
 \begin{CCode}
@@
 to append a new value, the array would have to read the current size, write the new value at the position size + 1 and then write the incremented size back into the member variable.
-Now if two threads simultainously try to do this, it may happen that both read the same size, write to the same position (with one overwriting the value of the other) and write back the same incremented size.
+Now if two threads simultaneously try to do this, it may happen that both read the same size, write to the same position (with one overwriting the value of the other) and write back the same incremented size.
 In this case one of the appended values would be lost.
 
-In a more complicated example, the array could have to resize its data buffer to accomodate the new value.
+In a more complicated example, the array could have to resize its data buffer to accommodate the new value.
 It would read the current size, allocate a new buffer, copy the values to the new buffer and then destroying the old one.
-Again if two threads try to do this simultainously, one of them could still be copying data, while the other is already at the point where it destroys the old buffer.
+Again if two threads try to do this simultaneously, one of them could still be copying data, while the other is already at the point where it destroys the old buffer.
 This would lead to the copying thread accessing freed memory.
@@
 
-This chapter describes how green threads were used to solve the following problems occuring when trying to implement threading support:
+This chapter describes how green threads were used to solve the following problems occurring when trying to implement threading support:
 %
@@
 In keeping the analogy of the interpreter being a software CPU, multithreading is implemented by having a separate interpreter with its own register set for each thread.
+Thus the words ``thread'' and ``interpreter'' are used interchangeably as there is a 1:1 relationship between them.
 When a user starts a new task, the scheduler first looks for an idle thread.
@@
 on interpreter startup and whenever a new thread is created.
-It creates the scheduler PMC and sets up alarm singal handling.
+It creates the scheduler PMC and sets up alarm signal handling.
 
@@
 Otherwise the current task would run until it finishes by itself, schedules other tasks or an alarm expires.
-this is an optimization reducing the runtime overhead of green threads in the important single tasking case to 0.
+This is an optimization reducing the runtime overhead of green threads in the important single tasking case to 0.
 
@@
 The exception handler does not necessarily have to have been set up in the same runloop from which the exception has been thrown.
-Parrot uses the runloop id to identifiy such situations and free the nested runloops up to the level of the exception handler.
+Parrot uses the runloop id to identify such situations and free the nested runloops up to the level of the exception handler.
 
@@
 \textit{Parrot\_thread\_outer\_runloop} is used as the thread's main function.
-It's very similiar to \textit{Parrot\_cx\_outer\_runloop}.
+It's very similar to \textit{Parrot\_cx\_outer\_runloop}.
 The most important difference is the way it waits for pending alarms in case no tasks are to be run.
 While \textit{Parrot\_cx\_outer\_runloop} waits for an alarm signal to arrive, \textit{Parrot\_thread\_outer\_runloop} waits for a dummy to be written to its \textit{notifierfd} pipe by \textit{Parrot\_cx\_run\_scheduler}.
-In addition it is holding the \textit{interp->thread\_data->interp\_lock} lock while executing a task.
-This is the lock which \textit{Parrot\_cx\_schedule\_task} is taking during its call to \textit{Parrot\_thread\_schedule\_task} because the latter needs to create new objects local to the thread.
-As this could cause garbage collection to be triggered, the thread itself should not be executing code at the same time.
 
@@
 The default implementation for all writing functions would just call \textit{cant\_do\_write\_method} which creates a runtime exception.
-All other methods call the vtable method of the same name on the \textit{target} passing \textit{interp} as the current interpreter and all other parameters unchanged.
-If the method returns a PMC, another proxy object is created and wrapped around it so it can be safely returned to the caller.
+
+All other methods call the vtable method of the same name on the \textit{target} passing the current interpreter as \textit{interp} and all other parameters unchanged.
+This causes all access to globals by the proxied function to go through the thread's interp and new PMCs created during the call to be allocated from the thread's memory pool.
+If the method returns a PMC from the target's interp, another proxy object has to be created and wrapped around it so it can be safely returned to the caller.
+
+To differ between PMCs originating from the target's \textit{interp} and those created on the thread's interp during the call the GC is told to set the new \textit{PObj\_is\_new\_FLAG} on newly created PMCs.
+The \textit{PARROT\_THR\_FLAG\_NEW\_PMC} flag on the interp itself is used to communicate this requirement to the GC.
+
+The alternative of calling the proxied function using the target's \textit{interp} would lead to concurrency issues with the garbage collector.
+Any call may cause garbage collection to run.
+Since the GC also scans the C stack the target's GC would find the thread's PMCs there and mark them confusing the thread's GC.
+Disabling the garbage collector would require a lock since the target's GC may already have started collecting.
+
+\subsubsection{Sub}
+
+The \textit{Sub} PMC represents executable subroutines.
+A \textit{Sub} does not only contain the code to execute but also the context in which to execute the code such as visible globals and namespaces.
+If a proxy to such a \textit{Sub} were created and \textit{invoke} called on it, the code would access this context directly since it belongs to the same interp as the proxied \textit{Sub} itself.
+Thus an op like \textit{get\_global} would fetch a global from an unproxied namespace and an unproxied global would be put into the target register.
+Since this is happening while running \textit{invoke} on the original \textit{Sub}, \textit{Proxy} cannot intercept the call and create a \textit{Proxy} for the result.
+
+This is the reason why \textit{Parrot\_thread\_create\_proxy} does not create a \textit{Proxy} for a \textit{Sub} but uses \textit{Parrot\_thread\_create\_local\_sub} to create a copy on the thread's \textit{interp} with proxies for all PMC attributes like \textit{namespace\_stash} and \textit{ctx}.
 
@@
 
+\subsection{The \textit{wait} op}
+
+Using \textit{Tasks} to write to write to shared data makes such actions inherently asynchronous.
+This is not always what is needed by the implemented algorithm.
+For example when the shared variable is a lock, processing should continue as soon as it's acquired.
+The \textit{wait} op is used to wait for a task's completion.
+The waiting task is added to the waited for task's \textit{waiters} list and preempted immediately.
+When a task finishes, the all tasks in the \textit{waiters} list are scheduled again for execution.
+Since for each task a local copy is created on the target thread, the running task not only checks its own \textit{waiters} list but also its partner's.
+
+If a task on the main thread was waiting for a task on another thread to finish and no other tasks are in the scheduler's queue on the main thread, the main thread would exit if no alarms are pending.
+To prevent this unintended exit all tasks are added to the scheduler's \textit{foreign\_tasks} list when they are scheduled on other threads.
+To end the program with other threads still running an explicit \textit{exit} op has to be issued.
+
 \subsection{Garbage Collection (3)}
 
-Since each interpreter has its own garbage collector and consequenty its own memory areas and can only access its own data directly, the garbage collector can simply act as if there were no other threads.
+Since each interpreter has its own garbage collector and consequently its own memory areas and can only access its own data directly, the garbage collector can simply act as if there were no other threads.
 Thus the main work has been to clean up the interpreter structure itself and to replace references to global data by proxies.
@@
 Code segments and vtables are only marked by the main thread's garbage collector.
+Code segments are managed in Parrot using \textit{PackFile} structures.
+Besides the code itself these structures contain tables of constants used in the code.
+These constants are expanded to PMCs by the bytecode loader when the code is loaded from disk or put there directly by the PIR compiler.
+Since \textit{PackFiles} are no PMCs, the Proxy PMC cannot be used to shield them from the GC.
+Accessing the contained constants would therefore lead to unproxied PMCs being accessed by different threads.
+Since PackFiles may only be loaded by the main thread anyway, the bytecode loader and the compiler were changed to flag all \textit{PackFile} constants as \textit{shared}.
+They still are used as is on all threads, but the GC knows to only handle \textit{shared} PMCs on the main thread.
+
+Scheduling tasks on other interpreters may cause garbage collection due to the need of creating proxy objects.
+To protect the garbage collector from concurrency issues, garbage collection is disabled during creation of the new \textit{Task} object and the proxies on the target thread's \textit{interp}.
+But since the target thread may itself be allocating memory or collecting garbage, the GC has to be protected by a lock as well.
+The \textit{interp->thread\_data->interp\_lock} lock is used for this.
+It is checked by all allocation and freeing functions which are also the only places where a garbage collection run may be triggered.
 

### 2012-03-29 Paper.tex
Mention the first test program

@@
 
+In addition to Parrot's extensive test suite, tests were conducted using three test programs.
+
+\section{tasks.pir}
+
+tasks.pir is a simple test program, executing two \textit{Tasks} called ``a'' and ``b'' which run in tight loops printing their name on each 100000th iteration.
+The program ends after running for 10 seconds.
+This is a test of thread creation and task scheduling but does not access any shared data.
+The full code is as follows:
+%
+\begin{GenericCode}
+.sub main :main
+    .local pmc task, a, b, a_task, b_task
+    task = get_global 'task'
+    a = new ['String']
+    a = "a"
+    b = new ['String']
+    b = "b"
+    a_task = new ['Task']
+    setattribute a_task, 'code', task
+    setattribute a_task, 'data', a
+    b_task = new ['Task']
+    setattribute b_task, 'code', task
+    setattribute b_task, 'data', b
+    schedule a_task
+    schedule b_task
+    sleep 10
+    exit 0
+.end
+
+.sub task
+    .param pmc name
+    .local int i
+start:
+    print name
+    i = 0
+loop:
+    inc i
+    if i >= 100000 goto start
+    goto loop
+.end
+\end{GenericCode}
+
 \chapter{Conclusion, Further Work and Experiences (3)}

### 2012-04-17 Paper.tex
More about tests

@@
 \usepackage{wrapfig}
+\usepackage{listings}
 \AddBibFile{literature.bib}
@@
 
+\section{moretasks.pir}
+
+moretasks.pir is an extended test where the main task creates 50000 child tasks.
+These tasks would poll a shared variable with pauses of 100ms.
+When the variable is set to 1, they schedule tasks on the main thread to write their number to a results array.
+When the results array contains the expected number of results the process is repeated.
+This test stresses reading and writing to shared data, allocation and garbage collection.
+This test proved to be valuable for finding all sorts of concurrency issues.
+The complete source code is available in appendix \ref{app:moretasks}.
+
+\section{matrix\_part.winxed}
+
+This test implements multiplication of a matrix by a vector using four threads.
+It is written in the \textit{Winxed} programming language.
+Winxed is a low level language with JavaScript like syntax and the possibility to include sections of PIR code verbatim making it possible to try experimental opcodes while writing more readable and concise code than with PIR alone.
+The complete source code is available in appendix \ref{app:matrix_part}.
+
+\section{chameneos.pir}
+
+Chameneos is a game useful for testing peer-to-peer cooperation of threads presented in \cite{Chameneos}. The complete source code is available in appendix \ref{app:chameneos_part}.
+
 \chapter{Conclusion, Further Work and Experiences (3)}
@@
 
+There are several places with an implicit assumption that new threads may be only started from the main thread.
+Similarly only the main thread may schedule tasks on other threads.
+Sub threads may schedule tasks only on the main thread, not on other sub threads.
+These restrictions are not inherent in the implemented threading model but were introduced due to time constraints.
+
+\appendix
+\chapter{moretasks.pir}
+\label{app:moretasks}
+
+\begin{GenericCode}
+.sub main :main
+    .local pmc task, sayer, starter, number, interp, tasks, results
+    .local int i, num_results, results_rem
+    interp = getinterp
+    sayer = get_global 'sayer'
+init:
+    starter = new ['Integer']
+    i = 1
+    starter = 0
+    say "1..100"
+    tasks = new ['ResizablePMCArray']
+    results = new ['ResizablePMCArray']
+start:
+    number = new ['String']
+    number = i
+    task = new ['Task']
+    push task, results
+    push task, starter
+    setattribute task, 'code', sayer
+    setattribute task, 'data', number
+    print "ok "
+    say number
+    push tasks, task
+    schedule task
+    inc i
+    if i > 50000 goto run
+    goto start
+run:
+    starter = 1
+check_results:
+    pass
+    num_results = results
+    results_rem = num_results % 1000
+    if results_rem != 0 goto skip_say
+    say num_results
+skip_say:
+    if num_results >= 50000 goto end
+    goto check_results
+end:
+    goto init
+.end
+
+.sub sayer
+    .param pmc name
+    .local pmc interp, task, starter, results, result_sub, result_task
+    .local int i
+    interp = getinterp
+    task = interp.'current_task'()
+    starter    = pop task
+    results    = pop task
+    result_sub = get_global 'push_result'
+start:
+    if starter > 0 goto run
+    sleep 0.1
+    goto start
+run:
+    result_task = new ['Task']
+    setattribute result_task, 'code', result_sub
+    setattribute result_task, 'data', results
+    push result_task, name
+    interp.'schedule_proxied'(result_task, results)
+.end
+
+.sub push_result
+    .param pmc results
+    .local pmc interp, task, number
+    interp = getinterp
+    task = interp.'current_task'()
+    number = pop task
+    push results, number
+.end
+\end{GenericCode}
+
+\chapter{matrix\_part.winxed}
+\label{app:matrix_part}
+
+\lstinputlisting{/home/nine/install/parrot/examples/threads/matrix_part.winxed}
+
+\chapter{chameneos.pir}
+\label{app:chameneos_part}
+
+\lstinputlisting{/home/nine/install/parrot/examples/threads/chameneos.pir}
+
 \MakeBibliography{References, Literature (1)}

### 2012-05-12 Paper.tex

@@
 
-\section{Erlang}
-
 \chapter{Parrot (4)}
@@
 A stack based virtual machine usually fetches the operands of an operation from the top of a stack and puts the result back on top.
-Thus the operands chosen implicitly by ordering of the operations allowing the opcodes to be quite small.
+Thus the operands are chosen implicitly by ordering of the operations allowing the opcodes to be quite small.
 In a register based VM on the other hand each operation has to specify the operands explicitly.
@@
 \end{CCode}
-In short it contains a pointer to a type specific data structure and some meta data.
+In short they contain a pointer to a type specific data structure and some meta data.
 The vtable is where the type's behavior is defined.
@@
 
-Control flow is modeled in Parrot using continuation passing.
+Control flow in Parrot is modeled using continuation passing.
 A continuation is a data structure containing the state of a program at a given point in its execution.
@@
 
-A runloop is in principle the most inner loop executing bytecode in Parrot.
+A runloop is in principle the inner most loop executing bytecode in Parrot.
 It is also a data structure containing some information needed to support exception handling.
@@
 \textit{Parrot\_cx\_schedule\_sleep} is the actual implementation of the \textit{sleep} op.
-Like in\textit{Parrot\_cx\_preempt\_task}, \textit{Parrot\_cx\_stop\_task} is used to get an updated task at the current execution position, but instead of pushing this on the run queue, it is used as a callback for a newly set alarm.
+Like in \textit{Parrot\_cx\_preempt\_task}, \textit{Parrot\_cx\_stop\_task} is used to get an updated task at the current execution position, but instead of pushing this on the run queue, it is used as a callback for a newly set alarm.
 Again the NULL opcode is used to stop processing of the current task.

### 2012-06-05 Paper.tex
New section "Alarms and Timers"

@@
 
+\section{Alarms and Timers}
+
+Timers used for \textit{sleep}, alarms and preemption use a common timer thread implemented in \textit{src/alarm.c}.
+This thread sleeps until a new alarm is set or the currently active one expires.
+When the latter happens it increases the global alarm serial and notifies all threads that a timer is expired.
+Since it does not keep a list of pending alarms but knows only about the next one, interpreter threads are obliged to re-set their next alarms upon notification.
+
 \section{Threads (8)}

### 2012-06-05 Paper.tex
Add more details about the usage of locks in the GC

@@
 It is checked by all allocation and freeing functions which are also the only places where a garbage collection run may be triggered.
+Note that despite the GC being protected from concurrent access, this alone would still not be enough to share a GC between different threads since these threads could be changing references to objects while the GC does it's marking and sweeping.
+When only creating new proxies (which contain no references visible to the GC), accessing a foreign GC is safe on the other hand.
 

### 2012-06-05 Paper.tex
Start writing the last chapter

@@
 
-\begin{itemize}
-\item Windows implementation for timer/alarm handling
-\item Replace thread\_data->interp\_lock usage by preallocated proxy objects
-\item Allow threads to create threads
-\item Allow multi level proxying (don't create a proxy for a proxy, unpack first)
-\item Allow redistribution of tasks to other threads
-\end{itemize}
+This work shows a way to successfully implement threading support in a dynamic language runtime after the fact.
+Changes to the interpreter itself could be kept at a reasonable level and very local to the threading implementation itself.
+Benchmarks have shown that no slowdowns were measurable in the still very important single threaded case, while the implementation scales well using multiple threads.
 
-There are several places with an implicit assumption that new threads may be only started from the main thread.
+\section{Further work}
+
+\subsection{Remove locking from the GC}
+
+The initial goal of an almost lock free implementation has not been met fully.
+The garbage collector has to take a lock in most of it's functions.
+Though these locks are used only when more than one interpreter thread is running, they still represent an overhead which could be avoided.
+A way to accomplish this would be to pre-allocate proxy objects instead of allocating them on demand.
+The allocation could be done as part of a thread's outer runloop.
+This would eliminate the need for the locks in the garbage collector.
+The number of proxy objects to pre-allocate would be an important knob for optimization, since too few objects would lead to starvation quickly when many tasks are scheduled.
+
+\subsection{Allow threads to create threads}
+
+In the current implementation there are several places with an implicit assumption that new threads may be only started from the main thread.
 Similarly only the main thread may schedule tasks on other threads.
@@
 These restrictions are not inherent in the implemented threading model but were introduced due to time constraints.
+
+\subsection{Allow multi level proxying}
+
+Communication between sub threads is currently restricted by the fact that proxy objects can only be transfered back to the thread which created them.
+To remove this restriction, \textit{Parrot\_thread\_create\_proxy} has to be extended to detect an attempted transfer of a proxy to a different thread and unwrap it first before creating a new proxy.
+
+\subsection{Allow redistribution of tasks to other threads}
+
+\subsection{Dynamic maximum number of threads}
 

### 2012-06-05 Praktikum.tex
Anfang des Praktikumsberichts

%%% Einfaches Template für den Praktikumsbericht (2. Bachelorarbeit)

\documentclass[praktikum,german]{hgbthesis}
\graphicspath{{images/}}   				% wo liegen die EPS-Bilder?
%\AddBibFile{literatur.bib}  % Angabe der BibTeX-Datei

%%%----------------------------------------------------------
\begin{document}
%%%----------------------------------------------------------

\author{Stefan Seifert}
\studiengang{Software Engineering}
\studienort{Hagenberg}
\abgabemonat{Februar}
\abgabejahr{2012}

\nummer{1010307037-B}
\betreuer{Mag. Thomas Seel}
\firma{%
   Atikon EDV \& Marketing GmbH\\
   4020 Leonding, Kornstraße 4
}
\firmenTel{+43 732 611266}
\firmenUrl{www.atikon.com}

%%%----------------------------------------------------------
\frontmatter
\maketitle
\tableofcontents
%%%----------------------------------------------------------

\chapter{Kurzfassung}
Umfang der Kurzfassung: ca.\ 200 Worte.

Zum allgemeinen Inhalt des Berichts: Dieser Bericht beschreibt den Ablauf des Praktikums, die Aufgaben und durchgeführten Projekte und Erfahrungen. Die eigenen Aktivitäten (Projekte) stehen dabei natürlich im Mittelpunkt und bilden den Hauptteil des Berichts. Wenn viele Kleinprojekte bearbeitet wurden, sollten einige davon exemplarisch
genauer beschrieben werden. Neben der eigentlichen Arbeit sollten aber auch folgende weitere Aspekte berücksichtigt werden:
%
\begin{itemize}
\item Abläufe (Workflows) innerhalb des Unternehmens bzw. in Projekten (grafische Darstellungen
können dabei nützlich sein)
\item Arbeits- und Führungsstil, Kommunikation innerhalb des Unternehmens
\item Kommunikation nach außen (Kunden, Partner)
\item Zeitsituation, Terminprobleme
\item Einbettung in das Team, soziale Erfahrungen
\item Einsatz von speziellen Techniken, Methoden und Werkzeugen.
\item Wichtige Herausforderungen oder Schwierigkeiten
\item Anforderungen in Bezug auf die Ausbildung im Studium (gut einsetzbare Kenntnisse, Defizite)
\end{itemize}
%
Die nachfolgenden Kapitelüberschriften sollen nur zur Orientierung für die Struktur des Berichts dienen, über die konkrete Einteilung und den Wortlaut kann man natürlich selbst entscheiden.

%%%----------------------------------------------------------
\mainmatter           %Hauptteil (ab hier arab. Seitenzahlen)
%%%----------------------------------------------------------

\chapter{Das Unternehmen}
Atikon EDV \& Marketing GmbH mit Sitz in Leonding bietet in den 11 Jahren ihres Bestehens mit 55 Mitarbeitern Marketingdienste für Steuerberater und Ärzte und entwickelt Beratungssoftware für Steuerberater. Diese Spezialisierung spiegelt sich in der Kundenstruktur wieder. Unter den ca. 3400 Kunden befinden sich in etwa 80 \% Steuerberater, 10 \% Ärzte und 10 \% sonstige Unternehmer.

Diesen Kunden werden im Marketingbereich Websites mit individuellem Design und verschiedenste Drucksorten angeboten. Dazwischen gibt es noch Hybridprodukte wie Steuernews und Steuerinfo, welche in die Webseite eingebunden oder gedruckt verteilt werden können. Abgerundet wird das Portfolio durch Onlinetools wie Steuerrechner, z.B. zur Brutto/Netto Berechnung oder zur Einkommensteuerberechnung und einem Newslettersystem.

Die Beratungssoftware wird vom Team der Windowsprogrammierung erstellt. Hier gibt es eigenvertriebene Produkte wie die "Formel-f" oder Software zur Bilanzpräsentation und Auftragsarbeiten, z.B. für die deutsche DATEV.

Das fünfköpfige Team der Webprogrammierung kümmert sich um Webanwendungen sowohl für den internen Gebrauch, als auch für externe Kunden. Dazu gehören:
%
\begin{itemize}
\item das Content Management System
\item Projektverwaltung
\item Onlinetools (z.B. die Steuerrechner)
\item Newslettersystem
\item Kundenprojekte wie z.B. ein Online-CRM-System
\end{itemize}

\chapter{Projekte und Tätigkeiten während des Praktikums}

\section{CMS3000}
Umfang: 2--3 Seiten (Projektziel(e), Projektumfeld)

Websites sind eines der Hauptprodukte der Firma. Die Kundenstruktur findet sich auch abgebildet auf die ca. 1600 bisher erstellten Websites wieder. Steuerberater sind zumeist eher kleine Unternehmen. Es gibt jedoch unter ihnen auch Netzwerke aus zahlreichen Einzelunternehmen mit zusammen 100en Standorten und 1000en Mitarbeitern. Dementsprechend sind auch bis auf einzelne Ausnahmen die meisten der von uns erstellten und betreuten Websites relativ klein. Die wichtigsten Verkaufsargumente sind die einerseits verfügbaren branchenspezifischen Inhalte wie z.B. regelmäßig aktualisierte Steuernews, andererseits aber die vollständige Individualisierung von Layouts und Websitestruktur.

Dementsprechend gestalten sich daher auch die Anforderungen an das Content Management System:
%
\begin{itemize}
\item Unterstützung für 1000e individuelle Websites in einem System
\item Einfache Verteilung und Updates von gemeinsam genutzten Inhalten
\item Völlige Freiheit bei Design und Struktur
\item Individuelle Erweiterungen
\item Einfache Bedienung für Kunden mit Eigenwartung
\end{itemize}

In den ersten 10 Jahren des Firmenbestehens hat das deutsche Produkt "ZMS" die Anforderungen am besten erfüllt. Es handelt sich dabei um freie, GPL lizenzierte Software, die in Python geschrieben ist und auf dem Zope Application Server basiert.

\chapter{Projektbeispiele}
Umfang: 5--6 Seiten (Umsetzung, grober Terminplan, Ergebnisse, Qualitätssicherungsmaßnahmen)

\chapter{Erfahrungen und Zusammenfassung}
Umfang: 1--2 Seiten

Zum allgemeinen Inhalt des Berichts: Dieser Bericht beschreibt den Ablauf des Praktikums, die Aufgaben und durchgeführten Projekte und Erfahrungen. Die eigenen Aktivitäten (Projekte) stehen dabei natürlich im Mittelpunkt und bilden den Hauptteil des Berichts. Wenn viele Kleinprojekte bearbeitet wurden, sollten einige davon exemplarisch und im Detail beschrieben werden.

%%%----------------------------------------------------------
% Quellenverzeichnis (sofern notwendig, sonst weglassen)
%\MakeBibliography{Quellenverzeichnis}

%%%Messbox zur Druckkontrolle
%\input{messbox.tex}

\end{document}

% vim: spell spelllang=de_at

### 2012-06-05 Paper.tex
Fix little style problem
Abstracts

@@
 
+Parrot is a runtime for dynamically typed programming languages. Despite several attempts in it's 10 years of history, it does provide any support for multithreaded computation.
+
+This paper shows a way to implement threading support in the runtime using a hybrid between lightweight (``green'') threads and operating system threads.
+These lightweight threads are used as messages in a system where reading shared data is allowed but only the one owner thread may write to it.
+The implementation is described in detail and benchmarks presented demonstrating the viability of this design.
+
 \begin{german}
 \chapter{Kurzfassung (1)}
+
+Parrot ist eine Laufzeitumgebung f{\"u}r dynamisch typisierte Programmiersprachen.
+Trotz mehrerer Anl{\"a}ufe in seiner zehnj{\"a}hrigen Geschichte, bietet es keine Unterst{\"u}tzung f{\"u}r Multithreading.
+
+Diese Arbeit zeigt einen Weg um Threadingunterst{\"u}tzung in der Laufzeitumgebung zu implementieren mit Hilfe eines hybriden Systems aus leichtgewichtigen und Betriebssystem- Threads.
+Diese leichtgewichtigen Threads werden als Nachrichten in einem System benutzt, in dem das Lesen von gemeinsamen Daten erlaubt ist, aber nur der eine besitzende Thread darauf schreiben darf.
+Die Implementierung wird detailiert beschrieben und mit Benchmarks die Brauchbarkeit des Designs gezeigt.
+
 \end{german}
@@
 When the results array contains the expected number of results the process is repeated.
+
 This test stresses reading and writing to shared data, allocation and garbage collection.
-This test proved to be valuable for finding all sorts of concurrency issues.
+It proved to be valuable for finding all sorts of concurrency issues.
 The complete source code is available in appendix \ref{app:moretasks}.

### 2012-06-05 Paper.tex
s/it's/its/

@@
 It is checked by all allocation and freeing functions which are also the only places where a garbage collection run may be triggered.
-Note that despite the GC being protected from concurrent access, this alone would still not be enough to share a GC between different threads since these threads could be changing references to objects while the GC does it's marking and sweeping.
+Note that despite the GC being protected from concurrent access, this alone would still not be enough to share a GC between different threads since these threads could be changing references to objects while the GC does its marking and sweeping.
 When only creating new proxies (which contain no references visible to the GC), accessing a foreign GC is safe on the other hand.

### 2012-06-05 Paper.tex
Benchmark results

@@
 
+The program consists of the parts initialization, computation and verification.
+Computation is parallelized using 4 tasks each calculating one fourth of the result vector.
+Runtime is compared to a simple single threaded implementation.
+Run times were measured using the \textit{time} command and are recorded in table \ref{benchmark:matrix}
+
+\begin{table}
+\caption{Runtime comparison for matrix multiplication}
+\label{benchmark:matrix}
+\begin{tabular}{|l||c|c||c|c|} \hline
+& \emph{singlethreaded} & \emph{computation} & \emph{multithreaded} & \emph{computation} \\
+\hline
+1. run  & 28.522s & 19.530s & 17.543s & 8.478s \\
+\hline
+2. run  & 28.427s & 19.463s & 17.320s & 8.283s \\
+\hline
+3. run  & 28.200s & 19.235s & 17.489s & 8.473s \\
+\hline
+average & 28.383s & 19.409s & 17.451s & 8.411s \\
+\hline
+\end{tabular}
+\end{table}
+
+As can be seen the multi threaded implementation gives an average speedup of 2.31 for the computation and 1.61 in total.
+
 \section{chameneos.pir}
 
-Chameneos is a game useful for testing peer-to-peer cooperation of threads presented in \cite{Chameneos}. The complete source code is available in appendix \ref{app:chameneos_part}.
+Chameneos is a game useful for testing peer-to-peer cooperation of threads presented in \cite{Chameneos}.
+It revolves around fictional creatures, each being of one of three colors.
+Two of these creatures may meet and exchange their colors with each one independently calculating and adopting a complementary color depending on those two.
+These creatures have to be implemented using preemptive kernel or lightweight threads.
+Since the color calculation is trivial this test's performance depends in full on the available thread synchronization primitives and writes to shared data.
+It therefore represents a worst case scenario for the implemented threading model.
+The complete source code is available in appendix \ref{app:chameneos_part}.
+
+Preliminary benchmarks have shown Parrot's performance to be within an order of magnitude with that of an optimized implementation in Perl 5.
+Since Parrot does not yet offer the user any synchronization primitives, locks had to be implemented using a shared variable which is written to only by the main thread.
+Replacing this very primitive method with a native semaphore implementation would probably reduce runtime to a small fraction.
+
+\section{mandel\_inter.winxed}
+
+Calculating an image of the Mandelbrot set is a common benchmark for multithreading implementations since calculations of points are independent of each other and are thus easily parallelizable.
+A simple implementation of the escape time algorithm written in Winxed has been used to determine scalability properties of the threading implementation.
+The image is split into lines which are calculated alternatedly by a configured number of tasks.
+Run times were measured using the \textit{time} command on an Intel Core i7 3770K processor with 16 GiB of RAM running openSUSE 12.1 and are recorded in table \ref{benchmark:mandel}
+
+\begin{table}
+\caption{Runtime comparison for Mandelbrot set calculation}
+\label{benchmark:mandel}
+\begin{tabular}{|l||c||c|c|c|c|} \hline
+& \emph{singlethreaded} & \emph{1 thread} & \emph{2 threads} & \emph{4 threads} & \emph{8 threads} \\
+\hline
+1. run  & 89.931s & 89.978s & 45.813s & 24.028s & 17.445s \\
+\hline
+2. run  & 89.707s & 89.871s & 45.906s & 24.048s & 17.695s \\
+\hline
+3. run  & 90.318s & 89.839s & 45.951s & 24.049s & 17.573s \\
+\hline
+average & 89.985s & 89.896s & 45.890s & 24.042s & 17.571s  \\
+\hline
+speedup & 1       &  1      &  1.959  &  3.739  &  5.116  \\
+\hline
+\end{tabular}
+\end{table}
+
+As can be seen in the results, the implementation scales nearly linearly up to 4 threads reflecting the CPUs 4 physical cores.
+Using 8 threads the speedup is only 1.368 compared to 4 threads but this seems to be more a limitation of the hardware than the implementation.
 

### 2012-06-05 Paper.tex
Finish of the finish

@@
 
+Practice will show how this threading model will fare with real world workloads. It very much depends on the ratio of read to write accesses to shared data.
+
 \section{Further work}
@@
 The initial goal of an almost lock free implementation has not been met fully.
-The garbage collector has to take a lock in most of it's functions.
+The garbage collector has to take a lock in most of its functions.
 Though these locks are used only when more than one interpreter thread is running, they still represent an overhead which could be avoided.
@@
 
+\subsection{Dynamic maximum number of threads}
+
+The maximum number of threads is determined by the MAX\_THREADS constant in \textit{include/parrot/thread.h}.
+Instead the value should be determined at runtime probably depending on the available hardware resources, demand and system load.
+
 \subsection{Allow multi level proxying}
@@
 
-\subsection{Dynamic maximum number of threads}
+Currently the scheduler picks the interpreter with the smallest number of tasks in the task queue for scheduling a new task.
+This can lead to the situation where threads sit idle while tasks have to wait in other thread's queues.
+Some tasks get scheduled on specific threads because they need write access to the thread's PMCs.
+To allow migration of waiting tasks to idle threads there needs to be a method to distinguish such tasks from tasks which are there simply because the thread's queue was the shortest at the time the task got scheduled.
+The more difficult part of implementing task migration is that proxies are allocated on the target interpreter at the time of scheduling.
+Since these proxies belong to the interpreter they are created on, new proxies would have to be created on the interpreter the task is being migrated to.
+It is yet unclear if this overhead can be offset by better utilization of idle threads.
+
+\section{Experiences}
+
+I picked a very advanced topic for my Bachelor's thesis.
+While I had to invest many more hours than strictly necessary to get a Bachelor's degree, the work was also very interesting, instructive and hopefully valuable to the Parrot project and the free software community.
+Considering that I had no prior experience with Parrot which is a highly complex piece of software, it is not surprising that I spent most of the time discovering how it works and how its many features interact.
+In addition, concurrency usually brings with it difficult to diagnose problems further complicated by Parrot's cross platform nature.
+
+It is therefore no exaggeration to state that without the help of the Parrot community I could not have come nearly as far as I did in the time I had.
+They are very welcoming and were supportive and grateful from the first day, boosting my motivation.
+I can only recommend to other students to approach the free software communities when looking for thesis topics.
 

### 2012-06-06 Praktikum.tex
Kurzfassung Praktikumsbericht

@@
-%%% Einfaches Template für den Praktikumsbericht (2. Bachelorarbeit)
-
 \documentclass[praktikum,german]{hgbthesis}
@@
 \chapter{Kurzfassung}
-Umfang der Kurzfassung: ca.\ 200 Worte.
 
-Zum allgemeinen Inhalt des Berichts: Dieser Bericht beschreibt den Ablauf des Praktikums, die Aufgaben und durchgeführten Projekte und Erfahrungen.
-Die eigenen Aktivitäten (Projekte) stehen dabei natürlich im Mittelpunkt und bilden den Hauptteil des Berichts.
-Wenn viele Kleinprojekte bearbeitet wurden, sollten einige davon exemplarisch
-genauer beschrieben werden. Neben der eigentlichen Arbeit sollten aber auch folgende weitere Aspekte berücksichtigt werden:
+Atikon EDV \& Marketing GmbH entwickelt Webseiten für hauptsächlich Steuerberater und Ärzte, sowie Software für Steuerberater.
+Als Leiter der Webprogrammierung führe ich ein Team von 5 Programmierern.
+Wir entwickeln Software für den internen Gebrauch sowie für externe Kunden, z.B.
+das Content Management System, die Projektverwaltung, Online-Steuerberechnung und Individualentwicklung wie z.B.
+ein CRM System für einen unserer Kunden.
+
+Mit inzwischen über 1100 aktiven Webseiten ergeben sich in der Programmierung interessante Aufgaben von der Skalierung des Systems über die User Interface Gestaltung bis zur Automatisierung von Routineaufgaben um die Wartung und Pflege der Webseiten so effizient wie möglich zu gestalten.
+
+Bei der Programmierung des Content Management Systems galt es ein bestehendes System abzulösen, das mit fast 1000 Webseiten an seine Grenzen gestoßen war.
+Ziel war dabei eine vollständige Übernahme von bestehenden Webseiten in das neue System, da diese weiterhin gewartet werden müssen und automatisiert Inhalte wie Steuernews darauf verteilt werden.
+Während das alte CMS auf Python basiert und eine objektorientierte Datenbank (ZODB) benutzt, ist das neue in Perl geschrieben und benutzt PostgreSQL als Datenbank.
+
+Meine Aufgaben waren dabei neben der Projektleitung:
 %
 \begin{itemize}
-\item Abläufe (Workflows) innerhalb des Unternehmens bzw. in Projekten (grafische Darstellungen
-können dabei nützlich sein)
-\item Arbeits- und Führungsstil, Kommunikation innerhalb des Unternehmens
-\item Kommunikation nach außen (Kunden, Partner)
-\item Zeitsituation, Terminprobleme
-\item Einbettung in das Team, soziale Erfahrungen
-\item Einsatz von speziellen Techniken, Methoden und Werkzeugen.
-\item Wichtige Herausforderungen oder Schwierigkeiten
-\item Anforderungen in Bezug auf die Ausbildung im Studium (gut einsetzbare Kenntnisse, Defizite)
+\item{Entwurf der Architektur des neuen Systems}
+\item{Implementierung der Grundstruktur}
+\item{Implementierung des Exports aus dem alten CMS}
+\item{Implementierung der Unterstützung für die Templatesprache des alten CMS}
+\item{Erweiterung des Inline::Python Perlmoduls um die von uns benötigten Features}
+\item{Implementierung von automatischen Kompatibilitätstests mittels Screenshotvergleich der Webseiten mit allen Unterseiten.}
 \end{itemize}
-%
-Die nachfolgenden Kapitelüberschriften sollen nur zur Orientierung für die Struktur des Berichts dienen, über die konkrete Einteilung und den Wortlaut kann man natürlich selbst entscheiden.
 

### 2012-06-06 Praktikum.tex
Praktikum: Webprogrammierung, Zope/ZMS

@@
 
-\chapter{Projekte und Tätigkeiten während des Praktikums}
+Die Software ist dabei größtenteils in Perl geschrieben. Als Datenbank kommt hauptsächlich PostgreSQL zum Einsatz. Alte Anwendungen sind teilweise in PHP oder Python geschrieben und setzen auf MySQL.
 
-\section{CMS3000}
-Umfang: 2--3 Seiten (Projektziel(e), Projektumfeld)
+\chapter{Ausgangslage}
+
+\section{Das abzulösende System}
 
@@
 
-In den ersten 10 Jahren des Firmenbestehens hat das deutsche Produkt "ZMS" die Anforderungen am besten erfüllt.
+In den ersten 10 Jahren des Firmenbestehens hat das deutsche Produkt ``ZMS'' die Anforderungen am besten erfüllt.
 Es handelt sich dabei um freie, GPL lizenzierte Software, die in Python geschrieben ist und auf dem Zope Application Server basiert.
+Zope bringt eine eigene, objektorientierte Datenbank mit, genannt ``ZODB''.
+Diese Datenbank ist im Prinzip ein serialisierter Graph von Pythonobjekten, der zusammen mit den Indizes in einer Datei gespeichert wird.
+Der Application Server bringt eine eigene Weboberfläche mit, in der diese Datenbank als hierarchische Ordner- und Dateistruktur dargestellt wird.
 
-\chapter{Projektbeispiele}
-Umfang: 5--6 Seiten (Umsetzung, grober Terminplan, Ergebnisse, Qualitätssicherungsmaßnahmen)
+ZMS führt diese Grundstruktur fort, indem es die Inhalte von Webseiten ebenfalls in einer Ordner- und Dokumentenstruktur präsentiert.
+Das Layout wird durch Templates definiert, die in der Zope-eigenen Templatesprache ``DTML'' geschrieben sind.
+Diese Sprache bietet in einer HTML-ähnlichen Syntax die üblichen Konstrukte wie Verzweigungen, Schleifen, Definition und Ausgabe von Variablen.
+Diese Konstrukte unterstützen dabei die Formulierung von komplexen Ausdrücken in Python:
+%
+\begin{GenericCode}
+    <dtml-var "getObjPropert('foo', REQUEST).upper()" html_quote>
+\end{GenericCode}
+
+In diesem Beispiel wird die \textit{getObjProperty} Methode des Kontextobjekts aufgerufen mit dem String ``foo'' und dem immer verfügbaren REQUEST-Objekt als Parameter.
+Als Ergebnis wird ein String erwartet, dessen upper() Methode den String in Großbuchstaben zurückliefert.
+Das Ergebnis des Ausdrucks wird ausgegeben, wobei Zeichen, die in HTML eine spezielle Bedeutung haben escaped werden.
+
+Das für uns wichtigste Feature des ZMS war die einfache Erweiterbarkeit der zur Verfügung stehenden Inhaltsobjekte.
+ZMS erlaubt die Definition von eigenen sogenannten ``speziellen Objekten'' mit beliebigen Datenfeldern und eigenem Ausgabetemplate.
+Das Eingabeinterface wird dabei automatisch generiert.
+Damit ist es z.B.
+sehr einfach möglich für eine Mitarbeiterseite ein Mitarbeiter-Objekt zu definieren wo der entsprechende Mitarbeiter mit Namen, Emailadresse und Durchwahl zusammen mit einem Bild ausgegeben wird.
+Dieser Mechanismus erlaubt es, unterschiedlichsten Kundenanforderungen gerecht zu werden.
+
+\section{Probleme}
+
+Die Probleme, die zum Entschluss einer Neuprogrammierung geführt haben, waren dabei folgende:
+
+\subsection{Zuverlässigkeit}
+
+Der Zope Server kann nur starten, wenn er die Zope-Datenbank lesen kann.
+Gleichzeitig ist die Weboberfläche des Zope-Servers die einzige Möglichkeit um die Zope-Datenbank manuell zu bearbeiten.
+Ist nun die Datenbank durch Programmfehler beschädigt, gibt es kaum Mittel um sie zu reparieren.
+Derartige Probleme haben die Firma schon komplette Arbeitstage gekostet und dazu geführt, dass am Schluss ein stündliches Backup der Datenbank erstellt wurde.
+
+\subsection{Performance}
+
+Operationen, die alle Websites betreffen wie zum Beispiel das Update von gemeinsamen Inhalten, das bloße Auflisten aller Websites oder ein Durchsuchen aller Websites zeigten ein höchst unberechenbares und exponentielles Laufzeitverhalten.
+Das Auflisten der Websites konnte von wenigen Sekunden bis zu mehreren Minuten dauern.
+Das Inhaltsupdate lief mehrere Stunden während derer der Zugriff extrem verlangsamt wurde, weshalb es nur über Nacht durchgeführt werden konnte.
+
+In Zope wird jeder Webrequest automatisch in einer Transaktion ausgeführt.
+Treten Zugriffskonflikte mit anderen Transaktionen auf, wird die Transaktion automatisch zurückgesetzt und der Request wiederholt.
+Dies passiert vollautomatisch bis zu drei Mal bevor Zope den Request mit einer Fehlermeldung beendet.
+Bei Transaktionen, die mehrere Stunden laufen ist die Chance auf ein erfolgreiches Durchkommen sehr klein und die Zeit bis zur erfolgreichen Erledigung sehr lang.
+Als Abhilfe wurden derartige Zugriffe möglichst unterteilt (z.B.
+das Inhaltsupdate nur für einzelne Websites) und von außen gesteuert durchgeführt.
+Dies hat die Komplexität und damit die Fehleranfälligkeit wieder erhöht.
+
+Die Speicherung als Objektgraph macht bestimmte Operationen, die in SQL Datenbanken sehr schnell gehen äußerst langsam, z.B.
+das Durchsuchen der gesamten Website nach Objekten eines bestimmten Typs um zum Beispiel die neuesten News aller Kategorien gesammelt auf einer Startseite anzuzeigen.
+
+\subsection{Wartbarkeit}
+
+DTML verleitet dazu Präsentation und Logik zu mischen und beides in einem Template einzubauen.
+Gleichzeitig macht es durch fortgeschrittene Features das Lesen von Code schwierig und unzuverlässig.
+Es gibt in DTML nur einen einzigen, großen Namensraum.
+In diesem sind die Methoden und Eigenschaften des impliziten Kontextobjekts genauso enthalten, wie eingebaute Funktionen der Sprache und die aller höher liegenden Objekte in der Hierarchie.
+Da Templates und Scripte als Datei-Objekte üblicherweise im Wurzelverzeichnis einer Website abgelegt werden, sind diese als Eigenschaften des Wurzelverzeichnises überall ohne weitere Qualifikation verfügbar.
+Das trifft jedoch auch auf Objekte zu, die weiter oben in der Hierarchie liegen.
+Zum Beispiel hat das Anlegen eines Ordners namens ``test'' in der obersten Hierarchieebene dazu geführt, dass bei manchen Websites Fehler ausgegeben wurden, weil in manchen Ausdrücken dieser Ordner statt der globalen Funktion gleichen Namens gefunden wurde und ein Ordner kein ausführbares Objekt ist.
+
+\chapter{CMS 3000}
+
+Die Entscheidung für eine Eigenentwicklung wurde aus zwei Gründen gefällt.
+Einerseits wurde bei einer Marktanalyse kein System mit einem ähnlich gut auf die Kundenstruktur passendem Featureprofil gefunden.
+Andererseits bestand der Wunsch für die bestehenden 1000 Websites einen Upgradepfad zu ermöglichen um den Wartungsaufwand zu minimieren.
 
 \chapter{Erfahrungen und Zusammenfassung}
-Umfang: 1--2 Seiten
-
-Zum allgemeinen Inhalt des Berichts: Dieser Bericht beschreibt den Ablauf des Praktikums, die Aufgaben und durchgeführten Projekte und Erfahrungen.
-Die eigenen Aktivitäten (Projekte) stehen dabei natürlich im Mittelpunkt und bilden den Hauptteil des Berichts.
-Wenn viele Kleinprojekte bearbeitet wurden, sollten einige davon exemplarisch und im Detail beschrieben werden.
 

### 2012-06-11 Praktikum.tex
CMS 3000 Grundstruktur

@@
 Ziel war dabei eine vollständige Übernahme von bestehenden Webseiten in das neue System, da diese weiterhin gewartet werden müssen und automatisiert Inhalte wie Steuernews darauf verteilt werden.
-Während das alte CMS auf Python basiert und eine objektorientierte Datenbank (ZODB) benutzt, ist das neue in Perl geschrieben und benutzt PostgreSQL als Datenbank.
+Während das alte CMS auf Python basiert und eine objektorientierte Datenbank (ZODB) benutzt, ist das neue in Perl geschrieben und benutzt \textit{PostgreSQL} als Datenbank.
 
@@
 
-Die Software ist dabei größtenteils in Perl geschrieben. Als Datenbank kommt hauptsächlich PostgreSQL zum Einsatz. Alte Anwendungen sind teilweise in PHP oder Python geschrieben und setzen auf MySQL.
+Die Software ist dabei größtenteils in Perl geschrieben.
+Als Datenbank kommt hauptsächlich \textit{PostgreSQL} zum Einsatz.
+Alte Anwendungen sind teilweise in PHP oder Python geschrieben und setzen auf MySQL.
 
@@
 
-In den ersten 10 Jahren des Firmenbestehens hat das deutsche Produkt ``ZMS'' die Anforderungen am besten erfüllt.
-Es handelt sich dabei um freie, GPL lizenzierte Software, die in Python geschrieben ist und auf dem Zope Application Server basiert.
-Zope bringt eine eigene, objektorientierte Datenbank mit, genannt ``ZODB''.
+In den ersten 10 Jahren des Firmenbestehens hat das deutsche Produkt \textit{ZMS} die Anforderungen am besten erfüllt.
+Es handelt sich dabei um freie, GPL lizenzierte Software, die in Python geschrieben ist und auf dem \text{Zope} Application Server basiert.
+\textit{Zope} bringt eine eigene, objektorientierte Datenbank mit, genannt ``ZODB''.
 Diese Datenbank ist im Prinzip ein serialisierter Graph von Pythonobjekten, der zusammen mit den Indizes in einer Datei gespeichert wird.
@@
 
-ZMS führt diese Grundstruktur fort, indem es die Inhalte von Webseiten ebenfalls in einer Ordner- und Dokumentenstruktur präsentiert.
-Das Layout wird durch Templates definiert, die in der Zope-eigenen Templatesprache ``DTML'' geschrieben sind.
+\textit{ZMS} führt diese Grundstruktur fort, indem es die Inhalte von Webseiten ebenfalls in einer Ordner- und Dokumentenstruktur präsentiert.
+Das Layout wird durch Templates definiert, die in der \textit{Zope}-eigenen Templatesprache \textit{DTML} geschrieben sind.
 Diese Sprache bietet in einer HTML-ähnlichen Syntax die üblichen Konstrukte wie Verzweigungen, Schleifen, Definition und Ausgabe von Variablen.
@@
 \begin{GenericCode}
-    <dtml-var "getObjPropert('foo', REQUEST).upper()" html_quote>
+    <dtml-var "getObjProperty('foo', REQUEST).upper()" html_quote>
 \end{GenericCode}
@@
 
-Das für uns wichtigste Feature des ZMS war die einfache Erweiterbarkeit der zur Verfügung stehenden Inhaltsobjekte.
-ZMS erlaubt die Definition von eigenen sogenannten ``speziellen Objekten'' mit beliebigen Datenfeldern und eigenem Ausgabetemplate.
+Das für uns wichtigste Feature des \textit{ZMS} war die einfache Erweiterbarkeit der zur Verfügung stehenden Inhaltsobjekte.
+\textit{ZMS} erlaubt die Definition von eigenen sogenannten ``speziellen Objekten'' mit beliebigen Datenfeldern und eigenem Ausgabetemplate.
 Das Eingabeinterface wird dabei automatisch generiert.
@@
 
-Der Zope Server kann nur starten, wenn er die Zope-Datenbank lesen kann.
-Gleichzeitig ist die Weboberfläche des Zope-Servers die einzige Möglichkeit um die Zope-Datenbank manuell zu bearbeiten.
+Der \textit{Zope}-Server kann nur starten, wenn er die \textit{Zope}-Datenbank lesen kann.
+Gleichzeitig ist die Weboberfläche des \textit{Zope}-Servers die einzige Möglichkeit um die \textit{Zope}-Datenbank manuell zu bearbeiten.
 Ist nun die Datenbank durch Programmfehler beschädigt, gibt es kaum Mittel um sie zu reparieren.
@@
 
-In Zope wird jeder Webrequest automatisch in einer Transaktion ausgeführt.
+In \textit{Zope} wird jeder Webrequest automatisch in einer Transaktion ausgeführt.
 Treten Zugriffskonflikte mit anderen Transaktionen auf, wird die Transaktion automatisch zurückgesetzt und der Request wiederholt.
-Dies passiert vollautomatisch bis zu drei Mal bevor Zope den Request mit einer Fehlermeldung beendet.
+Dies passiert vollautomatisch bis zu drei Mal bevor \textit{Zope} den Request mit einer Fehlermeldung beendet.
 Bei Transaktionen, die mehrere Stunden laufen ist die Chance auf ein erfolgreiches Durchkommen sehr klein und die Zeit bis zur erfolgreichen Erledigung sehr lang.
@@
 
-DTML verleitet dazu Präsentation und Logik zu mischen und beides in einem Template einzubauen.
+\textit{DTML} verleitet dazu Präsentation und Logik zu mischen und beides in einem Template einzubauen.
 Gleichzeitig macht es durch fortgeschrittene Features das Lesen von Code schwierig und unzuverlässig.
-Es gibt in DTML nur einen einzigen, großen Namensraum.
+Es gibt in \textit{DTML} nur einen einzigen, großen Namensraum.
 In diesem sind die Methoden und Eigenschaften des impliziten Kontextobjekts genauso enthalten, wie eingebaute Funktionen der Sprache und die aller höher liegenden Objekte in der Hierarchie.
@@
 Andererseits bestand der Wunsch für die bestehenden 1000 Websites einen Upgradepfad zu ermöglichen um den Wartungsaufwand zu minimieren.
+
+\section{Grundstruktur}
+
+Als Datenspeicher wurde \textit{PostgreSQL} gewählt, eine freie und zugleich mächtige und performante SQL Datenbank.
+Um die einfache Erweiterbarkeit des \textit{ZMS}, wo für jede Seite neue Objekttypen definiert weden können und sogar bestehende um neue Felder erweitert, zu erhalten, bekommt jede Website ein komplett eigenes Datenbankschema.
+Diese Schemata werden in der Datenbank als getrennte Namensräume (im \textit{PostgreSQL} Jargon ebenfalls ``schema'') ausgeführt.
+Da dadurch die Anzahl der Datenbanktabellen mit jeder Website stark steigt, wurden Voruntersuchungen durchgeführt, ob \textit{PostgreSQL} mit mehreren 100.000 Tabellen umgehen kann.
+Es wurden dabei keine inhärenten Einschränkungen gefunden.
+Die Skalierung hängt rein vom darunterliegenden Dateisystem ab.
+Das verwendete \textit{ext4} scheint gut zu skalieren.
+
+Die Wahl der Programmiersprache fiel auf Perl, da in der Firma viel Know How für diese Sprache vorhanden ist.
+Mit \textit{Catalyst} steht ein flexibles MVC Framework für Webanwendungen zur Verfügung.
+Im \textit{Perl Comprehensive Archive Network} (\textit{CPAN}) sind unzählige Libraries verfügbar, von denen ca.
+80 dann auch verwendet wurden.
+Weiters ausschlaggebend war die gute Unterstützung für UTF-8 codiertes Unicode.
+
+Ein Risikopunkt bei dieser Wahl war die Unterstützung für 10.000e bestehende Templates und die Python-Ausdrücke, die diese enthalten.
+Mögliche Varianten waren hier eine automatisierte Übersetzung von Python auf Perl und das Einbinden eines Python-Interpreters mittels des Inline::Python Perlmoduls.
+Den Ausschlag für letzteres gab dessen Fortgeschrittenheit im Vergleich zu den bestehenden Übersetzungslösungen.
+
+Die Anwendung wird vom Webserver über das FastCGI Protokoll angesprochen, da \textit{Catalyst} dafür alles nötige bereits mitbringt.
+\textit{memcached} wird als gemeinsamer Objektcache verwendet um die Datenbank zu entlasten.
 

### 2012-06-12 Praktikum.tex
Datenbankaufbau, Klassenhierarchie

@@
 
-Ein Risikopunkt bei dieser Wahl war die Unterstützung für 10.000e bestehende Templates und die Python-Ausdrücke, die diese enthalten.
+Ein Risikofaktor bei dieser Wahl war die Unterstützung für 10.000e bestehende Templates und die Python-Ausdrücke, die diese enthalten.
 Mögliche Varianten waren hier eine automatisierte Übersetzung von Python auf Perl und das Einbinden eines Python-Interpreters mittels des Inline::Python Perlmoduls.
@@
 \textit{memcached} wird als gemeinsamer Objektcache verwendet um die Datenbank zu entlasten.
+
+\section{Datenbankaufbau}
+
+Jede Website besitzt ein eigenes Datenbankschema, das in einem PostgreSQL-schema, benannt nach dem Domainnamen der Website, enthalten ist.
+Für jede Objektart gibt es hier eine eigene Tabelle.
+Alle Typen sind von einer gemeinsamen Basisklasse namens \textit{ZMS::Object} abgeleitet.
+Das spiegelt sich im Datenbankschema wieder.
+In einer Tabelle namens \textit{zms\_object} werden für jedes gespeicherte Objekt die Eigenschaften der Basisklasse gespeichert.
+Das sind neben der von der Datenbank vergebenen, Website-weit eindeutigen \textit{uid} und einer textuellen \textit{id}, die nur unter den Kindobjekten eines gemeinsamen Elternobjektes eindeutig sein muss, vor allem auch der Name der Tabelle, in der die typspezifischen Daten gespeichert sind.
+Weiters finden sich dort ein Verweis auf das Elternelement, Informationen zur Sortierreihenfolge, das \textit{active} Flag und Änderungsdaten.
+Für jeden Typ gibt es neben einer eigenen Tabelle auch eine writable View die die jeweilige Tabelle und \textit{zms\_object} zusammenführt.
+
+Das Datenbankschema selbst ist in den Tabellen \textit{zms\_meta\_objects} und \textit{zms\_object\_attrs} beschrieben, wo es für jeden Typ bzw.
+jedes Attribut eines Typs Einträge gibt.
+Diese Tabellen werden zusammen mit \textit{zms\_languages}, welche Informationen über die in dieser Website verfügbaren Sprachen enthält, zu Beginn jedes Requests gelesen.
+
+Die Objekte einer Website werden in einem Baum verwaltet.
+Um Abfragen auf Teilbäume effizienter zu machen, werden in der Tabelle \textit{zms\_object\_tree} Informationen über die Baumstruktur gespeichert.
+Sie erlaubt effiziente Abfragen auf komplette Unterbäume mit optionaler Begrenzung der Suchtiefe.
+Die Informationen werden von Triggern in der Datenbank automatisch bei Änderungen an den Daten aktualisiert.
+
+Mehrsprachigkeit wird unterstützt indem bei mehrsprachigen Objekteigenschaften jeweils ein Datenbankfeld für jede Sprache angelegt wird.
+Ist ein Feld in einer Sekundärsprache NULL wird ein Fallback auf die für diese Sprache angegebene Vorgängersprache durchgeführt.
+Werden neue Sprachen hinzugefügt, oder bestehende entfernt müssen daher alle Datenbanktabellen und Views entsprechend angepasst werden.
+
+\section{Klassenhierarchie}
+
+\textit{ZMS::Object} ist die Basisklasse für alle vom CMS verwalteten Objekte und enthält den Großteil des Interfaces für Templatecode.
+\textit{ZMS::ContainerObject} ist davon abgeleitet und ist ihrerseits die Basisklasse für alle Objekte, die sich wie Ordner verhalten und Kindobjekte im Baum enthalten können.
+Von diesen beiden Klassen sind alle Klassen von Inhaltsobjekten abgeleitet.
+Vom Benutzer erstellte Objekttypen werden von \textit{ZMS::Custom} repräsentiert.
+
+Die von den Inhaltsobjekten unterstützten Attribute sind als von \textit{ZMS::Attribute} abgeleitete Klassen implementiert.
+Sie enthalten Informationen über den möglicherweise benutzten Datentyp eines Datenbankfeldes, Code zum Rendern des Eingabeinterfaces und für die Eingabevalidierung.
+
+\textit{Catalyst} gibt eine Grundstruktur für den Aufbau der Anwendung mit Unterteilung in Models, Views und Controllern vor.
+\textit{ZMS::Model::DB} implementiert alle low level Datenbankzugriffe und generiert daher auch alle verwendeten SQL Statements.
+Um die Ausgabe der Website kümmert sich \textit{ZMS::Controller::Content}.
+Die Verwaltung der Inhalte wird von \textit{ZMS::Controller::Content::Management} gestellt.
+Die Verwaltung der Metadaten wie Typen, Sprachen und Benutzer finden sich im \textit{ZMS::Controller::System} Namensraum.
 

### 2012-06-19 Praktikum.tex
Section on DTML

@@
 
+\section{DTML}
+
+In \textit{ZMS::View::DTML} ist das Parsen und Ausführen von DTML Templates implementiert. DTML Anweisungen folgen ähnlich zu HTML immer dem Aufbau:
+<dtml-\textit{command} \textit{options}...
+\textit{attribute}="\textit{value}"...
+\textit{options}...>.
+Sie sind dabei jedoch unabhängig von der HTML Struktur und können an beliebigen Stellen des HTML Codes vorkommen.
+Beim Parsen wird das Templates mit Hilfe einer Regular Expression in Chunks unterteilt und dabei gleichzeitig der DTML-Ausdruck selbst geparst.
+Daraus wird anschließend Perlcode generiert, der mit mittels eval Anweisung ausgeführt wird.
+Für jede DTML-Anweisung konnte eine Perl-Entsprechung gefunden werden.
+
+Der generierte Perlcode wird in einem Cache-File zwischengespeichert. Benannt wird die Datei nach dem DTML Template: headCSS.dtml wird zu .headCSS.dtml.cache
+
+\section{Inline::Python}
+
 \chapter{Erfahrungen und Zusammenfassung}

### 2012-06-19 Praktikum.tex
Inline::Python und Tests

@@
 
+Inline::Python ist ein Perl-Modul, das einen Python-Interpreter in den Perl-Prozess einbindet.
+Diesen Interpreter kann man benutzen um im selben Prozess Python-Code auszuführen.
+Inline::Python kümmert sich dabei darum, die Daten von Perl nach Python bzw.
+umgekehrt zu übersetzen, sodass das Verwenden von Pythonfunktionen von Perlcode heraus und umgekehrt völlig transparent erscheint.
+
+Das Modul wurde von Neil Watkiss entwickelt, der jedoch die Weiterentwicklung nach Version 0.22 vom Jänner 2005 aufgeben musste.
+Diese Version konnte bereits vieles von der benötigten Funktionalität, enthielt jedoch noch sehr viele Fehler im Speichermanagement.
+Ich habe daher die Weiterentwicklung offiziell übernommen und war seither für alle Releases (aktuell ist Version 0.40) verantwortlich.
+Die Fehler rührten daher, dass beide Programmiersprachen eine Reference Count basierende Garbage Collection benutzen, jedoch mit subtilen Unterschieden.
+Z.B.
+hält in Python der Stack eine implizite Referenz auf den Rückgabewert einer Funktion, während in Perl die Funktion dafür verantwortlich \textit{sv\_2mortal()} aufzurufen, womit der Referenzzähler nach dem Ausführen der nächsten Perl-Anweisung heruntergezählt wird.
+Abgesehen davon muss in beiden Sprachen die Referenzzählung in C Code manuell mittels Aufruf entsprechender Makros durchgeführt werden.
+Diese Zählung war größtenteils falsch.
+Es wurde sowohl an den falschen Stellen hinauf- wie hinuntergezählt.
+Das hat dazu geführt, dass Inline::Python zwar oberflächlich funktioniert hat, aber viel Speicher verloren ging.
+Die Fehler haben sich teilweise gegenseitig aufgehoben, sodass beim Beheben eines Fehlers an einer Stelle dafür ein Fehler an einer anderen aufgetaucht ist.
+Beheben ließ sich das durch die Erweiterung der Testsuite um umfangreiche Reference Counting Tests und genaues Studium der Entwicklerdokumentation und des Quellcodes sowohl von Python als auch von Perl.
+
+\section{Tests}
+
+Vor der Übernahme der bestehenden Websites in das neue CMS wurden umfangreiche Tests durchgeführt. Diese liefen in drei Phasen ab:
+
+\subsection{Manuelle Tests}
+
+Zufällig ausgewählte Websites wurden manuell durchgeprüft, ob die Ausgabe mit der Originalversion übereinstimmt oder ob unerwartete Fehlermeldungen kommen.
+In der Anfangsphase wurde man hier noch schnell fündig.
+Nach Test und Debugging von ca.
+10 Websites war die Fundquote jedoch zu niedrig um den Aufwand zu rechtfertigen.
+
+\subsection{Tests auf Fehlermeldungen}
+
+In der nächsten Phase wurden alle 1500 Websites ins neue System importiert und mittels eines Webcrawlers nach Fehlermeldungen des CMS durchsucht.
+Auf diese Art wurden ca.
+100 kleinere Inkompatibilitäten gefunden und behoben.
+
+\subsection{Screenshotvergleiche}
+
+Nachdem keine Fehlermeldungen mehr übrig waren wurde der Crawler erweitert, sodass er durch Fernsteuerung eines Browsers automatisiert Screenshots jeder einzelnen Seite jeder einzelnen Website angefertigt hat und mit einem entsprechenden Screenshot der Originalseite verglichen hat.
+Waren die Bilder nicht pixelweise identisch wurde dann ein Differenzbild der Screenshots erzeugt.
+Diese Differenzbilder wurden dann manuell durchkontrolliert und dabei gefundene Fehler behoben.
+
 \chapter{Erfahrungen und Zusammenfassung}

### 2012-06-19 Praktikum.tex
Rest des Praktikumsberichts

@@
 
+Der erste Prototyp wurde von mir geschrieben und der Geschäftsführung präsentiert.
+Vor allem war der direkte Performancevergleich zum Altsystem überzeugend, weshalb das Projekt genehmigt wurde und in die Entwicklung ging.
+Nach Erstellung des Grundsystems kam nach und nach der Rest des Teams zum Projekt dazu.
+Mit der Umstellung auf die Entwicklung im Team wurden auch eine Testsuite eingeführt.
+
 \section{Grundstruktur}
@@
 Abgesehen davon muss in beiden Sprachen die Referenzzählung in C Code manuell mittels Aufruf entsprechender Makros durchgeführt werden.
+
 Diese Zählung war größtenteils falsch.
@@
 
+Das Projekt war in vielerlei Hinsicht sehr lehrreich.
+Auf technischer Seite waren vor allem das Ausreizen der Datenbank zur Speicherung hierarchischer Objektgraphen und die Weiterentwicklung von Inline::Python herausfordernd.
+Letzteres war für mich der erste Kontakt sowohl mit den Interna von Perl wie auch derer von Python.
+Es war für mich außerdem das erste Projekt bei dem eine mehrschichtige Cachinginfrastruktur zum Einsatz kam.
+Der Wert einer umfangreichen und gut strukturierten Testsuite hat sich im Laufe des Projektes mehrfach gezeigt und die Erfahrungen damit sind bereits ins nächste Projekt eingeflossen.
+
+Organisatorisch war es Neuland, ein Projekt dieser Größe neben dem Tagesgeschäft als noch sehr junges Team durchzuführen.
+Zwei der Teammitglieder waren am Anfang noch nicht mal mit Perl vertraut.
+Einen Termindruck von Auftraggeberseite gab es zwar nicht, aber je länger die Entwicklungs- und Umstellungsphase dauerte, desto mehr Websites galt es zu migrieren.
+Die Umsetzer entdeckten in dieser Zeit außerdem viele neue Features des Altsystems, die dann wiederum nachgebaut werden mussten.
+
+Während des Projekts wurde von Subversion auf git als Sourcecodeverwaltung migriert.
+Der Aufwand, sich in das neue System einzulernen wurde jedoch von der besseren Unterstützung bei der Entwicklung mehr als aufgewogen.
+
+Trotz massiver Verbesserung bei Performance und Usability im Vergleich zum abgelösten System lief die Einführung nicht ohne Widerstand von Benutzerseite.
+Dieser begründete sich vor allem auf die nötige Umgewöhnung sowie auf Verunsicherung über die Zukunft der Software, da sie nicht als das Ergebnis eines Teamprojekts wahrgenommen wurde, sondern als meine persönliche Schöpfung.
+Mit etwas Überzeugungsarbeit ließen sich diese Hürden jedoch überspringen.
+
+Abschließend sei die Weitsicht der Unternehmensführung hervorzuheben, die das Risiko einer derartigen Eigenentwicklung eingegangen ist.
+Dem Aufwand stand die zentrale strategische Bedeutung für das weitere Wachstum der Firma gegenüber.
+Die Chance solche Projekte durchführen zu können macht Atikon zu einem derart schönen Arbeitsplatz.
+
 %%%----------------------------------------------------------

### 2012-06-20 Praktikum.tex
Restrukturierung Praktikumsbericht

@@
 
-Atikon EDV \& Marketing GmbH entwickelt Webseiten für hauptsächlich Steuerberater und Ärzte, sowie Software für Steuerberater.
+Atikon EDV \& Marketing GmbH entwickelt Webseiten hauptsächlich für Steuerberater und Ärzte, sowie Software für Steuerberater.
 Als Leiter der Webprogrammierung führe ich ein Team von 5 Programmierern.
@@
 
-\chapter{Ausgangslage}
+\chapter{Projekt CMS 3000}
 
-\section{Das abzulösende System}
+\section{Einleitung}
 
@@
 
-Dementsprechend gestalten sich daher auch die Anforderungen an das Content Management System:
-%
-\begin{itemize}
-\item Unterstützung für 1000e individuelle Websites in einem System
-\item Einfache Verteilung und Updates von gemeinsam genutzten Inhalten
-\item Völlige Freiheit bei Design und Struktur
-\item Individuelle Erweiterungen
-\item Einfache Bedienung für Kunden mit Eigenwartung
-\end{itemize}
+\section{Projektziel}
 
-In den ersten 10 Jahren des Firmenbestehens hat das deutsche Produkt \textit{ZMS} die Anforderungen am besten erfüllt.
+Das Projekt, das in dieser Arbeit beschrieben wird, hatte zum Ziel, das bestehende Content Management System, welches durch das Wachstum des Kundenstammes an seine Grenzen geraten war, durch ein neu entwickeltes abzulösen.
+Ziele waren dabei, die Verbesserung von Peformance, Zuverlässigkeit und Usability sowie das Schaffen von mehr Anpassungsmöglichkeiten an unsere Anforderungen.
+
+\section{Das abzulösende System}
+
+In den ersten 10 Jahren des Firmenbestehens wurde das deutsche Produkt \textit{ZMS} eingesetzt.
 Es handelt sich dabei um freie, GPL lizenzierte Software, die in Python geschrieben ist und auf dem \text{Zope} Application Server basiert.
@@
 
-\chapter{CMS 3000}
+\section{Anforderungen}
+
+Entsprechend der Kundenstruktur gestalten sich auch die Anforderungen an das Content Management System:
+%
+\begin{itemize}
+\item Unterstützung für 1000e individuelle Websites in einem System
+\item Einfache Verteilung und Updates von gemeinsam genutzten Inhalten
+\item Völlige Freiheit bei Design und Struktur
+\item Individuelle Erweiterungen
+\item Einfache Bedienung für Kunden mit Eigenwartung
+\end{itemize}
 
@@
 Andererseits bestand der Wunsch für die bestehenden 1000 Websites einen Upgradepfad zu ermöglichen um den Wartungsaufwand zu minimieren.
+
+\chapter{Umsetzung}
 

### 2012-06-21 Praktikum.tex
Ergebniskapitel

@@
 
+\chapter{Ergebnis}
+
+\begin{figure}
+\includegraphics[width=1\textwidth]{curl_atikon}
+\caption{Durchschnittliche Antwortzeit von www.atikon.com vor und nach der Migration in Sekunden.}
+\label{fig:atikon}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=1\textwidth]{curl_lask}
+\caption{Durchschnittliche Antwortzeit von www.lask.at vor und nach der Migration in Sekunden.}
+\label{fig:lask}
+\end{figure}
+
+Bereits vor den umfangreichen Kompatibilitätstests wurde angefangen, neue Websites nur noch mit dem neuen CMS zu entwickeln.
+Damit konnten einerseits noch Fehler in der Benutzeroberfläche gefunden werden und andererseits weiterer Umstellungsaufwand vermieden werden.
+Die Umstellung der bestehenden Websites wurde dann nach Abschluss der Tests schubweise durchgeführt.
+Wie in Abbildung \ref{fig:atikon} und \ref{fig:lask} zu sehen ist, verbesserten sich die Antwortzeiten der umgestellten Websites teilweise erheblich.
+Durch die einfachere Parallelisierbarkeit sind jetzt auch Besucherabstürme auf der Homepage des LASK kein Problem mehr für den Server.
+Beschwerden über die Performance sind völlig verschwunden.
+
+PostgreSQL hat sich als sehr stabile Basis erwiesen.
+Von Beginn des Projektes an lief es zuverlässig und problemlos.
+War früher ein Update der verteilten Inhalte ein fragiler und langwieriger Prozess, der nur außerhalb der Geschäftszeiten durchgeführt werden konnte, wird es jetzt jederzeit bei Bedarf durchgeführt, da es keinen spür- oder messbaren Einfluss auf die Performance des Systems hat.
+
+Durch den performanten Zugriff auf die SQL Datenbank wurden neue Auswertungen ermöglicht.
+Abfragen, welche Websites welche Inhalte oder Onlinerechner eingebunden haben lassen sich jetzt innerhalb von Sekunden beantworten.
+
+Das Projekt kann daher als voller Erfolg bezeichnet werden.
+
 \chapter{Erfahrungen und Zusammenfassung}

### 2012-06-23 Paper.tex
Remove target page count from headlines
First round of fixes suggested by H. Dobler++

@@
 
-\chapter{Thanks (1)}
-
-Andrew "whiteknight" Whitworth for laying the foundations for my work, being of tremendous help at all stages of the project and for fixing some of my bugs.
-
-Nat "Chandon" Tuck for implementing green thread basics on which my work is based on.
-
-Christoph "cotto" Otto for bringing me into the Parrot project.
-
-Brian "benabik" Gernhardt for lurking in the IRC channel and at least trying to help with my many questions.
-
-Markus "zimmski" Zimmerman for correcting many errors.
-
-\chapter{Abstract (1)}
-
-Parrot is a runtime for dynamically typed programming languages. Despite several attempts in it's 10 years of history, it does provide any support for multithreaded computation.
-
-This paper shows a way to implement threading support in the runtime using a hybrid between lightweight (``green'') threads and operating system threads.
+\chapter{Acknowledgements}
+
+I want to thank:
+\begin{itemize}
+\item{Andrew "whiteknight" Whitworth for laying the foundations for my work, being of tremendous help at all stages of the project and for fixing some of my bugs.}
+\item{Nat "Chandon" Tuck for implementing the green thread basics on which my work is based on.}
+\item{Christoph "cotto" Otto for bringing me into the Parrot project.}
+\item{Brian "benabik" Gernhardt for lurking in the chat channel and at least trying to help with my many questions.}
+\item{Markus "zimmski" Zimmerman for correcting many errors.}
+\end{itemize}
+
+\chapter{Abstract}
+
+Parrot is a runtime system for dynamically typed programming languages. Despite several attempts in it's 10 years of history, it does not provide any support for multithreaded computation.
+
+This thesis shows a way to implement threading support in Parrot using a hybrid approach using a combination of lightweight (``green'') threads and operating system threads.
 These lightweight threads are used as messages in a system where reading shared data is allowed but only the one owner thread may write to it.
-The implementation is described in detail and benchmarks presented demonstrating the viability of this design.
+The implementation of this hybrid approach is described in detail and benchmarks are presented demonstrating the viability of this design.
 
 \begin{german}
-\chapter{Kurzfassung (1)}
+\chapter{Kurzfassung}
 
 Parrot ist eine Laufzeitumgebung f{\"u}r dynamisch typisierte Programmiersprachen.
-Trotz mehrerer Anl{\"a}ufe in seiner zehnj{\"a}hrigen Geschichte, bietet es keine Unterst{\"u}tzung f{\"u}r Multithreading.
-
-Diese Arbeit zeigt einen Weg um Threadingunterst{\"u}tzung in der Laufzeitumgebung zu implementieren mit Hilfe eines hybriden Systems aus leichtgewichtigen und Betriebssystem- Threads.
+Trotz mehrerer Anl{\"a}ufe in seiner zehnj{\"a}hrigen Geschichte, bietet Parrot keine Unterst{\"u}tzung f{\"u}r Multithreading.
+
+Diese Arbeit zeigt einen Weg, um Threadingunterst{\"u}tzung in der Laufzeitumgebung zu implementieren mit Hilfe eines hybriden Systems aus leichtgewichtigen und Betriebssystem-Threads.
 Diese leichtgewichtigen Threads werden als Nachrichten in einem System benutzt, in dem das Lesen von gemeinsamen Daten erlaubt ist, aber nur der eine besitzende Thread darauf schreiben darf.
@@
 
-\chapter{Introduction and Motivation (3)}
-
-\begin{wrapfigure}{r}{0.5\textwidth}
-A simple example of Perl:
-\include{perlexample}
-\end{wrapfigure}
-%
-On July 19th 2000, the Perl 6 design process was announced.
-Perl 5 had been a very flexible and widely used programming language but had started to show its age and suffered from early design decisions.
+\chapter{Introduction and Motivation}
+
+%
+On July 19th 2000, the Perl 6 design process was announced. Perl 5 had been a flexible and widely used programming language but had started to show its age and suffered from early design decisions.
+
+Am example of Perl:
+\input{perlexample}
+
 The Perl interpreter is written in C and has accumulated a lot of cruft over the years.
-The general consensus among the core developers was that the code had reached a state where maintenance was approaching impossibility\cite{Masak40451}.
+The general consensus among the core developers was that the code had reached a state where maintenance was approaching impossibility \cite{Masak40451}.
 An attempt to reimplement these internals had failed but led to the decision that the interpreter for a Perl 6 language should be developed independently of the needs of Perl 5.
-Since the Perl 6 syntax was very much in flux (and parts of it still are) the designers of these new internals tried to work very independently of any syntax related questions\cite{Developer}.
-Taking up the name born in an April Fool's Day joke announcing the merging of the Perl and Python programming languages, these new internals were called Parrot\cite{Perl}.
-
-Parrot evolved from being just the interpreter for the new version of Perl to being a language independent virtual machine providing features like garbage collection, exception handling and dynamic typing.
-At the time when the Parrot project started, the Java and .NET virtual machines were widely used, but both targeted statically typed languages.
+Since the Perl 6 syntax was very much in flux (and parts of it still are) the designers of these new internals tried to work very independently of any syntax related questions \cite{Developer}.
+Taking up the name born in an April Fool's Day joke announcing the merging of the Perl and Python programming languages, these new internals were called Parrot \cite{Perl}.
+
+Parrot evolved from being just the interpreter for the new version of Perl to being a language independent VM providing features like garbage collection, exception handling and dynamic typing.
+At the time when the Parrot project started, the Java and .NET VM were widely used, but both targeted statically typed languages.
 Parrot thus filled a quickly growing niche.
 
-The Perl 6 design process began with asking the Perl users for what they were expecting from the new version of the language.
-The very first feature that got asked for was well integrated multithreading support\cite{RFC1}.
+The Perl 6 design process began with asking the Perl users what they were expecting from the new version of the language.
+The very first feature that got asked for was well integrated multithreading support \cite{RFC1}.
 Perl 5 had two different implementations of thread support.
-In the first model, called 5005threads, data was shared by default and shared access to data had to be explicitly synchronized.
+In the first model, called \textit{5005threads}, data was shared by default and shared access to data had to be explicitly synchronized.
 This was similar to the models used by languages such as C or Java.
-The implementation however suffered from data corruption and crashes and thus was not recommended for production use\cite{ThreadManual}.
-Perl 5.6 introduced the newer model called ithreads, mostly as a way to emulate fork() on Win32 platforms.
-Perl 5.8 exposed this API (Application Programming Interface) to the user of the programming language.
+The implementation however suffered from data corruption and crashes and thus was not recommended for production use \cite{ThreadManual}.
+Perl 5.6 introduced the newer model called \textit{ithreads}, mostly as a way to emulate \textit{fork} on Win32 platforms.
+Perl 5.8 exposed this Application Programming Interface (API) to the user of the programming language.
 In this new model, all data is copied to each thread and afterwards thread local.
 Data must be explicitly shared between threads.
+
 In other words, in Perl threads are not lightweight at all.
@@
 
-Being born at a time when Perl 6 still looked much more similar to Perl 5 than it does nowadays, Parrot's threading support initially was very close to Perl's ithreads model.
-Previous attempts to change this into the more conventional model of data shared by default or implementing new technologies like Software Transactional Memory failed.
+Being born at a time when Perl 6 still looked much more similar to Perl 5 than it does nowadays, Parrot's threading support initially was very close to Perl's \textit{ithreads} model.
+Previous attempts to change this into the more conventional model of data shared by default or implementing new technologies like \textit{Software Transactional Memory} failed.
 For example Parrot has never supported running multiple threads and having garbage collection at the same time.
@@
 
-In the year 2005 development of faster CPUs (Central Processing Units) shifted from increased speed of a single core to adding more cores.
+In the year 2005 development of faster Central Processing Units (CPUs) shifted from increased speed of a single core to adding more cores.
 Modern processors contain up to 12 cores with even mobile phones having up to four.
@@
 In UNIX (and thus Perl) tradition, this is accomplished using multiple processes being a good solution for many use cases.
-For many others like Perl 6's auto threading of hyper operators, the cost of process setup and communication would be prohibitively high except for very large data sets.
+For many others like auto threading of hyper operators in Perl 6, the cost of process setup and communication would be prohibitively high except for very large data sets.
 
@@
 
-Low level programming languages like C provide only the bare necessities, leaving the responsibility for preventing data corruption entirely to the user.
-A high level language like Perl 6 on the other hand offers complex and compound data types, handles garbage collection and a very dynamic object system.
-Even seemingly simple things like a method call can become very complex sequences.
+Low level programming languages like C provide only the bare necessities, leaving the responsibility for preventing data corruption and synchronization entirely to the user.
+A high-level language like Perl 6 on the other hand provides complex and compound data types, handles garbage collection and a very dynamic object system.
+Even seemingly simple things like a method call can become very complex.
 In a statically typed programming language the definition of a class is immutable.
-Thus calling a method on an object contains just the steps of determining the object's class, fetching the required method from this class and calling it.
+Thus, calling a method on an object contains just the steps of determining the object's class, fetching the required method from this class and calling it.
 Calling the same method again can then even omit the first two steps since their results cannot change.
@@
 The inheritance hierarchy of the class may be changed by adding or removing parent classes.
-Methods may be added to or removed from classes (or objects) at runtime and even the way to find a method of a class may change.
+Methods may be added to or removed from classes (or objects) at runtime and even the way how to find a method of a class may change.
 So a simple method call results in the following steps:
@@
 \begin{itemize}
-\item determining the class of the object
-\item determining the method resolution method of the class
-\item finding the actual method to call
-\item calling the method
+\item determining the class of the object,
+\item determining the method resolution method of the class,
+\item finding the actual method to call,
+\item calling the method.
 \end{itemize}
 %
-These steps have to be repeated for every following method call, because their results may change any time.
+These steps have to be repeated for every method call, because the results may change any time.
 In a threaded environment, a thread running in parallel may change the underlying data and meta data in between those sequences and even between those steps.
-As a consequence, this meta data has to be protected from corruption introducing the need for many locks in a very performance critical area.
-
-Many interpreters for dynamic languages like Python or Ruby handle this problem by using a global interpreter lock to effectively serialize all operations.
-This is a proven and reliable way but again leaves much of the hardware's potential unused.
+As a consequence, this meta data has to be protected from corruption introducing the need for locks in a performance critical area.
+
+Many interpreters for dynamic languages like Python \cite{LearningPython} or Ruby \cite{TheRubyProgrammingLanguage} handle this problem by using a global interpreter lock to effectively serialize all operations.
+This is a proven and reliable way but leaves much of the hardware's potential unused.
 
@@
 
-During these years of back and forth and failed attempts of adding threading support to Parrot, the Perl 6 specification evolved to a point where the largest parts of the language were covered and its features implemented in the compilers.
+During the years of back and forth and failed attempts of adding threading support to Parrot, the Perl 6 specification evolved to a point where the largest parts of the language were covered and its features were implemented in the compilers.
 The lack of concurrency primitives in Parrot however prevents any progress in the area of concurrency support.
 
-Before the work on this paper, Parrot did not have any threading support at all. The previous, defunct implementation had been removed.
-
-This paper suggests a new approach based on a hybrid threading system.
-So called green threads are used to simplify the implementation of a nearly lock free multithreading implementation.
-This approach is based on a design by Andrew Whitworth and Nat Tuck.
-The goal of this paper is to demonstrate the advantages of this model and produce a working implementation which can be used to investigate the performance characteristics of a hybrid threading system.
-
-\chapter{Concurrency in other Programming Platforms (5)}
-
-This chapter is talking about programming platforms.
+Before the work on this thesis started, Parrot did not have any threading support at all. The previous, defunct implementation had been removed.
+
+This thesis suggests a new approach based on a hybrid threading system.
+So called \textit{green threads} are used to simplify the implementation of a nearly lock free multithreading implementation.
+This approach is based on a design by Andrew Whitworth and Nat Tuck \cite{WKnightThreading}.
+The goal of this thesis is to demonstrate the advantages of this model and produce a working implementation which can be used to investigate the performance characteristics of a hybrid threading system.
+
+\chapter{Concurrency in other Programming Platforms}
+
+This chapter is about programming platforms.
 A platform is seen as a combination of a programming language and a runtime.
-For example for the Python programming language there are multiple runtimes with different implementations of threading support.
+E.g.
+for the Python programming language there are multiple runtimes with different implementations of threading support.
 
@@
 
-In Java responsibility for preventing concurrency issues is obliged on the user code.
-The language provides synchronization primitives like mutexes, but the interpreter (the Java Virtual Machine, JVM) does not protect itself.
-The class library provides the user with high level data structures explicitly designed for multithreaded scenarios.
-
-Java version 1.1 used ``green threads'' to support multithreaded execution of Java programs.
-Green threads are threads simulated by the virtual machine but unable to use more than one CPU core for processing.
+In Java, the user is responsible for preventing concurrency issues.
+The language provides synchronization primitives like mutexes, but the interpreter (the Java Virtual Machine, JVM) does not protect the consistency of the provided data structures.
+The class library provides the user with high-level data structures explicitly designed for multithreaded scenarios.
+
+Java version 1.1 used \textit{green threads} to support multithreaded execution of Java programs.
+Green threads are threads simulated by the virtual machine (VM) but unable to use more than one CPU core for processing.
 Details are described in chapter~\ref{cha:green_threads}.
-Version 1.2 introduced native OS (Operating System) thread support which since has become the standard way to do multithreading in Java\cite{JavaThreadManual}.
+Version 1.2 introduced native Operating System (OS) threading support which since has become the standard way to do multithreading in Java \cite{JavaThreadManual}.
 
@@
 
-The CPython implementation of the Python runtime uses a Global Interpreter Lock (GIL) to protect its internal consistency\cite{PythonThreadingManual}.
+The CPython implementation of the Python runtime uses a \textit{Global Interpreter Lock} (GIL) to protect its internal consistency \cite{PythonThreadingManual}.
 This is a single lock taken whenever the interpreter executes Python bytecode.
@@
 The drawback is that Python code cannot benefit from having multiple CPU cores available.
-However I/O operations and calls to external libraries are done without holding the GIL, so in applications with multiple I/O bound threads, there may still be a performance benefit from using multithreading.
-
-To actually run Python code in parallel, multiple processes have to be used.
-The \textit{multiprocessing} module provides support for spawning processes using an API similar to the threading module\cite{PythonMultiProcessingManual}.
+However I/O operations and calls to external libraries are executed without holding the GIL, so in applications with multiple I/O bounded threads, there may still be a performance benefit from using multithreading.
+
+To run Python code in parallel, multiple processes have to be used.
+The \textit{multiprocessing} module provides support for spawning processes exposed through an API similar to the \textit{threading} module \cite{PythonMultiProcessingManual}.
 Since processes may not directly access other processes' memory, the \textit{multiprocessing} module provides several means of communication between processes: \textit{Queues}, \textit{Pipes} and shared memory support.
 
-\chapter{Parrot (4)}
-
-Parrot consists of the virtual machine (VM, also called interpreter), and various tools to facilitate the implementation of programming languages on top of the Parrot VM (the Parrot Compiler Toolkit).
-This paper concentrates on the virtual machine itself.
-The interpreter itself is written in C.
-Example code and test cases are written in PIR (Parrot Intermediate Representation) a high level assembly language abstracting away register allocations and function calling conventions.
-
-Contrary to other widely used virtual machines like the JVM or the CLR which are stack based, Parrot mirrors contemporary hardware CPUs more closely by being register based.
-A stack based virtual machine usually fetches the operands of an operation from the top of a stack and puts the result back on top.
-Thus the operands are chosen implicitly by ordering of the operations allowing the opcodes to be quite small.
-In a register based VM on the other hand each operation has to specify the operands explicitly.
-Compilers for stack machines are simpler because they do not have to care about register allocations and code is independent of prior or subsequent code\cite{VMShowdown}.
-The rationale behind giving up the simplicity of a stack based implementation is the hope of simplifying just in time compilation and improved performance of nested function and method calls.
-
-The current design uses four sets of registers with each and unlimited number of usable registers. The four sets are:
+\chapter{Parrot}
+
+Parrot consists of the VM (also called interpreter), and various tools to facilitate the implementation of programming languages on top of the Parrot VM (the \textit{Parrot Compiler Toolkit}).
+This thesis concentrates on the VM itself.
+The interpreter is written in C.
+Example code and test cases are written in \textit{Parrot Intermediate Representation} (PIR) a high-level assembly language abstracting register allocations and function calling conventions.
+
+Contrary to other widely used VMs like the JVM or the \textit{Common Language Runtime} (CLR) which are stack based, Parrot mirrors contemporary hardware CPUs more closely by being register based.
+A stack based VM usually pops the operands for an operation from the top of a stack and pushes the result back.
+Thus the operands are chosen implicitly by ordering of the operations allowing the opcodes to be without operands.
+In a register based VM on the other hand, each operation has to specify the operands explicitly.
+Compilers for stack machines are simpler because they do not have to care about register allocation and code is independent of prior or subsequent code \cite{VMShowdown}.
+The rationale behind giving up the simplicity of a stack based implementation is the hope of simplifying just-in-time (JIT) compilation and improved performance of nested function and method calls.
+
+The current design of Parrot uses four sets of registers where each one has an unlimited number of registers. The four sets are correspondingly to Parrot's types:
 %
 \begin{itemize}
-\item integer
-\item floating point
-\item string
-\item polymorphic container (PMC)
+\item integer,
+\item floating point,
+\item string and
+\item polymorphic container (PMC).
 \end{itemize}
 
-The first three should be self explanatory.
-Strings in Parrot are a low level type with the interpreter handling all memory allocation issues and Unicode encoding.
-String values themselves are immutable.
-
-\section{PMCs}
+The first three register types are self explanatory. String in Parrot is a low-level type with the interpreter handling all memory allocation issues and Unicode encoding. String values are immutable.
+
+\section{PolyMorphic Containers (PMCs)}
 \label{sec:PMCs}
 
-PMCs (PolyMorphic Container) are containers for all higher level types such as objects, arrays, hash tables or code.
-Thus they are very similar to Python's PythonObject types.
-PMCs are defined by C structs and are fully garbage collected.
+PMCs are containers for all high level types such as objects, arrays, hash tables or code.
+Thus, they are similar to Python's \textit{PythonObject} types.
+PMCs are defined by C structs and are garbage collected.
 Their definition looks like:
@@
 \end{CCode}
-In short they contain a pointer to a type specific data structure and some meta data.
-The vtable is where the type's behavior is defined.
-It contains a long list of function pointers forming Parrot's unified type interface.
+In short, they contain a pointer to a type specific data structure and some meta data.
+The \textit{vtable} is a virtual method table, where the type's behavior is defined.
+It contains a list of function pointers forming Parrot's unified type interface.
 This interface is a union of numeric, string, array, hash and object like behaviors.
 For example the  \textit{get\_integer} function returns an integer value for the data type.
-For a simple \textit{int} it is the contained value.
+For a simple \textit{int} it is the value.
 For an array it may be the number of elements.
@@
 In this way, a language implementer can define the basic types of the language and available operations on them.
-For each vtable entry, there is a corresponding op code in Parrot's bytecode.
-Thus an \textit{inc \$P0} instruction is calling the \textit{increment} vtable function of the PMC contained in the \textit{\$P0} register.
-
-PMCs are implemented in .pmc files with method bodies written in C. These files get preprocessed to plain C before compilation.
-
-\section{ParrotInterpreter}
-
-The interpreter itself is represented by a C struct called \textit{parrot\_interp\_t}.
+For each \textit{vtable} entry, there is a corresponding opcode in Parrot's bytecode.
+Thus an \textit{inc \$P0} instruction calls the \textit{increment} vtable function of the PMC pointed to by the \textit{\$P0} register.
+
+PMCs are implemented in \textit{pmc} files with method bodies written in C. These files are preprocessed to plain C before compilation.
+
+\section{\textit{ParrotInterpreter}}
+
+The Parrot interpreter is represented by a C struct called \textit{parrot\_interp\_t}.
 This structure contains pointers to the garbage collector's runtime data, the loaded types, vtables, the runloop, the current continuation and other global data.
@@
 
+There is also the \textit{ParrotInterpreter} PMC which exposes parts of the \textit{parrot\_interp\_t} struct to the user.
+
 \section{Continuation passing}
@@
 A continuation is a data structure containing the state of a program at a given point in its execution.
-It contains all information needed to continue a program in a certain state, e.g.
+It contains all information necessary to continue a program in a certain state, e.g.
 a call stack, the instruction pointer and contents of local variables.
-``Calling'' a continuation means restoring the state encapsulated in the continuation.
-
-When a function is called, instead of pushing all this information on the stack, the function is given a return continuation as part of its parameters.
+Calling a continuation means restoring the state encapsulated in the continuation.
+
+When a function is called, instead of pushing a return address on the stack, the function is given a return continuation as part of its parameters.
 Returning from a function means calling the return continuation with possible return values stored in the registers where the calling code is expecting them.
-Continuation passing makes things like tail call optimization \footnote{When as the last statement of a function another function is called and its value returned unchanged, the outer function can just pass on its return continuation thus saving the need to create a new continuation and one step on returning from the nested function} very simple and as described in chapter~\ref{cha:green_threads} is a very important piece in implementing green threads.
+Continuation passing makes things like tail call optimization\footnote{When in the last statement of a function another function is called and its value returned unchanged, the outer function can just pass on its return continuation thus saving the need to create a new continuation and one step on returning from the nested function.} simple and as described in chapter~\ref{cha:green_threads} is a important part of implementing green threads.
 
@@
 
-A runloop is in principle the inner most loop executing bytecode in Parrot.
+A runloop is the inner most loop executing bytecode in Parrot.
 It is also a data structure containing some information needed to support exception handling.
@@
 
-Before entering a runloop, the \textit{setjmp} C function is used to save the current stack position and register contents to a data structure stored in the runloop meta data.
-This is effectively creating something resembling a continuation at C level.
+Before entering a runloop, the \textit{setjmp} C function is used to save the current stack position and register contents in a data structure stored in the runloop meta data.
+This is effectively creating something resembling a continuation at the C level.
 When an exception is created within the interpreter, the runloop stack is searched for the runloop containing a suitable exception handler.
-\textit{longjmp} is then used to unwind the call stack up to the point where setjmp was called and the call environment restored.
+\textit{longjmp} is then used to unwind the call stack up to the point where \textit{setjmp} was called and the call environment is restored.
 While having great similarities with continuations, this mechanism is more limited.
-It only allows to jump back to a point higher in the call stack.
-
-\section{Garbage Collector}
-
-Parrot supports different GC (Garbage Collector) implementations which can be selected at interpreter startup. Currently there are four existing implementations of different algorithms:
+It only allows to jump back to a point ``higher'' in the call stack.
+
+\section{Garbage Collector (GC)}
+
+Parrot supports different Garbage Collector (GC) implementations which can be selected at interpreter startup. Currently, there are four implementations of different algorithms:
 \begin{itemize}
-\item Inf: a GC for debugging purposes never collecting any garbage
-\item MS: a basic mark and sweep implementation
-\item MS2: a non-recursive mark and sweep implementation
-\item GMS: a generational, non-compacting, mark and sweep GC
+\item Inf: a GC for debugging purposes never collecting any garbage.
+\item MS: a basic mark and sweep implementation.
+\item MS2: a non-recursive mark and sweep implementation.
+\item GMS: a generational, non-compacting, mark and sweep GC.
 \end{itemize}
-with GMS being the default.
-
-A mark and sweep garbage collector operates in two phases.
-In the mark phase, starting from a known root set of objects, the GC follows pointers down the object tree, marking each encountered object as alive.
-In the sweep phase, all objects not being alive are destroyed and the live flags reset.
-
-In a process with many objects, having to traverse the whole tree may take a considerable amount of time.
-To mitigate this, a generational garbage collector extends this algorithm by assuming that the longer an object has lived, the lower the chances are that it will become unused.
-So the objects are partitioned into different generations.
-The youngest generation would always be traversed while the older generations would be handled much less frequently or even never more at all.
-
-\section{Historical development (1)}
-
-The remainder of the previous threading implementation has been removed with the merging of the kill\_threads branch on September, 21st 2011.
-Previously much threading related code has been removed to clean up the code and improve performance.
+GMS is the default.
+
+A mark and sweep garbage collector operates in two phases: In the mark phase, starting from a known root set of objects, the GC follows pointers in the object graph, marking each encountered object as alive.
+In the sweep phase, all objects not being alive are destroyed and all the live flags are reset.
+
+In a process with many objects, having to traverse the whole graph may take a considerable amount of time.
+To mitigate this, a generational garbage collector extends this algorithm by assuming that the longer an object is alive, the lower the chances are that it will become unused.
+So, the objects are partitioned into different generations.
+The youngest generation will always be traversed while the older generations will be handled much less frequently or even never more at all.
+
+\section{Historical development}
+
+Much of Parrot's previous threading related code has been removed to clean up the code and improve performance.
 Since the existing threading support was known to be unreliable and seriously flawed, this was no trade off.
-
-A year before, Nat Tuck began working on a green\_threads branch during his Google Summer of Code internship.
+The final parts were removed by the merging of the \textit{kill\_threads} branch on September, 21st 2011.
+
+In 2010, Nat Tuck began working on a \textit{green\_threads} branch during his Google Summer of Code internship.
 The feature got prototyped using pure PIR and then implemented in Parrot's core.
 He got it to work in simple cases and started to work on operating system thread support but the internship ended before the code was ready to be merged into the master branch.
-The code lay dormant until the work on this paper started about a year later.
-
-\chapter{Green threads (3)}
+The code lay dormant until the work on this thesis started in 2011.
+
+\chapter{Green threads}
 \label{cha:green_threads}
@@
 
-Operating system threads are like multiple processes running at the same time but share their memory.
+OS threads are like multiple processes running at the same time but share their memory.
 The operating system is responsible for giving each of these threads CPU time to run.
-Operating system threads are the only of these options where more than one thread may be executed by the CPU(s) at the same time if the hardware permits.
+OS threads are the only of these options where more than one thread may be executed by the CPU(s) at the same time if the hardware permits.
 
@@
 
-"Green threads" or "lightweight threads" are threads managed by the virtual machine instead of by the operating system.
+"Green threads" or "lightweight threads" are threads managed by the VM instead of by the operating system.
 For better distinction green threads are henceforth referred to as "tasks".
 These tasks only run pseudo parallel.
-The virtual machine contains a scheduler and support for preempting running tasks.
+The VM contains a scheduler and support for preempting running tasks.
 Preemption means to suspend a task regardless of what it is currently doing and probably resuming it later on.
@@
 \item Garbage collection can be implemented like in a single threaded process.
-\item Critical sections can simply be protected by disabling the scheduler until the code is clear of the section.
+\item Critical sections can be protected by disabling the scheduler until the code is clear of the section.
 \end{itemize}
@@
 
-\chapter{Design of Hybrid Threads (3)}
+\chapter{Design of Hybrid Threads}
 \label{cha:design}
@@
 In keeping the analogy of the interpreter being a software CPU, multithreading is implemented by having a separate interpreter with its own register set for each thread.
-Thus the words ``thread'' and ``interpreter'' are used interchangeably as there is a 1:1 relationship between them.
+Thus the words \textit{thread} and \textit{interpreter} are used interchangeably as there is a 1:1 relationship between them.
 When a user starts a new task, the scheduler first looks for an idle thread.
@@
 It not only increases the instruction count and memory accesses but it also forces the CPU cores to coordinate and thus communicate.
-Even a seemingly simple operation like an atomic increment can take two orders of magnitude longer than a normal increment\cite{LockingInOSKernels}.
+Even a seemingly simple operation like an atomic increment can take two orders of magnitude longer than a normal increment \cite{LockingInOSKernels}.
 While the gain through being able to utilize multiple CPU cores may offset this cost, it is still impacting the very common case of having only a single thread running.
@@
 
-The solution for these problems implemented for this paper is to side step them altogether by disallowing write access to shared data.
+The solution for these problems implemented for this thesis is to side step them altogether by disallowing write access to shared data.
 The programmer (or in most cases the compiler) is obliged to declare a list of all shared variables before a newly created task is started.
@@
 These proxies contain references to the original objects.
-They use these references to forward all reading vtable functions to the originals.
+They use these references to forward all reading \textit{vtable} functions to the originals.
 Write access on the other hand would lead to a runtime error.
@@
 The data owning interpreter would preempt the currently running task and process the new write task.
-Put another way, the data owning interpreter is simply told what to write to its variables, so other threads don't have to.
+Put another way, the data owning interpreter is told what to write to its variables, so other threads don't have to.
 
@@
 
-In this scheme critical sections can simply be implemented by disabling preemption till the code is clear of the section.
+In this scheme critical sections can be implemented by disabling preemption till the code is clear of the section.
 With other threads not being allowed to write to the data and the current task running uninterrupted it it guaranteed to complete without the data being changed from outside.
@@
 In previous attempts to implement threading support in Parrot, the answer to this problem was to disable the garbage collector while multiple threads are active.
-While multithreaded garbage collection algorithms exist, their implementation would have been too complex and brittle for this paper\cite{VCGC}.
+While multithreaded garbage collection algorithms exist, their implementation would have been too complex and brittle for this thesis \cite{VCGC}.
 Other VMs solve this problem by suspending all threads while the GC is running.
@@
 
-\chapter{Implementation of Hybrid Threads (13)}
-
-\section{The Scheduler (3)}
+\chapter{Implementation of Hybrid Threads}
+
+\section{The Scheduler}
 
@@
 \textit{Parrot\_cx\_outer\_runloop} actually consists of two nested loops.
-The inner loop simply fetches tasks and executes as long as tasks are available.
+The inner loop fetches tasks and executes as long as tasks are available.
 Even when all tasks are finished there may still be alarms pending which upon expiry would trigger new tasks to be scheduled.
@@
 
-The \textit{schedule} op code is used to schedule a new task from user code.
+The \textit{schedule} opcode is used to schedule a new task from user code.
 It uses \textit{Parrot\_cx\_schedule\_task} which starts new worker threads if needed and possible and pushes the task on the target thread's scheduler's task queue.
@@
 
-\section{Scheduler PMC (1)}
+\section{Scheduler PMC}
 
@@
 
-\section{Runloops (1)}
+\section{Runloops}
 
 Runloops are created whenever a part of Parrot's C internals starts to execute bytecode.
-This can be simply at interpreter startup to start the actual program to execute or when some library function needs to execute a user callback.
+This can be at interpreter startup to start the actual program to execute or when some library function needs to execute a user callback.
 An exception handler can also be such a callback.
 Each runloop has a unique id.
-These ids are numbered from 0 and simply increased by 1 whenever a new runloop is created.
+These ids are numbered from 0 and increased by 1 whenever a new runloop is created.
 
@@
 
-\section{Threads (8)}
-
-\subsection{Creation (1)}
+\section{Threads}
+
+\subsection{Creation}
 \label{subsec:thread_creation}
@@
 The task is then scheduled with the new interpreter's scheduler using \textit{Parrot\_thread\_schedule\_task}.
-This function uses \textit{Parrot\_thread\_create\_local\_task} to create a corresponding task on the new interpreter and then simply pushes it onto the scheduler.
+This function uses \textit{Parrot\_thread\_create\_local\_task} to create a corresponding task on the new interpreter and then pushes it onto the scheduler.
 Since each interpreter has its own garbage collector, all objects used on the interpreter must origin from this GC's memory pools.
@@
 
-\subsection{Proxies (2)}
+\subsection{Proxies}
 
@@
 This preprocessing gets done by the \textit{tools/build/pmc2c.pl} Perl script.
-This script got extended by implementing the \textit{Parrot::Pmc2c::PMC::Proxy} module which creates default implementations for all vtable functions not otherwise defined in the \textit{proxy.pmc} file.
+This script got extended by implementing the \textit{Parrot::Pmc2c::PMC::Proxy} module which creates default implementations for all \textit{vtable} functions not otherwise defined in the \textit{proxy.pmc} file.
 The default implementation for all writing functions would just call \textit{cant\_do\_write\_method} which creates a runtime exception.
@@
 
-\subsection{Writing shared data (2)}
+\subsection{Writing shared data}
 
@@
 %
-Line 1 simply creates a new Task object.
+Line 1 creates a new Task object.
 The example subroutine is used for the task's \textit{code} attribute.
@@
 
-\subsection{Garbage Collection (3)}
-
-Since each interpreter has its own garbage collector and consequently its own memory areas and can only access its own data directly, the garbage collector can simply act as if there were no other threads.
+\subsection{Garbage Collection}
+
+Since each interpreter has its own garbage collector and consequently its own memory areas and can only access its own data directly, the garbage collector can act as if there were no other threads.
 Thus the main work has been to clean up the interpreter structure itself and to replace references to global data by proxies.
@@
 
-\chapter{Tests and Benchmarks (4)}
+\chapter{Tests and Benchmarks}
 
@@
 It is written in the \textit{Winxed} programming language.
-Winxed is a low level language with JavaScript like syntax and the possibility to include sections of PIR code verbatim making it possible to try experimental opcodes while writing more readable and concise code than with PIR alone.
+Winxed is a low-level language with JavaScript like syntax and the possibility to include sections of PIR code verbatim making it possible to try experimental opcodes while writing more readable and concise code than with PIR alone.
 The complete source code is available in appendix \ref{app:matrix_part}.
@@
 
-\chapter{Conclusion, Further Work and Experiences (3)}
+\chapter{Conclusion, Further Work and Experiences}
 
@@
 
-\MakeBibliography{References, Literature (1)}
+\MakeBibliography{References, Literature}
 

### 2012-06-23 Paper.tex
Second round of fixes suggested by H. Dobler++

@@
 
-This thesis shows a way to implement threading support in Parrot using a hybrid approach using a combination of lightweight (``green'') threads and operating system threads.
-These lightweight threads are used as messages in a system where reading shared data is allowed but only the one owner thread may write to it.
+This thesis shows a way to implement threading support in Parrot using a hybrid approach using a combination of lightweight (``green'') threads and OS threads.
+These lightweight threads are used as messages in a system where reading shared variables is allowed but only the one owner thread may write to it.
 The implementation of this hybrid approach is described in detail and benchmarks are presented demonstrating the viability of this design.
@@
 In other words, in Perl threads are not lightweight at all.
-They have severe impact on memory usage, writes to shared data are expensive and still not all features of the language are usable in threaded programs.
+They have severe impact on memory usage, writes to shared variables are expensive and still not all features of the language are usable in threaded programs.
 
@@
 
-A mark and sweep garbage collector operates in two phases: In the mark phase, starting from a known root set of objects, the GC follows pointers in the object graph, marking each encountered object as alive.
+A mark and sweep GC operates in two phases: In the mark phase, starting from a known root set of objects, the GC follows pointers in the object graph, marking each encountered object as alive.
 In the sweep phase, all objects not being alive are destroyed and all the live flags are reset.
@@
 In a process with many objects, having to traverse the whole graph may take a considerable amount of time.
-To mitigate this, a generational garbage collector extends this algorithm by assuming that the longer an object is alive, the lower the chances are that it will become unused.
+To mitigate this, a generational GC extends this algorithm by assuming that the longer an object is alive, the lower the chances are that it will become unused.
 So, the objects are partitioned into different generations.
@@
 The feature got prototyped using pure PIR and then implemented in Parrot's core.
-He got it to work in simple cases and started to work on operating system thread support but the internship ended before the code was ready to be merged into the master branch.
+He got it to work in simple cases and started to work on OS thread support but the internship ended before the code was ready to be merged into the master branch.
 The code lay dormant until the work on this thesis started in 2011.
@@
 
-Green threads are one way to model concurrent control flows in a program. To get a better understanding of what they are and how they work, the possible options are discussed first.
+Green threads are one way to model concurrent control flows in a program.
+To get a better understanding of what they are and how they work, the possible options for supporting concurrency are discussed before explaining green threads in detail.
 
@@
 \end{GenericCode}
-This function would return one integer for every call, counting from 0 to 100. An example usage would look like:
+This function returns one integer for every call, counting from 0 to 100. An example usage looks like:
 \begin{GenericCode}
+def count\_up():
     x = counter()
@@
 \end{GenericCode}
-This example would print the numbers 0 and 1.
-For the two functions to run in parallel, they have to cooperate.
-One by calling the other repeatedly and the callee by yielding control back to the caller.
+This example prints the numbers 0 and 1. For the two functions to run in parallel, they have to cooperate. One by calling the other repeatedly and the callee by yielding control back to the caller.
 
@@
 
-OS threads are like multiple processes running at the same time but share their memory.
-The operating system is responsible for giving each of these threads CPU time to run.
-OS threads are the only of these options where more than one thread may be executed by the CPU(s) at the same time if the hardware permits.
+OS threads are like multiple processes running in parallel but share their memory.
+The OS is responsible for giving each of these threads CPU time to run.
+OS threads are the only option where more than one thread may be executed by the CPU(s) at the same time if the hardware permits.
 
@@
 
-"Green threads" or "lightweight threads" are threads managed by the VM instead of by the operating system.
-For better distinction green threads are henceforth referred to as "tasks".
+\textit{Green threads} or \textit{lightweight threads} are threads managed by the VM instead of by the OS.
+For better distinction green threads are henceforth referred to as \textit{tasks}.
 These tasks only run pseudo parallel.
-The VM contains a scheduler and support for preempting running tasks.
+The VM has a scheduler and support for preempting running tasks.
 Preemption means to suspend a task regardless of what it is currently doing and probably resuming it later on.
-This is an implementation of the many-to-one threading model very similar to what an operating system running on a system with a single CPU core does.
+This is an implementation of the many-to-one threading model very similar to what an OS running on a system with a single CPU core does.
 %
@@
 \item They allow pseudo concurrent processing without endangering the interpreter's internal consistency.
-\item Green threads are very light weight with low memory overhead and close to zero creation time.
-\item They do not depend on OS threading support
+\item Green threads are light weight having low memory overhead and close to zero creation time.
+\item They do not depend on OS threading support.
 \item The interpreter controls the point of preemption of a green thread.
-\item The interpreter controls the exact scheduling policy and may allow the user to influence it or even take over completely.
-\item Garbage collection can be implemented like in a single threaded process.
+\item The interpreter controls the scheduling policy and may allow the user to influence it or even take over completely.
+\item Garbage collection can be implemented like in a singlethreaded process.
 \item Critical sections can be protected by disabling the scheduler until the code is clear of the section.
@@
 \begin{itemize}
-\item They do not allow more than one CPU core to be used for computations.
-\item Blocking calls like I/O block the whole interpreter including other green threads.
-\item The interpreter has to contain logic and timers to control green threads.
+\item Green threads do not allow more than one CPU core to be used for computations.
+\item Blocking calls like I/O block the interpreter including other green threads.
+\item The interpreter has to provide logic and timers to control green threads.
 \end{itemize}
 
-Green threads differ from coroutines in that the interpreter decides when a running task is to be preempted while a coroutine depends on manual yield calls.
-
-Following is a simple example to explain why concurrent processing can endanger the interpreter's internal consistency.
-The basic problem is concurrent writes to shared data.
+Green threads differ from coroutines in that the interpreter decides when a running task is to be preempted while a coroutine depends on explicit yield calls.
+
+Below is an example in order to explain why concurrent processing can endanger the interpreter's internal consistency.
+The basic problem are concurrent writes to shared variables.
 Assuming an implementation of an array class consisting of the field holding the data and the number of contained elements in a separate member variable:
@@
 \end{CCode}
-to append a new value, the array would have to read the current size, write the new value at the position size + 1 and then write the incremented size back into the member variable.
-Now if two threads simultaneously try to do this, it may happen that both read the same size, write to the same position (with one overwriting the value of the other) and write back the same incremented size.
-In this case one of the appended values would be lost.
-
-In a more complicated example, the array could have to resize its data buffer to accommodate the new value.
-It would read the current size, allocate a new buffer, copy the values to the new buffer and then destroying the old one.
-Again if two threads try to do this simultaneously, one of them could still be copying data, while the other is already at the point where it destroys the old buffer.
-This would lead to the copying thread accessing freed memory.
-
-When garbage collection is brought into the mix, the possibilities for corruption grow even more.
-At the point where the first thread needs to allocate the new buffer, the garbage collector could decide that it needs to run and clear some unused memory.
-It would have to traverse the object tree to mark alive objects.
-Between the mark and the sweep runs, the second thread could change the pointer to the buffer from the old one to the new one.
-So the old buffer would have been marked alive, while during the sweep run the new buffer would be in its place, but not yet marked alive.
+To append a new value, the array has to read the current size, write the new value at the position \textit{size} + 1 and then write the incremented size back into the member variable.
+Now, if two threads simultaneously try to do this, it may happen that both read the same size, write to the same position (with one overwriting the value of the other) and write back the same incremented size.
+In this case one of the appended values is lost.
+
+In a more complicated example, the array could have to resize its data buffer to accept the new value.
+It would read the current size, allocate a new buffer, copy the values to the new buffer and then destroy the old one.
+Again, if two threads try to do this simultaneously, one of them could still be copying data, while the other already destroys the old buffer.
+This leads to the copying thread accessing freed memory.
+
+When garbage collection is brought into the mix, the possibilities for corruption grow even further.
+At the point where the first thread needs to allocate the new buffer, the GC could decide that it needs to clear some unused memory.
+It would have to traverse the object graph to mark alive objects.
+Between the mark and the sweep phases, the second thread could change the pointer to the buffer from the old one to the new one.
+So the old buffer would have been marked alive, while during the sweep phase the new buffer would be in its place, but not yet marked alive.
 
@@
 
-Parrot's green threads implementation is based on Nat Tuck's green\_threads branch developed during his Google Summer of Code internship.
-
-In Parrot green threads are called \textit{Task}.
+Parrot's green threads implementation is based on Nat Tuck's \textit{green\_threads} branch developed during his Google Summer of Code internship \cite{ChandonBlog}.
+
+In Parrot, green threads are called \textit{Tasks}.
 Each task is assigned a fixed amount of execution time.
-On expiry a timer callback sets a flag which is checked for at execution of every branch operation.
+After this time is up a timer callback sets a flag which is checked at execution of every branch operation.
 Since the interpreter's state is well defined at this point, its internal consistency is guaranteed.
-The same goes for the garbage collector.
-Since task preemption is only done while executing user level code, the GC can do its work undisturbed and without the need for measures like locking.
-Since user level code is allowed to disable the scheduler, it can be guaranteed to run undisturbed through critical sections.
-
-The scheduler is implemented as a PMC (Parrot's polymorphic container) type.
-The plan is to allow the user to subclass this PMC thus allowing very fine grained control over the scheduling policy.
-Features a user could add this way would be for example giving different priorities to tasks or implementing the possibility to suspend and resume a task.
+The same holds for the GC.
+Since task preemption is only done while executing user-level code, the GC can do its work undisturbed and without the need for measures like locking.
+Since user-level code is allowed to disable the scheduler, it can be guaranteed to run undisturbed through critical sections.
+
+The scheduler is implemented as a PMC type.
+This allows the user to subclass this PMC thus allowing fine-grained control over the scheduling policy.
+Features, a user could add this way would be for example giving different priorities to tasks or implementing the possibility to suspend and resume a task.
 
@@
 
-This chapter describes how green threads were used to solve the following problems occurring when trying to implement threading support:
+This chapter describes how green threads are used to solve the following problems occurring when trying to implement threading support:
 %
 \begin{itemize}
-\item How to ensure internal interpreter consistency when doing writes to shared data.
-\item How to implement critical sections.
-\item How to handle garbage collection.
+\item How to ensure internal interpreter consistency when doing writes to shared variables?
+\item How to implement critical sections?
+\item How to handle GC?
 \end{itemize}
@@
 
-As described in the introduction, cross thread writes to shared data may endanger the internal consistency of the interpreter.
-Traditionally the response to this problem is the use of locks of varying granularity.
-Fine grained locking allows code to run well parallelized but taking and releasing locks also costs performance.
+As described in the introduction, cross-thread writes to shared variables may endanger the internal consistency of the interpreter.
+Traditionally, the solution to this problem is the use of locks of varying granularity.
+Fine-grained locking allows code to run in parallel but taking and releasing locks costs performance.
 It not only increases the instruction count and memory accesses but it also forces the CPU cores to coordinate and thus communicate.
 Even a seemingly simple operation like an atomic increment can take two orders of magnitude longer than a normal increment \cite{LockingInOSKernels}.
-While the gain through being able to utilize multiple CPU cores may offset this cost, it is still impacting the very common case of having only a single thread running.
-
-Too coarse locking on the other hand would reduce scalability and the performance gains through parallel execution by having threads wait for extended periods for locks to become open.
-In the extreme case of having a global interpreter lock it would even effectively serialize all computations costing much of the benefits of using threads in the first place.
-
-The other problem with locking is the possibility of introducing dead locks.
-For example, two functions F1 and F2 both use two resources A and B protected by locks.
-If F1 first locks A and then tries to lock B while F2 has already locked B and is now trying to lock A, the program would come to a halt.
+While the gain through being able to utilize multiple CPU cores may offset this cost, it is still impacting the common case of having only a singlethread running.
+
+Too coarse locking on the other hand would reduce scalability and the performance gains through parallel execution by having threads wait for extended periods for locks to become available.
+In the extreme case of having a global interpreter lock it would effectively serialize all computations costing much of the benefits of using threads in the first place.
+
+The other problem with locking is the possibility of introducing deadlocks.
+For example, two functions \textit{F1} and \textit{F2} both use two resources \textit{A} and \textit{B} protected by locks.
+If \textit{F1} first locks \textit{A} and then tries to lock \textit{B} while \textit{F2} has already locked \textit{B} and is now trying to lock \textit{A}, the program would come to a halt.
 Both functions would be left waiting for the other to unlock the resource which will never happen.
-With fine grained locking the possibilities for such bugs grow quickly.
-At the same time it is easy to miss a case where a lock would be appropriate leading to difficult to diagnose corruption bugs.
-
-The solution for these problems implemented for this thesis is to side step them altogether by disallowing write access to shared data.
+With fine-grained locking, the possibilities for such bugs grow quickly.
+At the same time, it is easy to miss a case where a lock would be appropriate leading to difficult to diagnose corruption bugs.
+
+The solution for these problems implemented in this thesis is to side step them altogether by disallowing write access to shared variables.
 The programmer (or in most cases the compiler) is obliged to declare a list of all shared variables before a newly created task is started.
@@
 
-In other words all data is owned by the thread creating it and only the owner may write to it. Other threads have only read access.
-
-For threads to be able to communicate with their creators and other threads, they still need to write to shared data.
+In other words, all data is owned by the thread creating it and only the owner may write to it. Other threads have only read access.
+
+For threads to be able to communicate with their creators and other threads, they still need to write to shared variables.
 This is where green threads come into play.
-Since green threads are very light weight, it is feasible for a thread to create a task just for updating a variable.
-This task would be scheduled on the interpreter owning this variable.
+Since green threads are light weight, it is feasible for a thread to create a task just for updating a variable.
+This task is scheduled on the interpreter owning this variable.
 To reduce latency, the task is flagged to run immediately.
-The data owning interpreter would preempt the currently running task and process the new write task.
-Put another way, the data owning interpreter is told what to write to its variables, so other threads don't have to.
+The data-owning interpreter will preempt the currently running task and process the new write task.
+Put another way, the data-owning interpreter is told what to write to its variables, so other threads don't have to.
 
@@
 
-In this scheme critical sections can be implemented by disabling preemption till the code is clear of the section.
+Critical sections can be implemented by disabling preemption until the code has left the section.
 With other threads not being allowed to write to the data and the current task running uninterrupted it it guaranteed to complete without the data being changed from outside.
 
-\section{Garbage Collection}
-
-A garbage collector has to traverse and process the entire tree of objects in memory.
-If during this traversal another thread is changing this tree by creating new objects, moving them around in the tree or removing them, the garbage collector's data would be compromised.
-For example it could happen that the GC does not mark referenced objects alive because they were added to parents having already been processed.
-
-In previous attempts to implement threading support in Parrot, the answer to this problem was to disable the garbage collector while multiple threads are active.
-While multithreaded garbage collection algorithms exist, their implementation would have been too complex and brittle for this thesis \cite{VCGC}.
+\section{Garbage Collection (GC)}
+
+A GC has to traverse and process the entire graph of objects in memory.
+If during this traversal another thread is changing this structure by creating new objects, moving them around in the graph or removing them, the GC's data would be compromised.
+For example, it could happen that the GC does not mark referenced objects alive because they were added to objects which have already been processed.
+
+In previous attempts to implement threading support in Parrot, the solution to this problem was to disable the GC while multiple threads were active.
+While multithreaded GC algorithms exist, their implementation would have been too complex and brittle for this thesis \cite{VCGC}.
 Other VMs solve this problem by suspending all threads while the GC is running.
@@
 
-Since cross threaded writes are already forbidden and all read access to other thread's data goes through the very narrow channel of proxy objects, forcing a complete separation of the thread's memory domains was only a small step.
-By having separate memory areas for each interpreter, it became possible to have each interpreter to run its own garbage collector.
-This way, the known to work, single threaded GC implementation can be used nearly unchanged.
-
-With separated memory domains it can happen that an object is created on one thread, used on another thread but not any more on the owner thread.
+Since cross-threaded writes are already forbidden and all read access to other thread's data goes through the narrow channel of proxy objects, forcing a complete separation of the thread's memory domains is only a small step.
+By having separate memory areas for each interpreter, it becomes possible to have each interpreter to run its own GC.
+This way, the known to work singlethreaded GC implementation can be used nearly unchanged.
+
+With separated memory domains, it can happen that an object is created on one thread, used on another thread but not any more on the owner thread.
 Without any additional measures, the GC would collect such objects, since it does not know about references from other threads.
-But since all objects which could be accessed from other threads have to be pushed onto the Task object representing these threads, the objects can still be referenced from the Task object.
+But since all objects which could be accessed from other threads have to be pushed onto the task object representing these threads, the objects can still be referenced from the task object.
 Since the task is still owned by the original thread, the GC knows that these objects are still in use.
@@
 
-\section{The Scheduler}
+\section{Scheduler}
 
 The scheduler is the place where most of the green thread logic is implemented.
-It consists of two parts: functions using the prefix \textit{Parrot\_cx} which are located in \textit{src/scheduler.c} and the Scheduler PMC type containing the scheduler's data and public interface.
-
-\textit{Parrot\_cx\_init\_scheduler} is called from \textit{Parrot\_interp\_initialize\_interpreter} whenever a new interpreter is created, e.g.
-on interpreter startup and whenever a new thread is created.
+It consists of two parts: functions using the prefix \textit{Parrot\_cx} which are located in \textit{src/scheduler.c} and the scheduler PMC type containing the scheduler's data and public interface.
+
+\textit{Parrot\_cx\_init\_scheduler} is called from \textit{Parrot\_interp\_initialize\_interpreter} whenever a new interpreter is created, e.g., on interpreter startup and whenever a new thread is created.
 It creates the scheduler PMC and sets up alarm signal handling.
@@
 This task is then put onto the run queue.
-The scheduler timer gets enabled and control given to \textit{Parrot\_cx\_outer\_runloop}.
-
-\textit{Parrot\_cx\_outer\_runloop} is the core loop taking tasks from the run queue and executing them.
-Despite its name, it should not be confused with runloops discussed in section~\ref{sec:Runloops}.
+The scheduler timer gets enabled and control is given to \textit{Parrot\_cx\_outer\_runloop}.
+
+\textit{Parrot\_cx\_outer\_runloop} is the loop taking tasks from the run queue and executing them.
+Despite its name, it may not be confused with runloops discussed in section~\ref{sec:Runloops}.
 \textit{Parrot\_cx\_outer\_runloop} actually consists of two nested loops.
-The inner loop fetches tasks and executes as long as tasks are available.
-Even when all tasks are finished there may still be alarms pending which upon expiry would trigger new tasks to be scheduled.
-For this reason there's an outer loop checking for pending alarms and if there are putting the thread to sleep until the next alarm expires.
+The inner loop fetches tasks from the task queue and executes as long as tasks are available.
+Even when all tasks are finished there may still be alarms pending which upon termination would trigger new tasks to be scheduled.
+For this reason, there is an outer loop checking for pending alarms.
+If it finds any it puts the thread to sleep until the next alarm expires.
 
@@
 Before calling the task, it checks if there are other tasks still in the queue.
-Only if other tasks are waiting, task preemption gets enabled.
-Otherwise the current task would run until it finishes by itself, schedules other tasks or an alarm expires.
-This is an optimization reducing the runtime overhead of green threads in the important single tasking case to 0.
-
-To enable task preemption, \textit{Parrot\_cx\_enable\_preemption} sets a flag on the scheduler PMC and uses \textit{Parrot\_cx\_set\_scheduler\_alarm} to set an alarm at a point \textit{PARROT\_TASK\_SWITCH\_QUANTUM} milliseconds in the future.
+Only if other tasks are waiting, task preemption is enabled.
+Otherwise, the current task will run until it finishes by itself, schedules other tasks or an alarm expires.
+This is an optimization reducing the runtime overhead of green threads in the important single tasking case to zero.
+
+To enable task preemption, \textit{Parrot\_cx\_enable\_preemption} sets a flag on the scheduler PMC and uses \textit{Parrot\_cx\_set\_scheduler\_alarm} to set an alarm at a point in time \textit{PARROT\_TASK\_SWITCH\_QUANTUM} milliseconds in the future.
 
@@
 This serial is checked for changes by \textit{Parrot\_alarm\_check} being called by \textit{Parrot\_cx\_check\_scheduler}.
-This function in turn is called once for every \textit{branch} op.
-When the serial is different from the last check or the \textit{SCHEDULER\_wake\_requested} flag is set, \textit{Parrot\_cx\_check\_scheduler} wakes the scheduler by calling \textit{Parrot\_cx\_run\_scheduler}.
+This function in turn is called once for every \textit{branch} operation.
+When the serial is different from the last check or the \textit{SCHEDULER\_wake\_requested} flag is set, \textit{Parrot\_cx\_check\_scheduler} wakes up the scheduler by calling \textit{Parrot\_cx\_run\_scheduler}.
 
 Since POSIX alarms can only be delivered to the main thread of a process, \textit{Parrot\_cx\_run\_scheduler} uses \textit{Parrot\_thread\_notify\_threads} to relay alarms to other thread's schedulers.
-For this purpose every thread has a pipe which \textit{Parrot\_thread\_notify\_threads} writes a dummy byte to.
-The next step for \textit{Parrot\_cx\_run\_scheduler} is to check for any expired alarms using \textit{Parrot\_cx\_check\_alarms}.
-An alarm contains a task as callback.
-On expiry this callback gets scheduled for immediate execution.
+For this purpose, every thread has a pipe which \textit{Parrot\_thread\_notify\_threads} writes a dummy byte to.
+The next step for the \textit{Parrot\_cx\_run\_scheduler} is to check for any expired alarms using \textit{Parrot\_cx\_check\_alarms}.
+An alarm has a task as callback.
+On expiry, this callback gets scheduled for immediate execution.
 As alarm signals do not contain any information about their origin, the scheduler does not know if the alarm was scheduled by user code or if it is the scheduler's preemption alarm.
@@
 The task itself is then appended at the end of the run queue.
-The NULL opcode return value is propagated all the way up through the stack back to the \textit{branch} op checking for alarms.
-The op then recognizes this as the signal to stop processing and to exit the runloop.
+The function returns a \textit{NULL} opcode which is propagated all the way up through the stack back to the \textit{branch} operation checking for alarms.
+The operation then recognizes this as the signal to stop processing and to exit the runloop.
 
@@
 
-\textit{Parrot\_cx\_schedule\_immediate} is used in various places like at alarm expiry for putting a task to the head of the run queue and immediately causing preemption of the currently running task.
+\textit{Parrot\_cx\_schedule\_immediate} is used in various places like at alarm expiry for putting a task at the head of the run queue and immediately causing preemption of the currently running task.
 It does this by setting \textit{SCHEDULER\_wake\_requested} and \textit{SCHEDULER\_resched\_requested} flags the same as when the preemption alarm expires.
-To be precise this mechanism leads to the current task being preempted at the next \textit{branch} op since this is the place when the mentioned flags will be checked.
+To be precise, this mechanism leads to the current task being preempted at the next \textit{branch} operation since this is the place where the mentioned flags will be checked.
 
 Alarms can be registered using \textit{Parrot\_cx\_schedule\_alarm}.
-It puts the alarm in the appropriate place in the ordered alarm list and uses \textit{Parrot\_alarm\_set} which sets the actual alarm but only if there's not another alarm already set for an earlier time since there can only be one POSIX alarm pending at any time.
+It puts the alarm in the appropriate place in the ordered alarm list and uses \textit{Parrot\_alarm\_set} which sets the actual alarm but only if there is not another alarm already set for an earlier time, since there can only be one POSIX alarm pending at any time.
 
@@
 Like in \textit{Parrot\_cx\_preempt\_task}, \textit{Parrot\_cx\_stop\_task} is used to get an updated task at the current execution position, but instead of pushing this on the run queue, it is used as a callback for a newly set alarm.
-Again the NULL opcode is used to stop processing of the current task.
+Again, the \textit{NULL} opcode is used to stop processing of the current task.
 
@@
 \begin{itemize}
-\item PMC *task\_queue: List of tasks/green threads waiting to run
-\item Parrot\_mutex task\_queue\_lock: a lock protecting the task\_queue so other threads can access it safely
-\item PMC *alarms: List of future alarms ordered by time
-\item PMC *all\_tasks: Hash of all active tasks by ID
-\item UINTVAL next\_task\_id: ID to assign to the next created task
-\item Parrot\_Interp interp: A reference to the scheduler's interpreter.
+\item \textit{PMC *task\_queue}: list of tasks/green threads waiting to run.
+\item \textit{Parrot\_mutex task\_queue\_lock}: a lock protecting the task\_queue so other threads can access it safely.
+\item \textit{PMC *alarms}: list of future alarms ordered by time.
+\item \textit{PMC *all\_tasks}: hash of all active tasks by ID.
+\item \textit{UINTVAL next\_task\_id}: ID to assign to the next created task.
+\item \textit{Parrot\_Interp interp}: a reference to the scheduler's interpreter.
 \end{itemize}
 
-\textit{push\_pmc} and \textit{unshift\_pmc} are used to add a task at the end respectively the beginning of the task\_queue.
+\textit{push\_pmc} and \textit{unshift\_pmc} are used to add a task at the end, respectively the beginning of the task\_queue.
 \textit{shift\_pmc} is used to fetch and remove a task from the tip of the \textit{task\_queue}.
@@
 These four methods use the \textit{task\_queue\_lock} to make accessing the \textit{task\_queue} thread safe.
-Thus the Scheduler PMC acts as a container while using \textit{task\_queue} to actually store the data.
-
-\textit{alarms} is list containing all pending alarms, sorted by their expiry time.
+Thus, the Scheduler PMC acts as a container while using \textit{task\_queue} to actually store the data.
+
+\textit{alarms} is list containing all pending alarms, sorted by their expiration time.
 Sorted insertion is used to keep alarms in order.
@@
 
-The \textit{active\_tasks} returns an array containing all tasks which have been run at least once and are not yet finished, e.g. they are currently being executed or are preempted.
+The \textit{active\_tasks} returns an array containing all tasks which have been run at least once and are not yet finished, e.g., they are currently being executed or are preempted.
 
@@
 \begin{itemize}
-\item UINTVAL id: Unique identifier for the task
-\item FLOATVAL birthtime: The creation time stamp of the task
-\item Parrot\_Interp interp: The interpreter which created the task
-\item PMC *code: The code to run
-\item PMC *data: Additional data for the task given as parameter to \textit{code}
-\item INTVAL killed: Flag marking killed tasks
-\item PMC *waiters: Tasks waiting on this one
-\item PMC *shared: List of variables shared with this task
-\item PMC *partner: Copy of this task on the other side of a GC barrier, meaning in another thread
+\item \textit{UINTVAL id}: Unique identifier for the task.
+\item \textit{FLOATVAL birthtime}: the creation time stamp of the task.
+\item \textit{Parrot\_Interp interp}: the interpreter which created the task.
+\item \textit{PMC *code}: the code to run.
+\item \textit{PMC *data}: additional data for the task given as parameter to \textit{code}.
+\item \textit{INTVAL killed}: flag marking killed tasks.
+\item \textit{PMC *waiters}: tasks waiting on this one.
+\item \textit{PMC *shared}: list of variables shared with this task.
+\item \textit{PMC *partner}: copy of this task on the other side of a GC barrier, meaning in another thread.
 \end{itemize}
@@
 It first checks the \textit{killed} flag to see if the task has been killed while waiting in the task queue.
-If it is still alive, the current recursion depth is saved to a local variable.
-This is a workaround needed because \textit{Parrot\_Sub\_invoke}, which ends up being called when invoking the task's \textit{code}, increments the recursion depth, but it does not get decremented anymore in case the task got preempted.
-This would lead to an ``recursion depth exceeded'' exception when many tasks are invoked during a program's runtime.
+Parrot manages a \textit{recursion depth} counter which records, how many levels the call stack has.
+This counter is incremented by \textit{Parrot\_Sub\_invoke} for each call to a subroutine and decremented when a subroutine returns.
+While \textit{Parrot\_Sub\_invoke} is used to start or resume a task, the executed subroutine does not return when the task is preempted.
+This leads to \textit{recursion depth} growing until a \textit{recursion depth exceeded} exception is thrown.
+Therefore, as a workaround, the current \textit{recursion depth} is saved to a local variable and restored when the task stops executing.
+
 The next steps are to add the task to the list of active tasks and to invoke the task's \textit{code}.
@@
 
-\textit{push\_pmc} and \textit{pop\_pmc} are used to add or remove variables to the task's list of shared data. For these variables, proxies will be created when the task is run on a different thread.
+\textit{push\_pmc} and \textit{pop\_pmc} are used to add or remove variables to the task's list of shared variables.
+For these variables, proxies will be created when the task is run on a different thread.
 
@@
 Runloops are created whenever a part of Parrot's C internals starts to execute bytecode.
-This can be at interpreter startup to start the actual program to execute or when some library function needs to execute a user callback.
+This can be at interpreter startup to start the actual program execution or when some library function needs to execute a user callback.
 An exception handler can also be such a callback.
@@
 As part of exception handling finalization, it cleans up, freeing any information no longer needed.
-The exception handler does not necessarily have to have been set up in the same runloop from which the exception has been thrown.
+The exception handler does not have to have been set up in the same runloop from which the exception has been thrown.
 Parrot uses the runloop id to identify such situations and free the nested runloops up to the level of the exception handler.
@@
 A continuation also contains the id of the runloop in which it has been created.
-When resuming a preempted task, this id would not be the same as the one of the runloop created for executing its code.
-This would lead to Parrot not finding the correct runloop when finalizing an exception.
-
-To mitigate this, the \textit{reset\_runloop\_id\_counter(interp)} function is used by \textit{Parrot\_cx\_outer\_runloop} to reset the global runloop id counter back to 0 when resuming a task.
-This way the task retains the same runloop id over its life time.
+When resuming a preempted task, this id is not the same as the one of the runloop created for executing its code.
+This leads to Parrot not finding the correct runloop when finalizing an exception.
+
+To mitigate this, the \textit{reset\_runloop\_id\_counter} function is used by \textit{Parrot\_cx\_outer\_runloop} to reset the global runloop id counter back to 0 when resuming a task.
+This way, the task retains the same runloop id over its life time.
 
@@
 It would also have to recreate these call and runloop stacks when resuming the task.
-Otherwise the task would be ended as soon as the executing callback would be completed since instead of some C function called by an op in the bytecode, the scheduler is the owner of the runloop.
-
-Thus the scheduler checks for the current runloop id and does not preempt a task if it is currently running a nested runloop.
+Otherwise, the task would be ended as soon as the executing callback would be completed since instead of some C function called by an operation in the bytecode, the scheduler is the owner of the runloop.
+
+Thus, the scheduler checks for the current runloop id and does not preempt a task if it is currently running a nested runloop.
 Furthermore even a manual yield to another task is not possible in such a situation.
-The only way to solve this problem would be to get rid of nested runloops in the interpreter.
+The only way to solve this problem is to get rid of nested runloops in the interpreter.
 
@@
 This thread sleeps until a new alarm is set or the currently active one expires.
-When the latter happens it increases the global alarm serial and notifies all threads that a timer is expired.
-Since it does not keep a list of pending alarms but knows only about the next one, interpreter threads are obliged to re-set their next alarms upon notification.
+When the latter happens, it increases the global alarm serial and notifies all threads that a timer is expired.
+Since it does not keep a list of pending alarms but knows only about the next one, interpreter threads are obliged to reset their next alarms upon notification.
 
 \section{Threads}
+
+This section describes how OS threads are used to execute tasks in parallel.
 
@@
 These interpreters are kept in the \textit{threads\_array} defined in \textit{src/thread.c}.
-It's a simple C array instead of one of Parrot's dynamic arrays implemented as PMCs because though ParrotInterpreter is implemented as a PMC itself, it is not fully garbage collected because of the bootstrapping issues this would create.
+It's a C array instead of one of Parrot's dynamic arrays implemented as PMCs because although ParrotInterpreter is implemented as a PMC itself, it is not garbage collected because of the bootstrapping issues this would create.
 
 \textit{src/thread.c} also contains thread management functions all prefixed by \textit{Parrot\_thread\_}.
-Threads are created when \textit{Parrot\_cx\_schedule\_task} determines that all existing threads are busy and new threads can yet be started.
-It uses \textit{Parrot\_thread\_create} to create a new interpreter by cloning the current one and giving it a cleared out \textit{thread\_data} structure.
+Threads are created when \textit{Parrot\_cx\_schedule\_task} determines that all existing threads are busy and new threads can be started.
+\textit{Parrot\_cx\_schedule\_task} uses \textit{Parrot\_thread\_create} to create a new interpreter by cloning the current one and giving it an empty \textit{thread\_data} structure.
 It also creates the \textit{thread\_data->notifierfd} pipe used to notify threads of expired alerts.
@@
 This function uses \textit{Parrot\_thread\_create\_local\_task} to create a corresponding task on the new interpreter and then pushes it onto the scheduler.
-Since each interpreter has its own garbage collector, all objects used on the interpreter must origin from this GC's memory pools.
-That's why \textit{Parrot\_thread\_create\_local\_task} creates a new \textit{local\_task}.
+Since each interpreter has its own GC, all objects used on the interpreter must origin from this GC's memory pools.
+That is why \textit{Parrot\_thread\_create\_local\_task} creates a new \textit{local\_task}.
 Proxies are created for the task's \textit{code} and \textit{data} attributes.
 The task object of the originating interpreter and the new \textit{local\_task} are linked using their \textit{partner} attributes containing a pointer to the other object.
-This is the point where for all variables in the original task's \textit{shared} array proxies are created an put into the \textit{shared} array of the \textit{local\_task}.
+This is the point, where for all variables in the original task's \textit{shared} array proxies are created an put into the \textit{shared} array of the \textit{local\_task}.
 \textit{Parrot\_thread\_create\_proxy} is used for this purpose.
@@
 
-\textit{Parrot\_thread\_run} is the function where the actual operating system thread is created.
-It uses macros defined in the \textit{include/parrot/thr\_*.h} include files which act as an OS abstraction layer.
+\textit{Parrot\_thread\_run} is the function where the actual OS thread is created. It uses macros defined in the \textit{include/parrot/thr\_*.h} include files which act as an OS abstraction layer.
 
 \textit{Parrot\_thread\_outer\_runloop} is used as the thread's main function.
-It's very similar to \textit{Parrot\_cx\_outer\_runloop}.
+It is very similar to \textit{Parrot\_cx\_outer\_runloop}.
 The most important difference is the way it waits for pending alarms in case no tasks are to be run.
@@
 They are the only means for a thread to access another thread's data and are implemented by the \textit{Proxy} PMC type.
-This type has only two attributes, none of which are garbage collected:
+This type has only two attributes, which are not garbage collected:
 %
 \begin{itemize}
-\item PMC *target: The PMC this object proxies to
-\item Parrot\_Interp interp: The interpreter owning target
+\item \textit{PMC *target}: the PMC this object proxies to.
+\item \textit{Parrot\_Interp interp}: the interpreter owning target.
 \end{itemize}
 
-As described in section~\ref{sec:PMCs} PMC source files to get preprocessed before compilation by a C compiler.
-This preprocessing gets done by the \textit{tools/build/pmc2c.pl} Perl script.
-This script got extended by implementing the \textit{Parrot::Pmc2c::PMC::Proxy} module which creates default implementations for all \textit{vtable} functions not otherwise defined in the \textit{proxy.pmc} file.
-The default implementation for all writing functions would just call \textit{cant\_do\_write\_method} which creates a runtime exception.
+As described in section~\ref{sec:PMCs}, PMC source files get preprocessed before compilation by a C compiler.
+This preprocessing is done by the \textit{tools/build/pmc2c.pl} Perl script.
+This script is extended by the \textit{Parrot::Pmc2c::PMC::Proxy} module which creates default implementations for all \textit{vtable} functions not otherwise defined in the \textit{proxy.pmc} file.
+The default implementation for all writing functions just calls \textit{cant\_do\_write\_method} which creates a runtime exception.
 
 All other methods call the vtable method of the same name on the \textit{target} passing the current interpreter as \textit{interp} and all other parameters unchanged.
-This causes all access to globals by the proxied function to go through the thread's interp and new PMCs created during the call to be allocated from the thread's memory pool.
-If the method returns a PMC from the target's interp, another proxy object has to be created and wrapped around it so it can be safely returned to the caller.
-
-To differ between PMCs originating from the target's \textit{interp} and those created on the thread's interp during the call the GC is told to set the new \textit{PObj\_is\_new\_FLAG} on newly created PMCs.
-The \textit{PARROT\_THR\_FLAG\_NEW\_PMC} flag on the interp itself is used to communicate this requirement to the GC.
-
-The alternative of calling the proxied function using the target's \textit{interp} would lead to concurrency issues with the garbage collector.
-Any call may cause garbage collection to run.
+This causes all access to globals by the proxied function to go through the thread's \textit{interp} and new PMCs created during the call to be allocated from the thread's memory pool.
+If the method returns a PMC from the target's \textit{interp}, another proxy object has to be created and wrapped around it so it can be safely returned to the caller.
+
+To differ between PMCs originating from the target's \textit{interp} and those created on the thread's \textit{interp} during the call, the GC is told to set the new \textit{PObj\_is\_new\_FLAG} on newly created PMCs.
+The \textit{PARROT\_THR\_FLAG\_NEW\_PMC} flag on the \textit{interp} is used to communicate this requirement to the GC.
+
+The alternative of calling the proxied function using the target's \textit{interp} leads to concurrency issues with the GC.
+Any call may cause GC to run.
 Since the GC also scans the C stack the target's GC would find the thread's PMCs there and mark them confusing the thread's GC.
-Disabling the garbage collector would require a lock since the target's GC may already have started collecting.
+Disabling the GC requires a lock since the target's GC may already have started collecting.
 
@@
 A \textit{Sub} does not only contain the code to execute but also the context in which to execute the code such as visible globals and namespaces.
-If a proxy to such a \textit{Sub} were created and \textit{invoke} called on it, the code would access this context directly since it belongs to the same interp as the proxied \textit{Sub} itself.
-Thus an op like \textit{get\_global} would fetch a global from an unproxied namespace and an unproxied global would be put into the target register.
+If a proxy to such a \textit{Sub} were created and \textit{invoke} called on it, the code would access this context directly since it belongs to the same \textit{interp} as the proxied \textit{Sub} itself.
+Thus, an operation like \textit{get\_global} fetches a global from an unproxied namespace and an unproxied global is be put into the target register.
 Since this is happening while running \textit{invoke} on the original \textit{Sub}, \textit{Proxy} cannot intercept the call and create a \textit{Proxy} for the result.
@@
 
-\subsection{Writing shared data}
-
-As described in chapter~\ref{cha:design} to write to shared data, a thread creates a small task and schedules it on the data owning interpreter. An example task looks like this:
+\subsection{Writing to shared variables}
+
+As described in chapter~\ref{cha:design}, to write to shared variables, a thread creates a task and schedules it on the data owning interpreter. An example task looks like this:
 %
@@
 This is a subroutine with just one parameter.
-The variable passed in as this parameter is the one the task should write to.
+The variable passed as this parameter is the one the task should write to.
 In this case the constant value 1 would be written to the variable.
-In PIR an assignment to a PMC gets translated as a method call.
-In this case the \textit{set\_integer\_native} would be called changing the variable's value.
-Since PMCs get passed by reference, it is the original variable which gets written to.
-
-The code to create the task would look like:
+In PIR, an assignment to a PMC gets translated to a method call.
+In this case, the \textit{set\_integer\_native} is called changing the variable's value.
+Since PMCs are passed by reference, it is the original variable which gets written to.
+
+Code to create the task looks like:
 %
@@
 %
-Line 1 creates a new Task object.
+Line 1 creates a new task object.
 The example subroutine is used for the task's \textit{code} attribute.
 \textit{shared\_variable} is used for \textit{data}.
-At this point \textit{shared\_variable} is actually the proxy object created for the shared integer PMC.
+At this point, \textit{shared\_variable} is actually the proxy object created for the shared integer PMC.
 The interpreter object contains a \textit{schedule\_proxied} method which is used to schedule the \textit{write\_task} on the thread owning the original variable.
 This owner thread's interpreter cannot be used directly for scheduling the task, since it would have to be stored in a register to be accessible to PIR code.
-But then the same problem as described in subsection~\ref{subsec:thread_creation} would occur with the ParrotInterpreter PMC tripping up the garbage collector.
+But then, the same problem as described in section~\ref{subsec:thread_creation} would occur with the \textit{ParrotInterpreter} PMC tripping up the GC.
 
@@
 
-\subsection{The \textit{wait} op}
-
-Using \textit{Tasks} to write to write to shared data makes such actions inherently asynchronous.
+\subsection{\textit{wait} operation}
+
+Using \textit{tasks} to write to shared variables makes such actions inherently asynchronous.
 This is not always what is needed by the implemented algorithm.
-For example when the shared variable is a lock, processing should continue as soon as it's acquired.
-The \textit{wait} op is used to wait for a task's completion.
+For example, when the shared variable is a lock, processing should continue as soon as it's acquired.
+The \textit{wait} operation is used to wait for a task's completion.
 The waiting task is added to the waited for task's \textit{waiters} list and preempted immediately.
-When a task finishes, the all tasks in the \textit{waiters} list are scheduled again for execution.
+When a task finishes, all the tasks in the \textit{waiters} list are scheduled again for execution.
 Since for each task a local copy is created on the target thread, the running task not only checks its own \textit{waiters} list but also its partner's.
 
-If a task on the main thread was waiting for a task on another thread to finish and no other tasks are in the scheduler's queue on the main thread, the main thread would exit if no alarms are pending.
-To prevent this unintended exit all tasks are added to the scheduler's \textit{foreign\_tasks} list when they are scheduled on other threads.
-To end the program with other threads still running an explicit \textit{exit} op has to be issued.
+If a task on the main thread was waiting for a task on another thread to finish and no other tasks are in the scheduler's queue on the main thread, the main thread exits if no alarms are pending.
+To prevent this unintended exit, all tasks are added to the scheduler's \textit{foreign\_tasks} list when they are scheduled on other threads.
+To end the program with other threads still running, an explicit \textit{exit} operation has to be used.
 
@@
 
-Since each interpreter has its own garbage collector and consequently its own memory areas and can only access its own data directly, the garbage collector can act as if there were no other threads.
-Thus the main work has been to clean up the interpreter structure itself and to replace references to global data by proxies.
-Only a few special cases had to be built into the GC itself.
-The global variable PMCNULL is used in many places in the code, but should be handled only by the main thread's garbage collector.
+Since each interpreter has its own GC and consequently its own memory areas and can only access its own data directly, the GC can act as if there were no other threads.
+After replacing references to global data by proxies, only a few special cases had to be built into the GC.
+The global variable \textit{PMCNULL} is used in many places in the code, but should be handled only by the main thread's GC.
 
 Before the implementation of threading support, having multiple interpreter objects in the system was used to implement things like a safe compartment where code can be run in a restricted environment.
-For such cases it was useful to share the garbage collector between interpreters.
+For such cases it was useful to share the GC between interpreters.
 This special handling had to be removed.
@@
 Child threads use these via proxies.
-In previous implementation attempts code segments, classes and vtables have been copied for each new interpreter making thread startup a very costly operation while still causing subtle problems.
-
-Code segments and vtables are only marked by the main thread's garbage collector.
+In previous implementation attempts code segments, classes and vtables have been copied for each new interpreter making thread startup a costly operation while still causing subtle problems.
+
+Code segments and vtables are only marked by the main thread's GC.
 Code segments are managed in Parrot using \textit{PackFile} structures.
-Besides the code itself these structures contain tables of constants used in the code.
+Besides the code itself, these structures contain tables of constants used in the code.
 These constants are expanded to PMCs by the bytecode loader when the code is loaded from disk or put there directly by the PIR compiler.
@@
 Accessing the contained constants would therefore lead to unproxied PMCs being accessed by different threads.
-Since PackFiles may only be loaded by the main thread anyway, the bytecode loader and the compiler were changed to flag all \textit{PackFile} constants as \textit{shared}.
+Since PackFiles may only be loaded by the main thread, the bytecode loader and the compiler were changed to flag all \textit{PackFile} constants as \textit{shared}.
 They still are used as is on all threads, but the GC knows to only handle \textit{shared} PMCs on the main thread.
@@
 Scheduling tasks on other interpreters may cause garbage collection due to the need of creating proxy objects.
-To protect the garbage collector from concurrency issues, garbage collection is disabled during creation of the new \textit{Task} object and the proxies on the target thread's \textit{interp}.
+To protect the GC from concurrency issues, garbage collection is disabled during creation of the new \textit{Task} object and the proxies on the target thread's \textit{interp}.
 But since the target thread may itself be allocating memory or collecting garbage, the GC has to be protected by a lock as well.
@@
 
+\subsection{Conclusion}
+
+This threading implementation is mostly contained in a few files.
+Only a few changes had to be done in other places like the \textit{ParrotInterpreter} and the GC.
+This is a major difference to the previous implementation where all PMCs contained threading related code.
+As the next chapter shows, this design helped to reach a working state and meet the performance goals quickly.
+
 \chapter{Tests and Benchmarks}
@@
 
-\section{tasks.pir}
-
-tasks.pir is a simple test program, executing two \textit{Tasks} called ``a'' and ``b'' which run in tight loops printing their name on each 100000th iteration.
+\section{\textit{tasks.pir}}
+
+tasks.pir is a simple test program, executing two tasks called \textit{a} and \textit{b} which run in tight loops printing their name on each 100000th iteration.
 The program ends after running for 10 seconds.
-This is a test of thread creation and task scheduling but does not access any shared data.
-The full code is as follows:
+This is a test of thread creation and task scheduling but does not access any shared variables.
+The code is as follows:
 %
@@
 
-\section{moretasks.pir}
-
-moretasks.pir is an extended test where the main task creates 50000 child tasks.
-These tasks would poll a shared variable with pauses of 100ms.
+The test shows that task creation and execution works and that tasks do run in parallel.
+
+\section{\textit{moretasks.pir}}
+
+\textit{moretasks.pir} is an extended test where the main task creates 50000 child tasks.
+These tasks poll a shared variable with pauses of 100 ms.
 When the variable is set to 1, they schedule tasks on the main thread to write their number to a results array.
-When the results array contains the expected number of results the process is repeated.
-
-This test stresses reading and writing to shared data, allocation and garbage collection.
+When the results array contains the expected number of results, the process is repeated.
+
+This test stresses reading and writing to shared variables, allocation and garbage collection.
 It proved to be valuable for finding all sorts of concurrency issues.
@@
 
-\section{matrix\_part.winxed}
-
-This test implements multiplication of a matrix by a vector using four threads.
-It is written in the \textit{Winxed} programming language.
-Winxed is a low-level language with JavaScript like syntax and the possibility to include sections of PIR code verbatim making it possible to try experimental opcodes while writing more readable and concise code than with PIR alone.
+The program runs stable and does not leak memory demonstrating that the GC works.
+
+\section{\textit{matrix\_part.winxed}}
+
+This test implements matrix multiplication \cite{MatrixMult} using four threads.
+For simplicity the second matrix only has one column.
+The program is written in the \textit{Winxed} programming language.
+Winxed is a low-level language with Javascript like syntax and the possibility to include sections of PIR code verbatim making it possible to try experimental opcodes while writing more readable and concise code than with PIR alone.
 The complete source code is available in appendix \ref{app:matrix_part}.
@@
 The program consists of the parts initialization, computation and verification.
-Computation is parallelized using 4 tasks each calculating one fourth of the result vector.
-Runtime is compared to a simple single threaded implementation.
-Run times were measured using the \textit{time} command and are recorded in table \ref{benchmark:matrix}
+Computation is parallelized using four tasks each calculating one fourth of the result vector.
+Runtime is compared to a simple singlethreaded implementation.
+Run times were measured using the \textit{time} command and are recorded in table \ref{benchmark:matrix}.
 
@@
 \hline
-1. run  & 28.522s & 19.530s & 17.543s & 8.478s \\
+1. run  & 28.522 s & 19.530 s & 17.543 s & 8.478 s \\
 \hline
-2. run  & 28.427s & 19.463s & 17.320s & 8.283s \\
+2. run  & 28.427 s & 19.463 s & 17.320 s & 8.283 s \\
 \hline
-3. run  & 28.200s & 19.235s & 17.489s & 8.473s \\
+3. run  & 28.200 s & 19.235 s & 17.489 s & 8.473 s \\
 \hline
-average & 28.383s & 19.409s & 17.451s & 8.411s \\
+average & 28.383 s & 19.409 s & 17.451 s & 8.411 s \\
 \hline
@@
 
-As can be seen the multi threaded implementation gives an average speedup of 2.31 for the computation and 1.61 in total.
-
-\section{chameneos.pir}
-
-Chameneos is a game useful for testing peer-to-peer cooperation of threads presented in \cite{Chameneos}.
+As can be seen, the multithreaded implementation gives an average speedup of 2.31 for the computation and 1.61 in total.
+
+\section{\textit{chameneos.pir}}
+
+\textit{Chameneos} is a game useful for testing peer-to-peer cooperation of threads presented in \cite{Chameneos}.
 It revolves around fictional creatures, each being of one of three colors.
-Two of these creatures may meet and exchange their colors with each one independently calculating and adopting a complementary color depending on those two.
+Two of these creatures may meet and exchange their colors, each one independently calculating and adopting a complementary color depending on those two.
 These creatures have to be implemented using preemptive kernel or lightweight threads.
-Since the color calculation is trivial this test's performance depends in full on the available thread synchronization primitives and writes to shared data.
+Since the color calculation is trivial this test's performance depends on the available thread synchronization primitives and writes to shared variables.
 It therefore represents a worst case scenario for the implemented threading model.
@@
 
-Preliminary benchmarks have shown Parrot's performance to be within an order of magnitude with that of an optimized implementation in Perl 5.
+Preliminary benchmarks have shown Parrot's performance to be within an order of magnitude of that of an optimized implementation in Perl 5.
 Since Parrot does not yet offer the user any synchronization primitives, locks had to be implemented using a shared variable which is written to only by the main thread.
-Replacing this very primitive method with a native semaphore implementation would probably reduce runtime to a small fraction.
-
-\section{mandel\_inter.winxed}
-
-Calculating an image of the Mandelbrot set is a common benchmark for multithreading implementations since calculations of points are independent of each other and are thus easily parallelizable.
+Replacing this primitive method with a native semaphore implementation would probably reduce runtime to a small fraction.
+
+\section{\textit{mandel\_inter.winxed}}
+
+Calculating an image of the Mandelbrot set \cite{Mandelbrot} is a common benchmark for multithreading implementations since calculations of points are independent of each other and are thus easily parallelizable.
 A simple implementation of the escape time algorithm written in Winxed has been used to determine scalability properties of the threading implementation.
 The image is split into lines which are calculated alternatedly by a configured number of tasks.
-Run times were measured using the \textit{time} command on an Intel Core i7 3770K processor with 16 GiB of RAM running openSUSE 12.1 and are recorded in table \ref{benchmark:mandel}
+Run times were measured using the \textit{time} command on an Intel Core i7 3770K processor with 16 GiB RAM running openSUSE 12.1 and are recorded in table \ref{benchmark:mandel}
 
@@
 
-As can be seen in the results, the implementation scales nearly linearly up to 4 threads reflecting the CPUs 4 physical cores.
-Using 8 threads the speedup is only 1.368 compared to 4 threads but this seems to be more a limitation of the hardware than the implementation.
+As can be seen, the implementation scales nearly linearly up to four threads reflecting the CPUs four physical cores.
+Using eight threads, the speedup is only 1.368 compared to four threads but this seems to be more a limitation of the hardware than the implementation.
 
@@
 
-This work shows a way to successfully implement threading support in a dynamic language runtime after the fact.
+This thesis shows a way to implement threading support in a dynamic language runtime.
 Changes to the interpreter itself could be kept at a reasonable level and very local to the threading implementation itself.
-Benchmarks have shown that no slowdowns were measurable in the still very important single threaded case, while the implementation scales well using multiple threads.
-
-Practice will show how this threading model will fare with real world workloads. It very much depends on the ratio of read to write accesses to shared data.
-
-\section{Further work}
+Benchmarks have shown that no slowdowns were measurable in the still very important singlethreaded case, while the implementation scales well using multiple threads.
+
+Practice will show how this threading model will fare with real world workloads. It very much depends on the ratio of read to write accesses to shared variables.
+
+\section{Further Work}
+
+There are several ways in which the current implementation can be improved.
 
@@
 
-The initial goal of an almost lock free implementation has not been met fully.
-The garbage collector has to take a lock in most of its functions.
+The initial goal of an almost lock free implementation has not been met.
+The GC has to take a lock in most of its functions.
 Though these locks are used only when more than one interpreter thread is running, they still represent an overhead which could be avoided.
-A way to accomplish this would be to pre-allocate proxy objects instead of allocating them on demand.
+A way to accomplish this is to pre-allocate proxy objects instead of allocating them on demand.
 The allocation could be done as part of a thread's outer runloop.
-This would eliminate the need for the locks in the garbage collector.
-The number of proxy objects to pre-allocate would be an important knob for optimization, since too few objects would lead to starvation quickly when many tasks are scheduled.
+This eliminates the need for the locks in the GC.
+The number of proxy objects to pre-allocate would be an important knob for optimization, since too few objects lead to starvation quickly when many tasks are scheduled.
 
@@
 
-In the current implementation there are several places with an implicit assumption that new threads may be only started from the main thread.
+In the current implementation, there are several places with an implicit assumption that new threads may be only started from the main thread.
 Similarly only the main thread may schedule tasks on other threads.
@@
 
-\subsection{Allow multi level proxying}
+\subsection{Allow multi-level proxying}
 
@@
 
-Currently the scheduler picks the interpreter with the smallest number of tasks in the task queue for scheduling a new task.
-This can lead to the situation where threads sit idle while tasks have to wait in other thread's queues.
+Currently, the scheduler picks the interpreter with the smallest number of tasks in the task queue for scheduling a new task.
+This can lead to a situation where threads sit idle while tasks have to wait in other thread's queues.
 Some tasks get scheduled on specific threads because they need write access to the thread's PMCs.
-To allow migration of waiting tasks to idle threads there needs to be a method to distinguish such tasks from tasks which are there simply because the thread's queue was the shortest at the time the task got scheduled.
+To allow migration of waiting tasks to idle threads, there has to be a method to distinguish such tasks from tasks which are there simply because the thread's queue was the shortest at the time the task got scheduled.
 The more difficult part of implementing task migration is that proxies are allocated on the target interpreter at the time of scheduling.
-Since these proxies belong to the interpreter they are created on, new proxies would have to be created on the interpreter the task is being migrated to.
+Since these proxies belong to the interpreter, they are created on, new proxies would have to be created on the interpreter the task is being migrated to.
 It is yet unclear if this overhead can be offset by better utilization of idle threads.
@@
 While I had to invest many more hours than strictly necessary to get a Bachelor's degree, the work was also very interesting, instructive and hopefully valuable to the Parrot project and the free software community.
-Considering that I had no prior experience with Parrot which is a highly complex piece of software, it is not surprising that I spent most of the time discovering how it works and how its many features interact.
+Considering that I had no prior experience with Parrot which is a complex piece of software, it is not surprising that I spent most of the time discovering how it works and how its many features interact.
 In addition, concurrency usually brings with it difficult to diagnose problems further complicated by Parrot's cross platform nature.
@@
 \label{app:moretasks}
+
+This program tests the GC by creating tasks and writing to shared variables.
 
@@
 
+This program implements multithreaded matrix multiplication.
+
 \lstinputlisting{/home/nine/install/parrot/examples/threads/matrix_part.winxed}
@@
 
+An implementation of the \textit{Chameneos} game testing thread synchronization.
+
 \lstinputlisting{/home/nine/install/parrot/examples/threads/chameneos.pir}
 
-\MakeBibliography{References, Literature}
+\MakeBibliography{References}
 

### 2012-06-25 Paper.tex
Line breaks in title
Examples for chapter 2

@@
 
-\title{Hybrid Threads for the Parrot Virtual Machine}
+\title{Hybrid Threads \\ for the \\ Parrot Virtual Machine}
 
@@
 
+Following is an example of a multithreaded program in Java. The program spawns 10,000 threads, each one waiting for the \textit{starter} variable to be set to 1 before adding it's name to an array:
+\input{thread-java}
+
+The program exploits the fact that all methods in the \textit{Vector} class are synchronized, i.e.
+thread-safe.
+If the \textit{results} array was a non-thread-safe \textit{ArrayList} instead, access would have to be locked manually:
+
+\begin{JavaCode}
+synchronize(results) {
+    results.add(name);
+}
+\end{JavaCode}
+
 \section{Python}
@@
 Since processes may not directly access other processes' memory, the \textit{multiprocessing} module provides several means of communication between processes: \textit{Queues}, \textit{Pipes} and shared memory support.
+
+Following is an example of a multithreaded program in Python.
+The program spawns 10,000 threads, each one waiting for the \textit{starter} variable to be set to 1 before appending it's name to an array:
+\input{thread-python}
+
+Even though many threads append to the same array, they do not have to take any locks. The GIL implicitly protects the array by allowing only one Python instruction to be run at any time.
 

### 2012-06-26 Paper.tex
First diagrams appear

@@
 \usepackage{listings}
+\usepackage{tikz}
+\usepackage{pgf-umlsd}
+\usepgflibrary{arrows}
+\usetikzlibrary{shadows}
 \AddBibFile{literature.bib}
@@
 
-\section{PolyMorphic Containers (PMCs)}
+\section{\textit{PolyMorphic Containers} (PMCs)}
 \label{sec:PMCs}
@@
 
-A runloop is the inner most loop executing bytecode in Parrot.
-It is also a data structure containing some information needed to support exception handling.
+A runloop is the inner most loop executing bytecode in Parrot. It consists of two steps as shown in the following pseudo code:
+\begin{GenericCode}
+while (pc != NULL) { // program counter points to an opcode
+    op = fetch_opcode_at(pc);
+    pc = execute_opcode(op);
+}
+\end{GenericCode}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.4]{runloops}
+\caption{Example of a callback executing in a nested runloop.}
+\label{fig:runloops}
+\end{figure}
+
+It is also a data structure containing information needed to support exception handling.
 When some operation in Parrot bytecode is calling a C function and this C function in turn is again executing Parrot bytecode, a nested runloop is started.
 Examples for such situations include calling a library function with a callback as parameter and exceptions thrown inside Parrot's C code calling a previously defined exception handler in user code.
+This is shown in figure~\ref{fig:runloops}.
 
@@
 Even a seemingly simple operation like an atomic increment can take two orders of magnitude longer than a normal increment \cite{LockingInOSKernels}.
-While the gain through being able to utilize multiple CPU cores may offset this cost, it is still impacting the common case of having only a singlethread running.
+While the gain through being able to utilize multiple CPU cores may offset this cost, it is still impacting the common case of having only a single thread running.
 
@@
 
-The solution for these problems implemented in this thesis is to side step them altogether by disallowing write access to shared variables.
+The solution for these problems implemented in this thesis is to sidestep them altogether by disallowing write access to shared variables.
 The programmer (or in most cases the compiler) is obliged to declare a list of all shared variables before a newly created task is started.
@@
 Put another way, the data-owning interpreter is told what to write to its variables, so other threads don't have to.
+
+\begin{figure}
+    \centering
+    \begin{sequencediagram}
+        \newthread{main}{:MainThread}
+        \newinst{shared}{:A}
+        \newthread{child}{:ChildThread}
+
+        \begin{callself}{main}{task1}{}
+        \end{callself}
+
+        \prelevel\prelevel
+        \begin{callself}{child}{task2}{}
+            \postlevel
+            \begin{messcall}[1]{child}{schedule(write\_task)}{main}
+                \begin{callself}{main}{write\_task}{}
+                    \begin{messcall}{main}{write}{shared}
+                    \end{messcall}
+                \end{callself}
+            \end{messcall}
+        \end{callself}
+
+        \prelevel\prelevel
+        \begin{callself}{main}{task1}{}
+        \end{callself}
+
+    \end{sequencediagram}
+    \caption{\textit{task2} running on \textit{ChildThread} is sending the \textit{write\_task} to the \textit{MainThread} to write to the shared variable \textit{A}.}
+    \label{fig:hybridthreads}
+\end{figure}
 

### 2012-06-27 Paper.tex
Fix a grammar error
Improved understandability of the implementation chapter

@@
 \usepgflibrary{arrows}
-\usetikzlibrary{shadows}
+\usetikzlibrary{shapes,arrows,shadows}
 \AddBibFile{literature.bib}
@@
 \begin{document}
+
+\tikzstyle{block} = [rectangle, draw, text width=12.3em, text centered, rounded corners, minimum height=2em]
+\tikzstyle{decision} = [diamond, draw, text width=4.5em, text badly centered, node distance=2cm, inner sep=0pt]
+\tikzstyle{line} = [draw, -latex']
 
@@
 Since cross-threaded writes are already forbidden and all read access to other thread's data goes through the narrow channel of proxy objects, forcing a complete separation of the thread's memory domains is only a small step.
-By having separate memory areas for each interpreter, it becomes possible to have each interpreter to run its own GC.
+By having separate memory areas for each interpreter, it becomes possible to have each interpreter run its own GC.
 This way, the known to work singlethreaded GC implementation can be used nearly unchanged.
@@
 The scheduler is the place where most of the green thread logic is implemented.
-It consists of two parts: functions using the prefix \textit{Parrot\_cx} which are located in \textit{src/scheduler.c} and the scheduler PMC type containing the scheduler's data and public interface.
-
-\textit{Parrot\_cx\_init\_scheduler} is called from \textit{Parrot\_interp\_initialize\_interpreter} whenever a new interpreter is created, e.g., on interpreter startup and whenever a new thread is created.
+It consists of two parts: functions using the prefix \textit{Parrot\_cx} which are located in \textit{src/scheduler.c} and the \textit{Scheduler} PMC type containing the scheduler's data and public interface.
+
+The process of executing and preempting tasks is pictured in figure~\ref{controlflow} and described in detail in the rest of this section.
+
+\begin{figure}
+\centering
+\begin{tikzpicture}[node distance = 1cm, auto]
+\node[block](outerrl){Parrot\_cx\_outer\_runloop};
+\node[block, below of=outerrl](nexttask){Parrot\_cx\_next\_task};
+\node[block, below of=nexttask](runloop){runloop};
+\node[decision, below of=runloop] (branch) {branch?};
+\node[block, below of=branch, node distance=2cm](checksched){Parrot\_cx\_check\_scheduler};
+\node[block, below of=checksched](runsched){Parrot\_cx\_run\_scheduler};
+\node[block, right of=runsched, node distance=13.8em](checkquant){Parot\_cx\_check\_quantum};
+\node[decision, below of=runsched] (taskdone) {Quantum done?};
+\node[block, below of=taskdone, node distance=2cm](preempt){Parrot\_cx\_preempt\_task};
+\node[block, right of=preempt, node distance=13.8em](stop){Parrot\_cx\_stop\_task};
+\path[line](outerrl) -- (nexttask);
+\path[line](nexttask) -- (runloop);
+\path[line](runloop) -- (branch);
+\path[line](branch.south) to node[right] {yes} (checksched.north);
+\path[line](branch.west) to [out=180,in=220] node[right] {no} (runloop.west);
+\path[line](checksched) -- (runsched);
+\path[line](runsched) -- (checkquant);
+\path[line](runsched) -- (taskdone);
+\path[line](taskdone.south) to node[right] {yes} (preempt.north);
+\path[line](taskdone.west) to [out=180,in=220] node[right] {no} (runloop.west);
+\path[line](preempt) -- (stop);
+\draw[->] (preempt.west) to [out=120,in=240] (outerrl.west);
+\end{tikzpicture}
+\caption{Control flow of task execution and preemption.}
+\label{fig:controlflow}
+\end{figure}
+
+\textit{Parrot\_cx\_init\_scheduler} is called from \textit{Parrot\_\-interp\-\_ini\-tialize\_\-inter\-preter} whenever a new interpreter is created, e.g., on interpreter startup and whenever a new thread is created.
 It creates the scheduler PMC and sets up alarm signal handling.
@@
 Despite its name, it may not be confused with runloops discussed in section~\ref{sec:Runloops}.
-\textit{Parrot\_cx\_outer\_runloop} actually consists of two nested loops.
+This function has to execute tasks as long as tasks are available.
+If no tasks are available, it lets the thread sleep to not uselessly burn CPU time.
+On the other hand it should simply end instead of sleep if no tasks are available, no alarms pending and no tasks are being executed on other threads.
+
+As can be seen in listing~\ref{fig:outerrunloop}, \textit{Parrot\_cx\_outer\_runloop} therefore consists of two nested loops.
 The inner loop fetches tasks from the task queue and executes as long as tasks are available.
@@
 
-\textit{Parrot\_cx\_next\_task} contains the code to take the next task from the run queue and executing it.
+\begin{figure}
+\centering
+\begin{GenericCode}
+do {
+    while (tasks_available()) {
+        Parrot_cx_next_task()
+        Parrot_cx_check_alarms()
+    }
+
+    clean_finished_foreign_tasks()
+
+    if (
+        not tasks_available()
+        and (alarms_pending() or foreign_tasks_active())
+    ) {
+        Parrot_thread_wait_for_notification()
+        Parrot_cx_check_alarms()
+    }
+} while (
+    alarms_pending()
+    or foreign_tasks_active()
+    or tasks_available()
+)
+\end{GenericCode}
+\caption{Pseudo code describing Parrot\_cx\_outer\_runloop.}
+\label{fig:outerrunloop}
+\end{figure}
+
+\textit{Parrot\_cx\_next\_task} contains the code to take the next task from the run queue and execute it.
 Before calling the task, it checks if there are other tasks still in the queue.
@@
 
-When an alarm is triggered, Parrot only increments the global \textit{alarm\_serial}.
-This serial is checked for changes by \textit{Parrot\_alarm\_check} being called by \textit{Parrot\_cx\_check\_scheduler}.
-This function in turn is called once for every \textit{branch} operation.
-When the serial is different from the last check or the \textit{SCHEDULER\_wake\_requested} flag is set, \textit{Parrot\_cx\_check\_scheduler} wakes up the scheduler by calling \textit{Parrot\_cx\_run\_scheduler}.
-
-Since POSIX alarms can only be delivered to the main thread of a process, \textit{Parrot\_cx\_run\_scheduler} uses \textit{Parrot\_thread\_notify\_threads} to relay alarms to other thread's schedulers.
-For this purpose, every thread has a pipe which \textit{Parrot\_thread\_notify\_threads} writes a dummy byte to.
-The next step for the \textit{Parrot\_cx\_run\_scheduler} is to check for any expired alarms using \textit{Parrot\_cx\_check\_alarms}.
-An alarm has a task as callback.
-On expiry, this callback gets scheduled for immediate execution.
-As alarm signals do not contain any information about their origin, the scheduler does not know if the alarm was scheduled by user code or if it is the scheduler's preemption alarm.
-\textit{Parrot\_cx\_check\_quantum} is therefore used to check if the current task has used its assigned quantum and if it should be preempted.
+Checking for expired alarms after every executed operation would be too costly.
+Therefore the \textit{branch} operation has been picked as the point where alarms are checked.
+To make this as cheap as possible, Parrot increments the global \textit{alarm\_serial}.
+This serial is compared to the value last seen by the \textit{Parrot\_cx\_check\_scheduler} function.
+If they are different, or the \textit{SCHEDULER\_wake\_requested} flag is set, \textit{Parrot\_cx\_check\_scheduler} wakes up the scheduler by calling \textit{Parrot\_cx\_run\_scheduler}.
+
+\textit{Parrot\_cx\_check\_quantum} is used to check if the alarm signal means that the current task has used its assigned quantum and should therefore be preempted.
 
@@
 
-The control flow ends up back at \textit{Parrot\_cx\_next\_task} at the point after executing the task. \textit{Parrot\_cx\_next\_task} then returns to \textit{Parrot\_cx\_outer\_runloop}.
+The control flow ends up back at \textit{Parrot\_cx\_next\_task} at the point after executing the task. \textit{Parrot\_cx\_next\_task} then returns to \textit{Parrot\_cx\_\-outer\_\-runloop}.
 
@@
 These interpreters are kept in the \textit{threads\_array} defined in \textit{src/thread.c}.
-It's a C array instead of one of Parrot's dynamic arrays implemented as PMCs because although ParrotInterpreter is implemented as a PMC itself, it is not garbage collected because of the bootstrapping issues this would create.
+It is a C array instead of one of Parrot's dynamic arrays implemented as PMCs because although ParrotInterpreter is implemented as a PMC itself, it is not garbage collected because of the bootstrapping issues this would create.
 
@@
 \textit{Parrot\_cx\_schedule\_task} uses \textit{Parrot\_thread\_create} to create a new interpreter by cloning the current one and giving it an empty \textit{thread\_data} structure.
-It also creates the \textit{thread\_data->notifierfd} pipe used to notify threads of expired alerts.
 
 The task is then scheduled with the new interpreter's scheduler using \textit{Parrot\_thread\_schedule\_task}.
-This function uses \textit{Parrot\_thread\_create\_local\_task} to create a corresponding task on the new interpreter and then pushes it onto the scheduler.
+This function uses \textit{Parrot\_\-thread\_\-create\_\-local\_\-task} to create a corresponding task on the new interpreter and then pushes it onto the scheduler.
 Since each interpreter has its own GC, all objects used on the interpreter must origin from this GC's memory pools.
-That is why \textit{Parrot\_thread\_create\_local\_task} creates a new \textit{local\_task}.
-Proxies are created for the task's \textit{code} and \textit{data} attributes.
-The task object of the originating interpreter and the new \textit{local\_task} are linked using their \textit{partner} attributes containing a pointer to the other object.
-This is the point, where for all variables in the original task's \textit{shared} array proxies are created an put into the \textit{shared} array of the \textit{local\_task}.
+That is why \textit{Parrot\_thread\_create\_local\_task} creates a new \textit{local\_task} as desribed in figure~\ref{fig:createlocaltask}.
+Proxies are created for the task's \textit{code} and \textit{data} attributes (lines 3 and 4).
+The task object of the originating interpreter and the new \textit{local\_task} are linked using their \textit{partner} attributes containing a pointer to the other object (lines 11 and 12).
+For all variables in the original task's \textit{shared} array proxies are created an put into the \textit{shared} array of the \textit{local\_task} (lines 6--9).
 \textit{Parrot\_thread\_create\_proxy} is used for this purpose.
+
+\begin{figure}
+\centering
+\begin{GenericCode}
+local_task = new Task()
+
+local_task->code = Parrot_thread_create_proxy(task->code)
+local_task->data = Parrot_thread_create_proxy(task->data)
+
+foreach (variable in task->shared) {
+    proxy = Parrot_thread_create_proxy(variable)
+    push(proxy, local_task->shared)
+}
+
+local_task->partner = task
+task->partner = local_task
+\end{GenericCode}
+\caption{Pseudo code describing Parrot\_thread\_create\_local\_task.}
+\label{fig:createlocaltask}
+\end{figure}
 
@@
 It is very similar to \textit{Parrot\_cx\_outer\_runloop}.
-The most important difference is the way it waits for pending alarms in case no tasks are to be run.
-While \textit{Parrot\_cx\_outer\_runloop} waits for an alarm signal to arrive, \textit{Parrot\_thread\_outer\_runloop} waits for a dummy to be written to its \textit{notifierfd} pipe by \textit{Parrot\_cx\_run\_scheduler}.
+The most important difference is that \textit{Parrot\_thread\_outer\_runloop} waits indefinitely for new tasks while \textit{Parrot\_cx\_outer\_runloop} ends if no tasks are queued and no alarms are pending.
 
@@
 As described in section~\ref{sec:PMCs}, PMC source files get preprocessed before compilation by a C compiler.
-This preprocessing is done by the \textit{tools/build/pmc2c.pl} Perl script.
+This preprocessing is done by the \textit{pmc2c.pl} Perl script.
 This script is extended by the \textit{Parrot::Pmc2c::PMC::Proxy} module which creates default implementations for all \textit{vtable} functions not otherwise defined in the \textit{proxy.pmc} file.
@@
 To differ between PMCs originating from the target's \textit{interp} and those created on the thread's \textit{interp} during the call, the GC is told to set the new \textit{PObj\_is\_new\_FLAG} on newly created PMCs.
-The \textit{PARROT\_THR\_FLAG\_NEW\_PMC} flag on the \textit{interp} is used to communicate this requirement to the GC.
+The \textit{PARROT\_\-THR\_\-FLAG\_\-NEW\_\-PMC} flag on the \textit{interp} is used to communicate this requirement to the GC.
 
@@
 \textit{shared\_variable} is used for \textit{data}.
-At this point, \textit{shared\_variable} is actually the proxy object created for the shared integer PMC.
+At this point, \textit{shared\_\-variable} is actually the proxy object created for the shared integer PMC.
 The interpreter object contains a \textit{schedule\_proxied} method which is used to schedule the \textit{write\_task} on the thread owning the original variable.
@@
 \textit{schedule\_proxied} uses \textit{Parrot\_thread\_create\_local\_task} which in this case detects that the \textit{data} given as parameter for the task's \textit{code} is actually a proxy already and unwraps the proxied object.
-\textit{Parrot\_cx\_schedule\_immediate} is then used to make the data owning interpreter execute the task as soon as possible.
+\textit{Parrot\_cx\_schedule\_\-immediate} is then used to make the data owning interpreter execute the task as soon as possible.
 
@@
 
-As can be seen, the multithreaded implementation gives an average speedup of 2.31 for the computation and 1.61 in total.
+As can be seen, the multithreaded implementation gives an average speed\-up of 2.31 for the computation and 1.61 in total.
 

### 2012-06-28 Paper.tex
Absolute final fixes

@@
 
+\tikzstyle{endpoint} = [circle, draw, text width=1.8em, text centered, rounded corners, minimum height=1.8em]
 \tikzstyle{block} = [rectangle, draw, text width=12.3em, text centered, rounded corners, minimum height=2em]
-\tikzstyle{decision} = [diamond, draw, text width=4.5em, text badly centered, node distance=2cm, inner sep=0pt]
+\tikzstyle{decision} = [diamond, draw, text width=4.5em, text badly centered, node distance=1.9cm, inner sep=0pt]
 \tikzstyle{line} = [draw, -latex']
@@
 
-\chapter{Introduction and Motivation}
+\chapter{Introduction and motivation}
 
@@
 
-\chapter{Concurrency in other Programming Platforms}
+\chapter{Concurrency in other programming platforms}
 
@@
 \section{Java}
+\label{sec:java}
 
@@
 
-Even though many threads append to the same array, they do not have to take any locks. The GIL implicitly protects the array by allowing only one Python instruction to be run at any time.
+Even though many threads append to the same array, they do not have to take any locks.
+The GIL implicitly protects the array by allowing only one Python instruction to be run at any time.
+This is a contrast to the Java version in section~\ref{sec:java} where the user has to prevent concurrency issues either by using a thread-safe data structure or through manual locking.
 
@@
 \item string and
-\item polymorphic container (PMC).
+\item PolyMorphic Container (PMC).
 \end{itemize}
@@
 
-\section{Garbage Collector (GC)}
+\section{Garbage collector (GC)}
 
@@
 
-\chapter{Design of Hybrid Threads}
+\chapter{Design of hybrid threads}
 \label{cha:design}
@@
 
-\section{Critical Sections}
+\section{Critical sections}
 
@@
 
-\section{Garbage Collection (GC)}
+\section{Garbage collection (GC)}
 
@@
 
-\chapter{Implementation of Hybrid Threads}
+\chapter{Implementation of hybrid threads}
 
@@
 
-The process of executing and preempting tasks is pictured in figure~\ref{controlflow} and described in detail in the rest of this section.
+The process of executing and preempting tasks is pictured in figure~\ref{fig:controlflow} and described in detail in the rest of this section.
 
@@
 \begin{tikzpicture}[node distance = 1cm, auto]
-\node[block](outerrl){Parrot\_cx\_outer\_runloop};
-\node[block, below of=outerrl](nexttask){Parrot\_cx\_next\_task};
+\node[endpoint](start){start};
+\node[block, below of=start, node distance=1.1cm](outerrl){Parrot\_cx\_outer\_runloop};
+\node[decision, below of=outerrl] (moretasks) {tasks available?};
+\node[block, below of=moretasks, node distance=2cm](nexttask){Parrot\_cx\_next\_task};
+\node[endpoint, right of=moretasks, node distance = 13.8em](end){end};
 \node[block, below of=nexttask](runloop){runloop};
-\node[decision, below of=runloop] (branch) {branch?};
+\node[decision, below of=runloop, node distance=1.7cm] (branch) {branch?};
 \node[block, below of=branch, node distance=2cm](checksched){Parrot\_cx\_check\_scheduler};
@@
 \node[block, right of=preempt, node distance=13.8em](stop){Parrot\_cx\_stop\_task};
-\path[line](outerrl) -- (nexttask);
+\path[line](start) -- (outerrl);
+\path[line](outerrl) -- (moretasks);
+\path[line](moretasks.south) to node[right] {yes} (nexttask.north);
+\path[line](moretasks.east) to  node[below] {no} (end.west);
 \path[line](nexttask) -- (runloop);
@@
 \path[line](branch.south) to node[right] {yes} (checksched.north);
-\path[line](branch.west) to [out=180,in=220] node[right] {no} (runloop.west);
+\path[line](branch.west) to [out=180,in=230] node[right] {no} (runloop.west);
 \path[line](checksched) -- (runsched);
 \path[line](runsched) -- (checkquant);
+\path[line](checkquant)[yshift=-0.3cm] -- (runsched)[yshift=-0.3cm];
 \path[line](runsched) -- (taskdone);
 \path[line](taskdone.south) to node[right] {yes} (preempt.north);
-\path[line](taskdone.west) to [out=180,in=220] node[right] {no} (runloop.west);
+\path[line](taskdone.west) to [out=150,in=230] node[right] {no} (runloop.west);
 \path[line](preempt) -- (stop);
-\draw[->] (preempt.west) to [out=120,in=240] (outerrl.west);
+\path[line](stop)[yshift=-0.3cm] -- (preempt)[yshift=-0.3cm];
+\draw[->] (preempt.west) to [out=110,in=250] (outerrl.west);
 \end{tikzpicture}
@@
 \textit{get\_integer} returns the number of tasks in the \textit{task\_queue}.
-These four methods use the \textit{task\_queue\_lock} to make accessing the \textit{task\_queue} thread safe.
+These four methods use the \textit{task\_queue\_lock} to make accessing the \textit{task\_queue} thread-safe.
 Thus, the Scheduler PMC acts as a container while using \textit{task\_queue} to actually store the data.
@@
 \item \textit{PMC *shared}: list of variables shared with this task.
-\item \textit{PMC *partner}: copy of this task on the other side of a GC barrier, meaning in another thread.
+\item \textit{PMC *partner}: copy of this task in another thread.
 \end{itemize}
@@
 
-\section{Alarms and Timers}
+\section{Alarms and timers}
 
@@
 
-\subsection{Garbage Collection}
+\subsection{Garbage collection}
 
@@
 The global variable \textit{PMCNULL} is used in many places in the code, but should be handled only by the main thread's GC.
-
-Before the implementation of threading support, having multiple interpreter objects in the system was used to implement things like a safe compartment where code can be run in a restricted environment.
-For such cases it was useful to share the GC between interpreters.
-This special handling had to be removed.
 
@@
 
-\chapter{Tests and Benchmarks}
+\chapter{Tests and benchmarks}
 
@@
 \hline
-1. run  & 89.931s & 89.978s & 45.813s & 24.028s & 17.445s \\
+1. run  & 89.931 s & 89.978 s & 45.813 s & 24.028 s & 17.445 s \\
 \hline
-2. run  & 89.707s & 89.871s & 45.906s & 24.048s & 17.695s \\
+2. run  & 89.707 s & 89.871 s & 45.906 s & 24.048 s & 17.695 s \\
 \hline
-3. run  & 90.318s & 89.839s & 45.951s & 24.049s & 17.573s \\
+3. run  & 90.318 s & 89.839 s & 45.951 s & 24.049 s & 17.573 s \\
 \hline
-average & 89.985s & 89.896s & 45.890s & 24.042s & 17.571s  \\
+average & 89.985 s & 89.896 s & 45.890 s & 24.042 s & 17.571 s  \\
 \hline
-speedup & 1       &  1      &  1.959  &  3.739  &  5.116  \\
+speedup &  1.000  &  1.001  &  1.959  &  3.739  &  5.116  \\
 \hline
@@
 
-\chapter{Conclusion, Further Work and Experiences}
+\chapter{Conclusion, further work and experiences}
 
@@
 
-\section{Further Work}
+\section{Further work}
 
@@
 \appendix
-\chapter{moretasks.pir}
+\chapter{\textit{moretasks.pir}}
 \label{app:moretasks}
@@
 
-\chapter{matrix\_part.winxed}
+\chapter{\textit{matrix\_part.winxed}}
 \label{app:matrix_part}
@@
 
-\chapter{chameneos.pir}
+\chapter{\textit{chameneos.pir}}
 \label{app:chameneos_part}

### 2012-06-28 Praktikum.tex
Absolute final fixes

@@
 Wir entwickeln Software für den internen Gebrauch sowie für externe Kunden, z.B.
-das Content Management System, die Projektverwaltung, Online-Steuerberechnung und Individualentwicklung wie z.B.
+das Content Management System, die Projektverwaltung, Online-Steuerberech\-nung und Individualentwicklung wie z.B.
 ein CRM System für einen unserer Kunden.
@@
 \item{Erweiterung des Inline::Python Perlmoduls um die von uns benötigten Features}
-\item{Implementierung von automatischen Kompatibilitätstests mittels Screenshotvergleich der Webseiten mit allen Unterseiten.}
+\item{Implementierung von automatischen Kompatibilitätstests mittels Vergleich von Screenshots der Webseiten mit allen Unterseiten.}
 \end{itemize}
@@
 Die Objekte einer Website werden in einem Baum verwaltet.
-Um Abfragen auf Teilbäume effizienter zu machen, werden in der Tabelle \textit{zms\_object\_tree} Informationen über die Baumstruktur gespeichert.
+Um Abfragen auf Teilbäume effizienter zu machen, werden in der Tabelle \textit{zms\_object\_\-tree} Informationen über die Baumstruktur gespeichert.
 Sie erlaubt effiziente Abfragen auf komplette Unterbäume mit optionaler Begrenzung der Suchtiefe.
@@
 
-Die von den Inhaltsobjekten unterstützten Attribute sind als von \textit{ZMS::Attribute} abgeleitete Klassen implementiert.
+Die von den Inhaltsobjekten unterstützten Attribute sind als von \textit{ZMS::\-Attribute} abgeleitete Klassen implementiert.
 Sie enthalten Informationen über den möglicherweise benutzten Datentyp eines Datenbankfeldes, Code zum Rendern des Eingabeinterfaces und für die Eingabevalidierung.
@@
 Um die Ausgabe der Website kümmert sich \textit{ZMS::Controller::Content}.
-Die Verwaltung der Inhalte wird von \textit{ZMS::Controller::Content::Management} gestellt.
+Die Verwaltung der Inhalte wird von \textit{ZMS::Con\-troller::Content::Management} gestellt.
 Die Verwaltung der Metadaten wie Typen, Sprachen und Benutzer finden sich im \textit{ZMS::Controller::System} Namensraum.
